---
title: "Regression Deep Dive"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

## Quick Recap: What We Covered Last Time {.smaller}

Last lecture, we introduced regression:

::: incremental
-   **Simple regression** models the relationship between one predictor and an outcome
-   **Multiple regression** adds controls - allowing us to "hold constant" confounders
-   **R-squared** tells us how much variation our model explains
-   **Residual plots** help us diagnose problems
:::

. . .

Today, we'll dig in. We'll formalize the intuition, learn what can go wrong, and extend regression to handle more complex questions.

## Today

::: incremental
1.  **Deeper on coefficients**: Binary vs. continuous predictors, non-linear terms
2.  **The R-squared trap**: Why blindly adding variables is dangerous
3.  **When regression fails**: Omitted variable bias, collider bias, Simpson's Paradox
4.  **Interactions**: What if the effect of X varies
5.  **Marginal effects**: How to best interpret interaction models
:::

# Part 1: Understanding Coefficients

## Binary vs. Continuous Predictors {.smaller}

Last time, our predictors were continuous (spending in dollars, poverty rate). But predictors can also be **binary** - taking values of 0 or 1.

. . .

Imagine we want to predict children's test scores based on whether their mother graduated from high school:

$$
\text{Kid Score} = 78 + 12 \times \text{Mom HS Grad} + \epsilon
$$

. . .

::: incremental
-   **If Mom didn't graduate** (Mom HS Grad = 0): Predicted score = 78 + 12(0) = **78**
-   **If Mom graduated** (Mom HS Grad = 1): Predicted score = 78 + 12(1) = **90**
-   $\beta_1 = 12$ is just the **difference in group means** — just like a t-test
:::

## Bivariate Regression with a Binary Predictor

With a binary predictor, regression gives us the difference in **conditional expectations**:

$$\beta_1 = E[Y | X = 1] - E[Y | X = 0]$$

. . .

This is equivalent to a difference in means test — regression and t-tests are doing the same thing here.

. . .

::: callout-tip
## Key Insight

Simple regression with a binary predictor is just a fancy t-test. The machinery is the same; regression just gives us a more flexible framework.
:::

## Continuous Predictors: The Slope {.smaller}

Now imagine we have *years of education* instead of a binary variable:

$$
\text{Kid Score} = 40 + 3.5 \times \text{Mom Education Years} + \epsilon
$$

. . .

::: incremental
-   $\beta_0 = 40$: Predicted score when Mom has 0 years of education
-   $\beta_1 = 3.5$: For **each additional year** of Mom's education, the kid's predicted score increases by 3.5 points
-   This is the **slope** — the rate of change in Y as X increases by one unit
:::

## The Slope Is a Partial Derivative {.smaller}

You don't need calculus for this class, but it's worth seeing the connection once:

$$
\frac{d \widehat{Y}}{d X} = \beta_1
$$

. . .

This just says: "if I nudge X up by a tiny amount, how much does predicted Y change?" The answer is $\beta_1$ — the slope.

. . .

With multiple predictors, the same idea applies to each one:

$$
\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 \quad \Longrightarrow \quad \frac{\partial \widehat{Y}}{\partial X_1} = \beta_1
$$

. . .

The partial derivative just means "the rate of change in Y with respect to $X_1$, **holding** $X_2$ fixed." That's exactly what "holding constant" means — the calculus and the intuition are saying the same thing.

## The Linearity Assumption

With a continuous predictor, our model assumes the effect is **constant**:

$$
E[Y | X = 12] - E[Y | X = 11] = E[Y | X = 16] - E[Y | X = 15] = \beta_1
$$

. . .

In words: going from 11 to 12 years of education has the **same** predicted effect as going from 15 to 16 years.

. . .

Is this realistic? Probably not! The jump from no high school diploma to a diploma might matter more than the jump from a master's to a PhD.

. . .

What can we do about it?

## Adding Non-Linear Terms

We can include **transformations** of X in our model. The most common: a quadratic term.

$$
Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \epsilon
$$

. . .

Now the effect of X is no longer constant — it changes as X changes:

$$
\frac{\partial Y}{\partial X} = \beta_1 + 2\beta_2 X
$$

## Plot: Linear vs Quadratic

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4

library(tidyverse)

set.seed(30317)
n <- 50
MomEdu <- runif(n, min = 0, max = 24)
KidScore <- 40 + 3.5 * MomEdu - 0.1 * (MomEdu^2) + rnorm(n, 0, 2)

mod_linear <- lm(KidScore ~ MomEdu)
mod_quad <- lm(KidScore ~ MomEdu + I(MomEdu^2))

pred_data <- data.frame(MomEdu = seq(0, 24, length.out = 200))
pred_data$Linear <- predict(mod_linear, newdata = pred_data)
pred_data$Quadratic <- predict(mod_quad, newdata = pred_data)

ggplot(mapping = aes(x = MomEdu, y = KidScore)) +
  geom_point(data = data.frame(MomEdu, KidScore), color = "#522D80", size = 2) +
  geom_line(data = pred_data, aes(y = Linear, color = "Linear"), linewidth = 1) +
  geom_line(data = pred_data, aes(y = Quadratic, color = "Quadratic"), linewidth = 1) +
  scale_color_manual(values = c("Linear" = "#522D80", "Quadratic" = "#F66733")) +
  labs(title = "Linear vs. Quadratic Fit",
       x = "Mom Education (years)", y = "Kid Score", color = "Model") +
  theme_minimal(base_size = 14)
```

The quadratic model captures **diminishing returns** — once mom has a college degree, further education doesn't help much.

## When to Include a Non-Linear Term?

::: incremental
-   If you have a **theory** that suggests diminishing (or increasing) returns
    -   Or, in our example, returns that go negative after a certain point
-   If your **residual plot** shows a curve when you fit a linear model
-   **Be careful**: adding extra predictors always "improves" the fit (more on this next)
-   Another common transformation: $\log(X)$ if you expect declining marginal returns
:::

## Making a Residual Plot in R {.smaller}

A residual plot is just three steps: fit the model, get the residuals, and plot them against the fitted values.

```{r}
#| echo: true
#| fig-width: 8
#| fig-height: 4

# Step 1: Fit the model (we already did this)
linear_fit <- lm(KidScore ~ MomEdu)

# Step 2: Get predicted values and residuals
resid_data <- data.frame(
  fitted = fitted(linear_fit),
  residual = resid(linear_fit)
)

# Step 3: Plot residuals vs. fitted values
ggplot(resid_data, aes(x = fitted, y = residual)) +
  geom_point(color = "#522D80", size = 2, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#F66733") +
  labs(x = "Predicted Value", y = "Residual") +
  theme_minimal(base_size = 14)
```

. . .

**That's it.** If the points scatter randomly around the dashed line, your model is doing fine. If you see a pattern: a curve, a funnel, a trend -something is off.

## Spotting Non-Linearity in Residuals {.smaller}

Here's the same education data. We fit a straight line, but the true relationship is curved. The residual plot makes the problem obvious:

```{r}
#| fig-width: 10
#| fig-height: 4

library(patchwork)

# Reuse the data from the earlier slide
linear_fit <- lm(KidScore ~ MomEdu)

resid_data <- data.frame(
  MomEdu = MomEdu,
  KidScore = KidScore,
  fitted = fitted(linear_fit),
  residual = resid(linear_fit)
)

p1 <- ggplot(resid_data, aes(x = MomEdu, y = KidScore)) +
  geom_point(color = "#522D80", size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "#F66733", linewidth = 1.2) +
  labs(title = "Linear Fit Forced Through Curved Data",
       x = "Mom Education (years)", y = "Kid Score") +
  theme_minimal(base_size = 12)

p2 <- ggplot(resid_data, aes(x = fitted, y = residual)) +
  geom_point(color = "#522D80", size = 2, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#F66733", linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "gray40", linewidth = 0.8) +
  labs(title = "Residual Plot Shows the U-Shape",
       x = "Fitted Values", y = "Residual") +
  theme_minimal(base_size = 12)

p1 + p2
```

. . .

The U-shape in the residuals tells us the linear model is **systematically wrong** — it overpredicts in the middle and underpredicts at the extremes. This is your cue to try a quadratic (or log) term.

## What a Good Residual Plot Looks Like {.smaller}

Now let's fit the quadratic model to the same data and check the residuals again:

```{r}
#| fig-width: 10
#| fig-height: 4

quad_fit <- lm(KidScore ~ MomEdu + I(MomEdu^2))

resid_quad <- data.frame(
  MomEdu = MomEdu,
  KidScore = KidScore,
  fitted = fitted(quad_fit),
  residual = resid(quad_fit)
)

p1 <- ggplot(resid_quad, aes(x = MomEdu, y = KidScore)) +
  geom_point(color = "#522D80", size = 2, alpha = 0.7) +
  geom_line(aes(y = fitted), color = "#F66733", linewidth = 1.2) +
  labs(title = "Quadratic Fit (Matches the Curve)",
       x = "Mom Education (years)", y = "Kid Score") +
  theme_minimal(base_size = 12)

p2 <- ggplot(resid_quad, aes(x = fitted, y = residual)) +
  geom_point(color = "#522D80", size = 2, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#F66733", linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "gray40", linewidth = 0.8) +
  labs(title = "Residuals: No Pattern",
       x = "Fitted Values", y = "Residual") +
  theme_minimal(base_size = 12)

p1 + p2
```

. . .

**This is what we want.** The residuals scatter randomly around zero with no curve or funnel. The loess smoother is basically flat — our model is capturing the real relationship.

## Multiple Predictors Revisited {.smaller}

Last time we added poverty as a control. We can add as many predictors as we want:

$$
\text{Score} = 10 + 2.4 \times \text{Mom Yrs} + 2.1 \times \text{Dad Yrs} + 4 \times \text{Fam Income} + \epsilon
$$

. . .

::: incremental
-   Each $\beta_j$ is the average change in Y for a one-unit change in $X_j$, **holding all other variables fixed**
-   Controls are a powerful tool; but doesn't mean we should throw every variable we can find into the model
-   **Bad controls introduce bias** (we'll see why shortly)
-   **Collinearity** between predictors can make the "holding constant" interpretation tricky
:::

# Part 2: The 'Add More Controls' Trap

## A Quick R-squared Refresher {.smaller}

From last time: $R^2$ measures what fraction of the variation in Y our model explains.

$$R^2 = 1 - \frac{\text{Residual SS}}{\text{Total SS}} = 1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$$

. . .

Think of it as: how much better is our model at predicting Y compared to just guessing the mean every time?

. . .

::: callout-note
Higher $R^2$ = better model fit. But there's a catch...
:::

## The R-squared Trap: A Demonstration {.smaller}

```{r}
#| echo: true

library(tidyverse)

set.seed(30317)

n <- 100
x1 <- rnorm(n)  # Meaningful predictor
noise <- rnorm(n, 0, 4)  # Pure noise — NO relationship with Y
y <- 2 + 3 * x1 + rnorm(n, 0, 2)  # True relationship only involves x1

# Fit two models
model1 <- lm(y ~ x1)
model2 <- lm(y ~ x1 + noise)

# Compare R-squared
cat("Model 1 (just x1):", round(summary(model1)$r.squared, 4), "\n")
cat("Model 2 (x1 + noise):", round(summary(model2)$r.squared, 4), "\n")
```

. . .

$R^2$ **increased** even though we added a completely meaningless variable!

## Why Does This Happen? {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 3.5

rsq1 <- summary(model1)$r.squared
rsq2 <- summary(model2)$r.squared

plot_data <- data.frame(
  Model = c("Just Spending", "Spending + Random Noise"),
  R_squared = c(rsq1, rsq2)
)

ggplot(plot_data, aes(x = Model, y = R_squared)) +
  geom_bar(stat = "identity", fill = "#522D80", alpha = 0.8, width = 0.5) +
  geom_text(aes(label = sprintf("%.4f", R_squared)), vjust = 1.5, size = 5,
            color = "white", fontface = "bold") +
  labs(title = expression(paste(R^2, " Always Increases When You Add Predictors")),
       subtitle = "Even completely random variables 'explain' some variance by chance",
       y = expression(R^2)) +
  theme_minimal(base_size = 14)
```

::: incremental
-   Each predictor "explains" some variance **by chance**, even if it has no real relationship with Y
-   OLS is a very efficient algorithm — it will find *some* pattern, even in random noise
-   This means $R^2$ alone can't tell you if a model is good
:::

## The Fix: Adjusted R-squared {.smaller}

Adjusted $R^2$ penalizes you for adding predictors that don't pull their weight:

$$\text{Adjusted } R^2 = 1 - \frac{(1 - R^2)(n - 1)}{n - p - 1}$$

. . .

::: incremental
-   $n$ = number of observations
-   $p$ = number of predictors
-   Unlike regular $R^2$, adjusted $R^2$ can actually **decrease** if you add a useless predictor
-   Use adjusted $R^2$ when comparing models with different numbers of predictors
:::

. . .

```{r}
#| echo: true

cat("Model 1 Adj R²:", round(summary(model1)$adj.r.squared, 4), "\n")
cat("Model 2 Adj R²:", round(summary(model2)$adj.r.squared, 4), "\n")
```

Adjusted $R^2$ went down, telling us the noise variable isn't helping.

## Key Takeaways on Model Fit

::: incremental
-   $R^2$ is useful but always goes up when you add predictors: don't be fooled
-   **Adjusted** $R^2$ corrects for this and is better for comparing models
-   If you care about **prediction**, cross-validation is even better (we'll discuss near end of semester)
-   If you care about **causation**, the right model is the one that matches your causal theory, so $R^2$ is less relevant
:::

# Part 3: When Regression Goes Wrong

## Three Threats {.smaller}

When we run a regression and interpret coefficients, we're implicitly making claims about relationships. Three common pitfalls:

::: incremental
1.  **Omitted Variable Bias**: Leaving out a confounder that distorts our estimate
2.  **Collider Bias**: Controlling for the *wrong* thing and creating a spurious relationship
3.  **Simpson's Paradox**: An aggregate trend that reverses when you look within groups
:::

. . .

We'll use **DAGs** (Directed Acyclic Graphs) to think through each one. DAGs are just arrow diagrams showing which variables cause which.

## Omitted Variable Bias: The DAG {.smaller}

```{r}
#| fig-width: 8
#| fig-height: 3.5
#| fig-align: center

library(dagitty)
library(ggdag)

dag_ovb <- dagitty('dag{
    "Family Income" [pos="0,0"]
    "Preschool" [pos="1,1"]
    "Graduation" [pos="2,0"]
    "Family Income" -> "Preschool"
    "Family Income" -> "Graduation"
    "Preschool" -> "Graduation"
}')

ggdag(dag_ovb, text = FALSE, use_labels = "name") +
  theme_dag() +
  coord_cartesian(clip = "off") +
  labs(title = "Omitted Variable Bias",
       subtitle = "Family Income causes BOTH Preschool attendance and Graduation") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        legend.position = "none",
        plot.margin = margin(t = 20, r = 10, b = 10, l = 10))
```

. . .

If we regress Graduation on Preschool **without** controlling for Family Income, we'll overestimate the preschool effect — because part of what we're picking up is actually the effect of income.

## OVB: Seeing It in Action {.smaller}

```{r}
#| echo: true
library(tidyverse)
library(dagitty)
library(ggdag)

set.seed(30317)
N <- 1000

# Family income (the confounder)
family_income <- rnorm(N)

# Preschool attendance (caused by income)
preschool <- 0.5 * family_income + rnorm(N)

# Graduation rate (caused by BOTH preschool and income)
graduation <- 0.4 * preschool + 0.3 * family_income + rnorm(N)

# Biased model (omitting income)
model_biased <- lm(graduation ~ preschool)

# Unbiased model (including income)
model_unbiased <- lm(graduation ~ preschool + family_income)
```

. . .

```{r}
#| fig-width: 10
#| fig-height: 4

coef_data <- data.frame(
  Model = c("Without Income\n(Biased)", "With Income\n(Unbiased)"),
  Coefficient = c(coef(model_biased)["preschool"],
                 coef(model_unbiased)["preschool"])
)

ggplot(coef_data, aes(x = Model, y = Coefficient)) +
  geom_bar(stat = "identity", fill = "#522D80", alpha = 0.8, width = 0.5) +
  geom_hline(yintercept = 0.4, color = "#F66733", linetype = "dashed", linewidth = 1) +
  geom_text(aes(label = sprintf("%.3f", Coefficient)), vjust = 1.5, size = 5,
            color = "white", fontface = "bold") +
  annotate("text", x = 2.4, y = 0.41, label = "True Effect = 0.4",
           color = "#F66733", fontface = "italic", size = 4) +
  labs(title = "Omitted Variable Bias in Action",
       subtitle = "Leaving out Family Income inflates the Preschool coefficient",
       y = "Estimated Effect of Preschool") +
  theme_minimal(base_size = 14)
```

We *know* the true effect is 0.4 (we simulated it!), but omitting income makes preschool look more impactful than it really is.

## What Makes Something an OVB Threat? {.smaller}

Not every omitted variable creates bias. For OVB to occur, the omitted variable must:

::: incremental
1.  Be correlated with the **predictor** (X): Family income affects who goes to preschool
2.  **Cause** the **outcome** (Y): Family income directly affects graduation rates
3.  Not be caused by X: If the omitted variable is *caused* by your predictor, including it would be a different kind of problem...
:::

. . .

::: callout-warning
## Common Misconception

It's not enough for a variable to be correlated with X and Y. It must **cause** Y. Correlation alone can mislead us.
:::

## Post-Treatment Bias {.smaller}

Consider another potential control. Some argue that *natural aptitude* also contributes to graduation rates, and we should capture this by controlling for **reading scores in first grade**.

. . .

Should we control for first-grade reading scores in our model?

. . .

**No!** Preschool might *cause* higher reading scores. If we control for reading scores, we're "blocking" part of the very effect we're trying to measure.

. . .

::: callout-important
## Rule of Thumb

Only control for variables that are **not caused by your treatment**. If the treatment can influence the control variable, including it will bias your estimate. This is called **post-treatment bias**.
:::

## Collider Bias: A Different Problem Entirely {.smaller}

OVB is about *leaving out* the right variable. Collider bias is about *including* the wrong one.

::: incremental
-   A **collider** is a variable that is *caused by* both X and Y
-   Conditioning on a collider (either by including it in the model or by sample selection) creates a spurious relationship between X and Y
-   Classic example: does being tall make you a worse free-throw shooter?
:::

## Shaq Example {.smaller}

```{r}
#| fig-width: 8
#| fig-height: 3.5
#| fig-align: center

dag_nba <- dagitty('dag{
    Height [pos="0,1"]
    "Free Throw %" [pos="2,1"]
    "NBA Status" [pos="1,0"]
    Height -> "NBA Status"
    "Free Throw %" -> "NBA Status"
}')

ggdag(dag_nba, text = FALSE, use_labels = "name") +
  theme_dag() +
  coord_cartesian(clip = "off") +
  labs(title = "NBA Status Is a Collider",
       subtitle = "Both Height and Free Throw % affect NBA chances") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        legend.position = "none",
        plot.margin = margin(t = 20, r = 10, b = 10, l = 10))
```

. . .

::: incremental
-   Height helps you make the NBA
-   Free throw shooting helps you make the NBA
-   **But height and FT% have no direct relationship** in the general population
:::

## In the General Population {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

set.seed(30317)
N <- 10000

height <- rnorm(N, mean = 74, sd = 4)  # inches
ft_pct <- rnorm(N, mean = 70, sd = 15)

# NBA selection depends on both
nba_prob <- plogis(-2 + 0.25 * (height - 74) + 0.05 * (ft_pct - 70))
made_nba <- rbinom(N, 1, nba_prob)

basketball_data <- data.frame(height, ft_pct, made_nba)
nba_data <- basketball_data |> filter(made_nba == 1)

ggplot(basketball_data, aes(x = height, y = ft_pct)) +
  geom_point(alpha = 0.1, color = "#522D80") +
  geom_smooth(method = "lm", color = "#F66733", linewidth = 1.2) +
  labs(title = "Height vs Free Throw % (All Basketball Players)",
       subtitle = paste("Correlation:",
                       round(cor(basketball_data$height,
                               basketball_data$ft_pct), 3)),
       x = "Height (inches)", y = "Free Throw %") +
  theme_minimal(base_size = 14)
```

No relationship — as expected. Being tall doesn't make you a better or worse shooter.

## But If We Only Look at NBA Players... {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

ggplot(nba_data, aes(x = height, y = ft_pct)) +
  geom_point(alpha = 0.4, color = "#522D80") +
  geom_smooth(method = "lm", color = "#F66733", linewidth = 1.2) +
  labs(title = "Height vs Free Throw % (NBA Players Only)",
       subtitle = paste("Correlation:",
                       round(cor(nba_data$height, nba_data$ft_pct), 3)),
       x = "Height (inches)", y = "Free Throw %") +
  theme_minimal(base_size = 14)
```

Now it looks like taller players are *worse* shooters! But this is an illusion created by **conditioning on the collider**.

## Why Does This Happen? {.smaller}

Think about what it takes to make the NBA:

::: incremental
-   **Shorter players** need to be exceptional shooters to compensate — think about how good a 5'5" NBA player would have to be!
-   **Taller players** can make it with weaker shooting — think Shaq
-   So among NBA players, height and shooting are **negatively correlated** — not because one causes the other, but because NBA selection (the collider) creates this pattern
:::

## Simpson's Paradox: A Real-World Trap {.smaller}

One of the most dangerous ways aggregate data can mislead us comes from the COVID-19 pandemic.

. . .

In 2021, news outlets reported a startling statistic: in some countries, the **death rate among vaccinated people was higher** than among the unvaccinated.

. . .

Anti-vaccine groups seized on this as proof that vaccines were harmful. But the data told the opposite story once you looked more carefully.

. . .

::: callout-warning
## The Paradox

A trend that appears in aggregate data can reverse — or even disappear — when the data is split into meaningful groups.
:::

## Setting Up the Vaccine Example {.smaller}

```{r}
#| echo: true

set.seed(2021)
N <- 10000

# Age group: older people have higher death rates AND higher vaccination rates
age_group <- sample(c("Under 50", "50-69", "70+"), N,
                    prob = c(0.5, 0.3, 0.2), replace = TRUE)

# Pre-existing conditions: more common in older people
has_condition <- case_when(
  age_group == "Under 50" ~ rbinom(N, 1, 0.10),
  age_group == "50-69"    ~ rbinom(N, 1, 0.30),
  age_group == "70+"      ~ rbinom(N, 1, 0.55)
)

# Vaccination: older people and those with conditions vaccinated at MUCH higher rates
vax_prob <- case_when(
  age_group == "Under 50" & has_condition == 0 ~ 0.30,
  age_group == "Under 50" & has_condition == 1 ~ 0.50,
  age_group == "50-69"    & has_condition == 0 ~ 0.70,
  age_group == "50-69"    & has_condition == 1 ~ 0.85,
  age_group == "70+"      & has_condition == 0 ~ 0.85,
  age_group == "70+"      & has_condition == 1 ~ 0.92
)
vaccinated <- rbinom(N, 1, vax_prob)
```

## Setting Up the Vaccine Example (cont.) {.smaller}

```{r}
#| echo: true

# Death probability: age and conditions increase risk; vaccine REDUCES risk
base_death <- case_when(
  age_group == "Under 50" ~ 0.001,
  age_group == "50-69"    ~ 0.015,
  age_group == "70+"      ~ 0.080
)

death_prob <- base_death *
  (1 + 2.0 * has_condition) *   # Pre-existing conditions double+ risk
  (1 - 0.85 * vaccinated)       # Vaccine reduces death risk by 85%

died <- rbinom(N, 1, pmin(death_prob, 1))

covid_data <- tibble(
  age_group = factor(age_group, levels = c("Under 50", "50-69", "70+")),
  has_condition = has_condition,
  vaccinated = factor(vaccinated, labels = c("Unvaccinated", "Vaccinated")),
  died = died
)
```

. . .

We built in an **85% reduction in death risk** from vaccination. The vaccine clearly works in our simulation. But watch what happens in the aggregate...

## The Misleading Aggregate {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4
#| fig-align: center

overall_rates <- covid_data |>
  group_by(vaccinated) |>
  summarize(deaths = sum(died), n = n(),
            death_rate = deaths / n * 100, .groups = "drop")

ggplot(overall_rates, aes(x = vaccinated, y = death_rate, fill = vaccinated)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = sprintf("%.2f%%\n(%d / %d)", death_rate, deaths, n)),
            vjust = 1.3, size = 5, color = "white", fontface = "bold") +
  scale_fill_manual(values = c("Unvaccinated" = "#522D80", "Vaccinated" = "#F66733")) +
  labs(title = "Overall Death Rates: Vaccinated vs. Unvaccinated",
       subtitle = "Wait... vaccinated people die at a HIGHER rate?",
       x = "", y = "Death Rate (%)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

. . .

If you stopped here, you'd conclude vaccines are harmful. But this is **Simpson's Paradox** in action.

## Now Break It Down by Age Group {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4
#| fig-align: center

age_rates <- covid_data |>
  group_by(age_group, vaccinated) |>
  summarize(deaths = sum(died), n = n(),
            death_rate = deaths / n * 100, .groups = "drop")

ggplot(age_rates, aes(x = vaccinated, y = death_rate, fill = vaccinated)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = sprintf("%.2f%%", death_rate)),
            position = position_dodge(width = 0.6), vjust = 1.3, size = 3.5,
            color = "white", fontface = "bold") +
  facet_wrap(~ age_group, scales = "free_y") +
  scale_fill_manual(values = c("Unvaccinated" = "#522D80", "Vaccinated" = "#F66733")) +
  labs(title = "Death Rates by Age Group: The Real Story",
       subtitle = "Within every age group, vaccination dramatically reduces death rates",
       x = "", y = "Death Rate (%)", fill = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

. . .

**Within every age group**, vaccinated people die at a much lower rate. The vaccine works — the aggregate just hid it.

## Why Does Simpson's Paradox Happen Here? {.smaller}

::: incremental
-   **Older people** have much higher baseline death rates (8% vs. 0.1%)
-   **Older people** were vaccinated at much higher rates (85-92% vs. 30%)
-   So the vaccinated group is **disproportionately old**
-   When you compare all vaccinated to all unvaccinated, you're really comparing mostly-elderly to mostly-young
:::

## The Composition Effect {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4.5
#| fig-align: center

vax_age <- covid_data |>
  group_by(vaccinated, age_group) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(vaccinated) |>
  mutate(pct = n / sum(n) * 100)

ggplot(vax_age, aes(x = vaccinated, y = pct, fill = age_group)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", pct)),
            position = position_stack(vjust = 0.5), size = 5,
            color = "white", fontface = "bold") +
  scale_fill_manual(values = c("Under 50" = "#8FD694", "50-69" = "#F66733", "70+" = "#522D80")) +
  labs(title = "Age Composition of Each Group",
       subtitle = "The vaccinated group is much older — that's why the aggregate is misleading",
       x = "", y = "Percent", fill = "Age Group") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

## The Lesson for Policy Analysts {.smaller}

::: callout-tip
## When You See a Surprising Aggregate Result

Always ask: "Could this be driven by group composition?" If the groups differ on both the treatment AND the outcome, the aggregate trend can be misleading.
:::

. . .

This isn't just academic. Simpson's Paradox with COVID vaccines:

::: incremental
-   Fueled vaccine hesitancy and cost lives
-   Was used (intentionally or not) to mislead the public
-   Could have been avoided by reporting age-stratified data from the start
:::

. . .

The fix is the same as always: **control for the confounder**. In a regression, adding age group as a predictor reveals the true protective effect of vaccination.

## Outliers and Influence {.smaller}

One more diagnostic: some data points have outsized impact on the regression line.

```{r}
#| fig-width: 10
#| fig-height: 4

set.seed(30317)
n_main <- 40
x_main <- rnorm(n_main, mean = 0, sd = 1)
y_main <- 2 + 3 * x_main + rnorm(n_main, sd = 1)

x_high_off <- c(8, 10)
y_high_off <- c(40, -5)

x_youtlier <- 0
y_youtlier <- 50

x <- c(x_main, x_high_off, x_youtlier)
y <- c(y_main, y_high_off, y_youtlier)
df <- data.frame(x = x, y = y)

model <- lm(y ~ x, data = df)
df$cooks_dist <- cooks.distance(model)
cook_cutoff <- 4 / nrow(df)
df$high_influence <- df$cooks_dist > cook_cutoff

ggplot(df, aes(x = x, y = y)) +
  geom_point(aes(color = high_influence), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#522D80", linewidth = 1.2) +
  scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "#F66733"),
                     labels = c("Normal", "High Influence")) +
  labs(title = "Some Points Pull the Line More Than Others",
       subtitle = paste("Points with Cook's Distance >",
                       round(cook_cutoff, 3), "are flagged"),
       x = "X", y = "Y", color = "Influence") +
  theme_minimal(base_size = 14)
```

## Dealing with Influential Points {.smaller}

::: incremental
-   **Don't automatically delete outliers**: they might be real and important!
-   **Do check** if the data was entered correctly (typos, coding errors)
-   **Do exclude** values that are impossible (voter turnout \> 100%, negative ages)
-   **Cook's distance** is a common measure of influence — how much the regression line changes if you drop that one point
    -   Rule of thumb: Cook's D \> 4/n is worth investigating
-   **Report both**: run the analysis with and without influential points and see if your conclusions change
    -   This is often called the "jacknife" approach
:::

# Part 4: Interactions

## When One Size Doesn't Fit All {.smaller}

So far, our models assume the effect of spending is the **same** for every school district, regardless of poverty level.

. . .

But what if the effect of funding **depends** on school characteristics?

. . .

::: incremental
-   Maybe an extra \$1,000 matters more for a school in a high-poverty district than a wealthy one
-   Maybe the return on funding levels off for the richest districts
-   These are **conditional hypotheses** - we need **interaction terms** to test them
:::

## The Policy Question

::: incremental
-   How does funding affect student achievement?
-   Does the effect differ between high-poverty and low-poverty schools?
-   Basic regression: `Achievement ~ Funding + Poverty`
-   But what if the effect is **not the same** across school types?
-   **Interaction model**: `Achievement ~ Funding * Poverty`
:::

## Setting Up the Data {.smaller}

```{r}
#| echo: true

library(tidyverse)

set.seed(30317)
n <- 1000

funding <- rnorm(n, 5000, 1000)
high_poverty <- rbinom(n, 1, 0.3)  # 30% of schools are high poverty

sim_data <- data.frame(
  high_poverty = high_poverty,
  funding = funding,
  achievement = 70 + 0.0013 * funding +      # Base funding effect
    -20 * high_poverty +                       # Poverty penalty
    0.0024 * funding * high_poverty +          # ADDITIONAL effect in high poverty schools
    rnorm(n, 0, 2)
)
```

. . .

Notice what we built in: funding has a base effect (0.0013 per dollar) that is **larger** for high-poverty schools (an additional 0.0024 per dollar).

## Visualizing: One Line vs. Two {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4

p1 <- ggplot(sim_data, aes(x = funding, y = achievement)) +
  geom_point(aes(color = factor(high_poverty)), alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1.2) +
  scale_color_manual(values = c("#522D80", "#F66733"),
                     labels = c("Low Poverty", "High Poverty"),
                     name = "School Type") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "One Line for Everyone",
       x = "Per-pupil Funding", y = "Achievement") +
  theme_minimal(base_size = 12)

p2 <- ggplot(sim_data, aes(x = funding, y = achievement,
                            color = factor(high_poverty))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(values = c("#522D80", "#F66733"),
                     labels = c("Low Poverty", "High Poverty"),
                     name = "School Type") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Separate Lines by Poverty",
       x = "Per-pupil Funding", y = "Achievement") +
  theme_minimal(base_size = 12)

p1 + p2
```

The right panel tells a different story — funding has a **steeper slope** for high-poverty schools.

## The Interaction Model in R {.smaller}

```{r}
#| echo: true

# Model WITHOUT interaction
model_no_int <- lm(achievement ~ funding + high_poverty, data = sim_data)

# Model WITH interaction
model_int <- lm(achievement ~ funding + high_poverty + funding:high_poverty,
                data = sim_data)
summary(model_int)
```

Note: `funding:high_poverty` is the interaction term. You can also write `funding * high_poverty`, which automatically includes both main effects and the interaction.

## Interpreting the Interaction Model {.smaller}

The full model:

$$
Y = \beta_0 + \beta_1 \times \text{Funding} + \beta_2 \times \text{High Poverty} + \beta_3 \times \text{Funding} \times \text{High Poverty}
$$

. . .

::: incremental
-   $\beta_0$: Achievement when funding = 0 and poverty = 0 (not very interpretable)
-   $\beta_1$: Effect of each dollar of funding **when poverty = 0** (low-poverty schools)
-   $\beta_2$: Difference in achievement for high-poverty schools **when funding = 0** (also not very interpretable)
-   $\beta_3$: The **additional** effect of each dollar of funding for high-poverty schools
:::

. . .

So the **total effect** of funding for high-poverty schools is $\beta_1 + \beta_3$.

## What's Happening Under the Hood {.smaller}

Again, you don't need calculus - but it's nice to see why the coefficients work the way they do. Start with the model:

$$
Y = \beta_0 + \beta_1 \text{Funding} + \beta_2 \text{Poverty} + \beta_3 (\text{Funding} \times \text{Poverty})
$$

. . .

Now take the partial derivative with respect to Funding:

$$
\frac{\partial Y}{\partial \text{Funding}} = \beta_1 + \beta_3 \times \text{Poverty}
$$

. . .

That's it. The effect of funding **depends on the value of poverty**. The $\beta_3$ term is what makes it vary: without it, we'd just get $\beta_1$ for everyone.

. . .

::: incremental
-   When Poverty = 0: effect of funding = $\beta_1$ (just the base slope)
-   When Poverty = 1: effect of funding = $\beta_1 + \beta_3$ (base + additional)
-   The partial derivative is just formalizing what the coefficient table already told us
:::

## Don't Over-Interpret the Coefficients! {.smaller}

This is the most common mistake with interaction models:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2
$$

::: incremental
-   $\beta_1$ is **NOT** the average effect of $X_1$ on Y
-   $\beta_2$ is **NOT** the average effect of $X_2$ on Y
-   Instead, $\beta_1$ is the effect of $X_1$ **when** $X_2 = 0$
-   And $\beta_2$ is the effect of $X_2$ **when** $X_1 = 0$
-   To get the effect at other values, you need **marginal effects**
:::

## Continuous Interactions {.smaller}

Our poverty variable doesn't have to be binary. What if we use the actual poverty rate (0–100%)?

```{r}
#| echo: true

set.seed(30317)
n <- 1000

funding <- rnorm(n, 5000, 1000)
poverty <- runif(n, 0, 1)  # poverty rate from 0 to 100%

sim_data_cont <- data.frame(
  poverty = poverty,
  funding = funding,
  achievement = 70 + 0.001 * funding +
    -25 * poverty +
    0.002 * funding * poverty +
    rnorm(n, 0, 5)
)

model_cont <- lm(achievement ~ funding * poverty, data = sim_data_cont)
```

Now the effect of funding changes **continuously** with the poverty rate:

$$
\frac{\partial Y}{\partial \text{Funding}} = \beta_1 + \beta_3 \times \text{Poverty}
$$

## Visualizing the Continuous Interaction {.smaller}

```{r}
#| fig-width: 10
#| fig-height: 4

ggplot(sim_data_cont, aes(x = funding, y = achievement)) +
  geom_point(aes(color = poverty), alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  scale_color_viridis_c(name = "Poverty Rate", labels = scales::percent) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Effect of School Funding on Achievement",
       subtitle = "Color indicates school poverty rate",
       x = "Per-pupil Funding", y = "Achievement Score") +
  theme_minimal(base_size = 14)
```

The overall line hides an important story — the effect of funding varies by poverty level.

## The Continuous Interaction Output {.smaller}

```{r}
#| echo: true
summary(model_cont)
```

## When to Include Interactions {.smaller}

Any time you have a **conditional hypothesis**: a theory that says the effect of X depends on some other variable:

::: incremental
-   "Funding will have a **greater impact** on student achievement when poverty is higher"
-   "The effect of free trade agreements on voting depends on **how exposed** the voter's industry is to trade"
-   "Campaign spending matters more in **competitive** districts than in safe ones"
:::

. . .

::: callout-tip
## Theory First

Don't include interactions just because you can. Need a theory about *why* the effect should vary.
:::

# Part 5: Marginal Effects

## Why We Need Marginal Effects {.smaller}

In an interaction model, the regression table doesn't directly give us the effect of funding at different poverty levels. We need to calculate **marginal effects**.

. . .

From our model:

$$\frac{\partial Y}{\partial \text{Funding}} = \beta_1 + \beta_3 \times \text{Poverty}$$

. . .

::: incremental
-   At poverty = 0%: effect of funding = $\beta_1$
-   At poverty = 50%: effect of funding = $\beta_1 + 0.5 \times \beta_3$
-   At poverty = 100%: effect of funding = $\beta_1 + \beta_3$
-   We can calculate this for any poverty level!
:::

## Computing Marginal Effects Manually {.smaller}

```{r}
#| echo: true

# Get coefficients
coefs <- coef(model_cont)

# Calculate the effect of funding at different poverty levels
marginal_effects <- data.frame(
  poverty = seq(0, 1, by = 0.1)
) |>
  mutate(
    effect = coefs["funding"] + coefs["funding:poverty"] * poverty
  )

marginal_effects
```

. . .

As poverty increases, the effect of funding gets larger — exactly what we built into the simulation.

## The `marginaleffects` Package {.smaller}

Computing SEs for marginal effects by hand is tricky. The `marginaleffects` package does this for us:

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 3.5

library(marginaleffects)

# Calculate and plot marginal effects of funding across poverty levels
plot_slopes(model_cont,
           variables = "funding",
           condition = "poverty") +
  labs(title = "Marginal Effect of Funding on Achievement",
       subtitle = "Effect increases with poverty rate",
       x = "Poverty Rate",
       y = "Marginal Effect of $1 in Funding") +
  theme_minimal(base_size = 14)
```

The shaded band is a 95% confidence interval. Where the band includes zero, the effect is not statistically significant at that poverty level.

## Key Rules for Marginal Effects

::: incremental
-   **Always** report marginal effects when you have interactions: don't just report the regression table
-   Calculate them at **meaningful values** of the moderator (e.g., low/medium/high poverty, not just 0)
-   **Include confidence intervals**: the standard errors on the coefficients alone can be misleading
-   The `marginaleffects` package makes this easy: `plot_slopes()` for visualization, `slopes()` for tables
:::

# Wrapping Up

## What We've Covered Today

::: incremental
1.  **Coefficient interpretation** goes beyond "slope": binary predictors give differences in means, continuous predictors give rates of change, and quadratic terms allow curvature
2.  **R-squared always increases** when you add predictors, use adjusted R-squared to compare models fairly
3.  **Omitted variable bias** inflates estimates when confounders are left out: Use DAGs to think through what to control for
4.  **Collider bias** creates spurious relationships when you control for the wrong thing
5.  **Interactions** let us model effects that vary + **marginal effects** are the right way to interpret them
:::

## The Bias Cheat Sheet {.smaller}

| Problem | What Happened | Example | Fix |
|------------------|-------------------|------------------|------------------|
| **OVB** | Left out a confounder | Preschool effect without income | Include the confounder |
| **Post-treatment** | Controlled for an outcome of treatment | Preschool effect controlling for reading scores | Don't control for post-treatment variables |
| **Collider** | Conditioned on a shared effect | NBA height/shooting | Don't control for (or select on) colliders |
| **Simpson's** | Aggregate hides group-level patterns | COVID vaccines by age | Analyze within groups |

## Looking Ahead

::: incremental
-   So far we've been describing **associations** and trying to reduce bias through controls
-   But the gold standard for causal inference is **research design** is structuring the study so that confounding is eliminated, not just reduced
-   Coming up: **randomized experiments**, **difference-in-differences**, and **regression discontinuity**
-   These tools are the backbone of rigorous policy evaluation
:::

## Your Turn: Practice Problem {.smaller}

A researcher wants to know if police body cameras reduce use-of-force incidents. They run:

$$\text{Force Incidents} = \beta_0 + \beta_1 \times \text{Cameras} + \beta_2 \times \text{City Size} + \epsilon$$

. . .

Think about:

::: incremental
1.  What confounders might be missing? (Think about what types of cities adopt cameras early)
2.  Would you add an interaction term? If so, which one and why?
3.  Could there be a collider problem? (What if we only look at departments under federal oversight?)
4.  If you found that cameras *increase* force incidents, what might explain that?
:::