---
title: "Introduction to Regression"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
---

## Where We Are {.smaller}

Now we know how to:

-   Test whether a result could be random noise (hypothesis testing)
-   Quantify how precise our estimates are (confidence intervals)
-   Compare two groups with t-tests

. . .

What if we want to go beyond comparing two groups?

-   Does higher per-pupil **spending** predict better **graduation rates**?
-   Can we **account for** other factors, like district poverty, at the same time?

## When t-tests Aren't Enough

t-tests compare two groups: treatment vs. control, Men vs Women, etc.

. . .

But many policy questions involve **continuous** predictors:

-   How does *each additional dollar* of spending relate to graduation rates?
-   What's the relationship between *officer staffing levels* and *crime rates*?
-   Does *distance to a polling place* affect *turnout*?

. . .

**Regression** lets us model these relationships — and control for confounding factors.

## The Big Idea: Conditional Expectation

Remember the conditional expectation function (CEF) from Lecture 5?

$$E[Y | X = x]$$

"What's the average value of Y for observations where X equals x?"

. . .

In Lecture 5, we asked: What's the average wait time for urban vs. rural residents?

. . .

**Regression** takes this further: instead of comparing group averages, it models *how the average changes continuously* as X increases.

# Simple Linear Regression

## Starting with a Picture

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

library(tidyverse)

set.seed(42)

# Simulate realistic district-level data
n <- 200
poverty_rate <- runif(n, 5, 40)
spending <- 8 + 0.15 * (50 - poverty_rate) + rnorm(n, 0, 2)  # spending in thousands
grad_rate <- 60 + 1.8 * spending - 0.4 * poverty_rate + rnorm(n, 0, 5)

districts <- tibble(
  spending = round(spending, 1),
  grad_rate = pmin(round(grad_rate, 1), 99),
  poverty_rate = round(poverty_rate, 1)
)

ggplot(districts, aes(x = spending, y = grad_rate)) +
  geom_point(alpha = 0.6, color = "#522D80", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "#F66733", linewidth = 1.5) +
  labs(
    title = "Per-Pupil Spending and Graduation Rates Across 200 Districts",
    x = "Per-Pupil Spending (thousands of $)",
    y = "Graduation Rate (%)"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

Each dot is a school district. The line summarizes: "On average, how do graduation rates change as spending increases?"

## The Linear Regression Model {.smaller}

We assume the relationship between X and Y can be described by a line:

$$Y = \beta_0 + \beta_1 X + \epsilon$$

. . .

-   **Y**: The outcome (graduation rate)
-   **X**: The predictor (per-pupil spending)
-   $\beta_0$ (intercept): The predicted graduation rate when spending = 0
-   $\beta_1$ (slope): How much the graduation rate changes, on average, for each additional \$1,000 in spending
-   $\epsilon$ (error): Everything else that affects graduation rates but isn't in our model

## How Does R Find the Line? {.smaller}

R finds the line that minimizes the **sum of squared residuals** — the total squared distance between each point and the line.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5

model_simple <- lm(grad_rate ~ spending, data = districts)

districts_with_pred <- districts |>
  mutate(predicted = predict(model_simple))

# Show just 15 points for clarity
set.seed(10)
sample_districts <- districts_with_pred |> slice_sample(n = 15)

ggplot(sample_districts, aes(x = spending, y = grad_rate)) +
  geom_segment(aes(xend = spending, yend = predicted),
               color = "#F66733", alpha = 0.6, linewidth = 0.8) +
  geom_point(size = 3, color = "#522D80") +
  geom_smooth(data = districts, method = "lm", se = FALSE,
              color = "#522D80", linewidth = 1.2) +
  labs(
    title = "Ordinary Least Squares: Minimizing the Residuals",
    subtitle = "Orange lines = residuals (what the model gets wrong)",
    x = "Per-Pupil Spending (thousands of $)",
    y = "Graduation Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

The orange lines are **residuals** — the gap between what we predicted and what actually happened. OLS finds the line that makes these gaps as small as possible (in total).

## Running Regression in R: `lm()`

```{r}
#| echo: true

# Run a simple regression: graduation rate predicted by spending
model_simple <- lm(grad_rate ~ spending, data = districts)
```

. . .

The formula `grad_rate ~ spending` means: "Predict graduation rate from per-pupil spending."

. . .

This is the same `~` syntax you've seen in `t.test()`. In R, `~` always means "as a function of."

## The Output: Don't Panic {.smaller}

```{r}
#| echo: true

summary(model_simple)
```

## Reading the Output: What to Focus On {.smaller}

That's a lot of numbers. For now, focus on **three things**:

. . .

**1. The coefficients table** (middle section):

| Term | Estimate | What it means |
|-------------------|----------------------|--------------------------------|
| (Intercept) | `r round(coef(model_simple)[1], 1)` | Predicted graduation rate if spending were \$0 (not meaningful here) |
| spending | `r round(coef(model_simple)[2], 1)` | Each additional \$1,000 in spending is associated with a `r round(coef(model_simple)[2], 1)` percentage point increase in graduation rate |

. . .

**2. The p-values** (column `Pr(>|t|)`): Is each coefficient significantly different from zero?

. . .

**3. R-squared** (bottom): How much variation in graduation rates does spending explain?

## Interpreting the Slope

Our model says:

$$\text{Grad Rate} = `r round(coef(model_simple)[1], 1)` + `r round(coef(model_simple)[2], 1)` \times \text{Spending}$$

. . .

**In words:** For every additional \$1,000 in per-pupil spending, graduation rates are about `r round(coef(model_simple)[2], 1)` percentage points higher, on average.

. . .

::: callout-warning
**Caution:** This is an *association*, not a causal effect! Districts that spend more may also differ in other ways (wealthier families, more experienced teachers, etc.). We're describing a pattern — not proving that spending *causes* higher graduation rates.
:::

## What About the Intercept?

The intercept (`r round(coef(model_simple)[1], 1)`) is the predicted graduation rate when spending = \$0.

. . .

**Does this make sense?** Not really — no district spends \$0 per pupil.

. . .

::: callout-tip
## Rule of Thumb

The intercept is often not substantively interesting. It anchors the line mathematically, but don't over-interpret it — especially when X = 0 is outside the range of your data.
:::

## R-squared: How Explanatory Is Our Model? {.smaller}

**R-squared** ($R^2$) tells us what fraction of the variation in Y our model explains.

. . .

Our model: $R^2$ = `r round(summary(model_simple)$r.squared, 2)`

. . .

**Translation:** Spending explains about `r round(summary(model_simple)$r.squared * 100)`% of the variation in graduation rates across districts.

. . .

That means the other `r round((1 - summary(model_simple)$r.squared) * 100)`% is driven by factors we haven't included — poverty, teacher quality, curriculum, demographics, and so on.

. . .

::: callout-note
A low R-squared doesn't mean the model is useless. It just means there's a lot of unexplained variation — which is normal in social science. Even small R-squared values can reveal important relationships.
:::

## How R-squared Is Calculated {.smaller}

R-squared compares two things: how much graduation rates vary *overall*, and how much they vary *after* our model has had its say.

. . .

We start by decomposing the total variation in Y into two pieces:

$$\underbrace{\sum(Y_i - \bar{Y})^2}_{\text{Total SS}} = \underbrace{\sum(\hat{Y}_i - \bar{Y})^2}_{\text{Model SS}} + \underbrace{\sum(Y_i - \hat{Y}_i)^2}_{\text{Residual SS}}$$

. . .

-   **Total SS:** How spread out graduation rates are around their overall mean
-   **Model SS:** How much of that spread our regression line accounts for
-   **Residual SS:** What's left over — the stuff the model can't explain

. . .

Then:

$$R^2 = \frac{\text{Model SS}}{\text{Total SS}} = 1 - \frac{\text{Residual SS}}{\text{Total SS}}$$

It's just the share of total variation that the model captures.

## What Are Residuals? {.smaller}

A residual is just the difference between what actually happened and what our model predicted:

$$e_i = Y_i - \hat{Y}_i$$

. . .

-   $Y_i$ is the actual graduation rate for district $i$
-   $\hat{Y}_i$ is what our regression line predicted for that district
-   $e_i$ is the residual — positive means the model underpredicted, negative means it overpredicted

. . .

If the model were perfect, every residual would be zero. They won't be — but we want them to be small and random, with no pattern.

## Computing Residuals in R {.smaller}

R makes this easy. Once you've fit a model with `lm()`, you can get the predicted values and residuals directly:

```{r}
#| echo: true

# Add predicted values and residuals to our data
districts_with_pred <- districts |>
  mutate(
    predicted = predict(model_simple),
    residual = grad_rate - predicted
  )

# Look at the first few rows
districts_with_pred |>
  select(spending, grad_rate, predicted, residual) |>
  head(5)
```

. . .

You can also use `resid(model_simple)` as a shortcut — it gives you the same residual column directly.

## The Residual Plot {.smaller}

Now we plot the residuals against the predicted values. This is the single most useful diagnostic plot for regression.

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4

ggplot(districts_with_pred, aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.5, color = "#522D80", size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "#F66733", linewidth = 1) +
  labs(
    x = "Predicted Graduation Rate (%)",
    y = "Residual (Actual - Predicted)"
  ) +
  theme_minimal(base_size = 14)
```

## Reading a Residual Plot {.smaller}

What are we looking for?

. . .

-   **Random scatter around zero:** Good: the model's errors aren't systematic
-   **A funnel shape (wider on one side):** Bad: that's heteroskedasticity (remember Lecture 5?). The model is less reliable for some districts than others.
-   **A curve or pattern:** Bad: the relationship probably isn't linear, and our straight line is missing something

. . .

Our plot looks pretty good, no obvious patterns, roughly even spread. That's what we want to see. But what does it look like when things go wrong?

## A Concerning Residual Plot: Non-Linearity {.smaller}

Here's a case where we fit a straight line to data that's actually curved. Imagine a program where small class sizes help a lot, but the benefit levels off: a linear model would miss this:

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

set.seed(123)
n <- 200
x_class <- runif(n, 10, 40)
# True relationship is curved: y = 95 - 30*log(x/10)
y_scores <- 95 - 30 * log(x_class / 10) + rnorm(n, 0, 3)

bad_model <- lm(y_scores ~ x_class)

library(patchwork)

p1 <- ggplot(tibble(x = x_class, y = y_scores), aes(x, y)) +
  geom_point(alpha = 0.5, color = "#522D80", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#F66733", linewidth = 1.2) +
  labs(title = "The Data (with a straight line forced through it)",
       x = "Class Size", y = "Test Scores") +
  theme_minimal(base_size = 13)

p2 <- ggplot(tibble(predicted = predict(bad_model),
                     residual = resid(bad_model)), aes(predicted, residual)) +
  geom_point(alpha = 0.5, color = "#522D80", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#F66733", linewidth = 1) +
  labs(title = "The Residual Plot",
       x = "Predicted Test Score", y = "Residual") +
  theme_minimal(base_size = 13)

p1 + p2
```

See the U-shape in the residuals? The model overpredicts in the middle and underpredicts at the extremes.

## A Concerning Residual Plot: Heteroskedasticity {.smaller}

Now a different problem. Imagine we're looking at 911 call volume across cities of different sizes. The relationship is linear, but the *spread* gets wider for bigger cities.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

set.seed(42)
n <- 300
city_pop <- runif(n, 5, 500)
calls_911 <- 2 + 0.8 * city_pop + rnorm(n, 0, city_pop * 0.15)

het_model <- lm(calls_911 ~ city_pop)

p1 <- ggplot(tibble(x = city_pop, y = calls_911), aes(x, y)) +
  geom_point(alpha = 0.5, color = "#522D80", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#F66733", linewidth = 1.2) +
  labs(title = "The Data (line fits OK, but look at the spread)",
       x = "City Population (thousands)", y = "Annual 911 Calls (thousands)") +
  theme_minimal(base_size = 13)

p2 <- ggplot(tibble(predicted = predict(het_model),
                     residual = resid(het_model)), aes(predicted, residual)) +
  geom_point(alpha = 0.5, color = "#522D80", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#F66733", linewidth = 1) +
  labs(title = "The Residual Plot",
       x = "Predicted 911 Calls", y = "Residual") +
  theme_minimal(base_size = 13)

p1 + p2
```

The line itself might be fine, but our standard errors are wrong — we'd need **robust standard errors** to fix that.

# But Wait: Correlation Is Not Causation

## The Problem with Simple Regression {.smaller}

We found that higher spending is associated with higher graduation rates. But consider:

. . .

-   **Wealthier districts** tend to spend more on schools
-   **Wealthier districts** also have lower poverty rates, more parental involvement, better health outcomes...
-   So is it the *spending* that matters, or the *wealth*?

. . .

```{r}
#| echo: false
#| fig-width: 8.5
#| fig-height: 3

library(patchwork)

p1 <- ggplot(districts, aes(x = poverty_rate, y = spending)) +
  geom_point(alpha = 0.5, color = "#522D80", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#F66733") +
  labs(title = "Poverty vs. Spending", x = "Poverty Rate (%)", y = "Spending ($K)") +
  theme_minimal(base_size = 12)

p2 <- ggplot(districts, aes(x = poverty_rate, y = grad_rate)) +
  geom_point(alpha = 0.5, color = "#522D80", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#F66733") +
  labs(title = "Poverty vs. Graduation Rate", x = "Poverty Rate (%)", y = "Graduation Rate (%)") +
  theme_minimal(base_size = 12)

p1 + p2
```

Poverty is correlated with *both* spending and graduation rates. This is **confounding**.

## The (partial) Solution: Multiple Regression

What if we could ask: "What's the relationship between spending and graduation rates *for districts with the same poverty rate*?"

. . .

This is exactly what **multiple regression** does. By including poverty as a second predictor, we can isolate spending's relationship with graduation rates while *holding poverty constant*.

. . .

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \epsilon$$

where $X_1$ = spending and $X_2$ = poverty rate.

# Multiple Regression

## Adding a Control Variable

```{r}
#| echo: true

# Multiple regression: graduation rate predicted by spending AND poverty
model_multiple <- lm(grad_rate ~ spending + poverty_rate, data = districts)
```

. . .

The formula `grad_rate ~ spending + poverty_rate` means: "Predict graduation rate from spending, *controlling for* poverty rate."

. . .

**The `+` in the formula doesn't mean addition**: it means "include both predictors in the model."

## Multiple Regression Output {.smaller}

```{r}
#| echo: true

summary(model_multiple)
```

## Reading This Output: Focus on the Same Three Things {.smaller}

**1. The coefficients table:**

| Term | Estimate | What it means |
|-------------------|----------------------|--------------------------------|
| (Intercept) | `r round(coef(model_multiple)[1], 1)` | Predicted grad rate when spending = 0 and poverty = 0 |
| spending | `r round(coef(model_multiple)[2], 1)` | Each \$1,000 more in spending predicts `r round(coef(model_multiple)[2], 1)` pp higher grad, **poverty held constant** |
| poverty_rate | `r round(coef(model_multiple)[3], 2)` | Each 1 pp increase in poverty predits a `r round(coef(model_multiple)[3], 2)` pp change in grad, **spending held constant** |

. . .

**2. The p-values:** Are the coefficients significantly different from zero?

. . .

**3. R-squared:** `r round(summary(model_multiple)$r.squared, 2)` — the model now explains `r round(summary(model_multiple)$r.squared * 100)`% of the variation (up from `r round(summary(model_simple)$r.squared * 100)`% with spending alone).

## The Key Phrase: "Holding Constant" {.smaller}

The most important change from simple to multiple regression:

. . .

**Simple regression:** "Districts that spend \$1,000 more have graduation rates that are `r round(coef(model_simple)[2], 1)` points higher, on average."

. . .

**Multiple regression:** "Among districts *with the same poverty rate*, those that spend \$1,000 more have graduation rates that are `r round(coef(model_multiple)[2], 1)` points higher, on average."

. . .

::: callout-tip
## Why This Matters

By controlling for poverty, we've isolated the spending–graduation relationship from the confounding effect of wealth. The coefficient on spending changed from `r round(coef(model_simple)[2], 1)` to `r round(coef(model_multiple)[2], 1)` — this tells us some of what looked like a "spending effect" was actually a "poverty effect."
:::

## A Visual Intuition

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

districts <- districts |>
  mutate(poverty_group = case_when(
    poverty_rate < 15 ~ "Low Poverty (<15%)",
    poverty_rate < 25 ~ "Medium Poverty (15-25%)",
    TRUE ~ "High Poverty (>25%)"
  )) |>
  mutate(poverty_group = factor(poverty_group,
    levels = c("Low Poverty (<15%)", "Medium Poverty (15-25%)", "High Poverty (>25%)")))

ggplot(districts, aes(x = spending, y = grad_rate, color = poverty_group)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("#522D80", "#F66733", "gray40")) +
  labs(
    title = "Spending vs. Graduation Rate, by Poverty Level",
    subtitle = "Multiple regression estimates the spending effect WITHIN each poverty group",
    x = "Per-Pupil Spending (thousands of $)",
    y = "Graduation Rate (%)",
    color = "District Poverty"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "bottom"
  )
```

Multiple regression is essentially comparing districts *within* the same poverty level. The slope within each group is the controlled effect of spending.

## Correlation Still Isn't Causation {.smaller}

Even with controls, we should be cautious:

. . .

::: callout-warning
## Important Limitations

-   We can only control for variables we **measure and include**
-   There could be other confounders we haven't thought of (teacher quality, parental education, school leadership...)
-   **Omitted variable bias**: leaving out important confounders can distort our estimates
-   Multiple regression reduces confounding but doesn't eliminate it
:::

. . .

**In a few weeks,** we'll learn techniques that *can* identify causal effects: randomized experiments, difference-in-differences, regression discontinuity, and instrumental variables.

# Practical Considerations

## Checking Assumptions {.smaller}

Linear regression assumes:

::: incremental
1.  **Linearity:** The relationship between X and Y is approximately linear
    1.  If it's not, we can add non-linear terms
2.  **Independence:** One district's graduation rate doesn't depend on another's
3.  **Constant variance:** The spread of residuals is roughly the same across all values of X
    1.  If not, we can use robust standard errors or other techniques (bootstrapping)
4.  **Normality of residuals:** The errors are approximately normally distributed (less important with large samples, thanks to the CLT!)
:::

. . .

You don't need to memorize proofs. But you should be able to spot when these assumptions are badly violated by looking at residual plots.

## When to Use What

| Situation | Tool | Policy Example |
|----------------------|-------------------|--------------------------------|
| Compare two groups | t-test | Did the mailer increase turnout? |
| Test if a proportion differs from a value | One-sample t-test | Is approval above 50%? |
| One continuous predictor | Simple regression | Does spending predict graduation rates? |
| Multiple predictors, control for confounders | Multiple regression | Effect of spending, controlling for poverty |

. . .

**The progression:** t-tests are a special case of regression. Everything you've learned builds toward a unified framework.

## Your Turn: What Would You Advise? {.smaller}

A state education department runs a regression and finds:

$$\text{Grad Rate} = 55 + 2.1 \times \text{Spending} - 0.6 \times \text{Poverty}$$

-   Both coefficients are statistically significant (p \< 0.01)
-   $R^2$ = 0.52

. . .

**The governor asks:** "If we increase per-pupil spending by \$2,000 statewide, will graduation rates rise by about 4 percentage points?"

. . .

Think about: Can we interpret this coefficient causally? What would you say? What would you want to know before making a recommendation?

## Summary: Your Regression Toolkit

::: incremental
1.  **Regression** models the relationship between a continuous predictor and an outcome, going beyond the two-group comparisons of t-tests

2.  **The slope (**$\beta_1$) tells you: for each 1-unit increase in X, Y changes by $\beta_1$ on average

3.  **R-squared** tells you how much variation your model explains, but low R-squared doesn't mean the model is useless

4.  **Multiple regression** lets you control for confounders by including additional predictors. The key phrase is "holding constant"

5.  **Association is not causation** — even with controls, omitted variables can bias our estimates
:::