---
title: "Panel Data, Fixed Effects, and Multilevel Models"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

## Panel Data

::: incremental
-   You download education data from the state. You notice the same schools appear year after year. You have **panel data**.

-   Panel data (or longitudinal data) tracks the same units over multiple time periods

    -   Test scores for schools measured annually

    -   State-level unemployment rates over decades

    -   A survey where the same individuals are re-interviewed each year

-   Because the same units appear multiple times, our standard OLS assumption of independent errors is violated

    -   Are independent errors plausible when the same school appears 10 times in your data?
:::

## Hierarchical Data

::: incremental
-   Sometimes data is **nested** in ways that create a hierarchy

    -   Students are nested in classrooms

    -   Classrooms are nested in schools

    -   Schools are nested in districts

-   If we don't account for this structure, our standard errors will be wrong and our estimates may be biased
:::

## Why Should MPAs Care?

::: incremental
-   Almost all policy data has one of these structures

    -   Program evaluations track participants over time (panel)

    -   Education data has students in schools in districts (hierarchical)

    -   Health data has patients in hospitals in regions (hierarchical)

-   If you ignore this structure, you will get wrong answers, sometimes dramatically wrong

-   Today we learn the workhorse tool for dealing with this: **fixed effects**
:::

## Today's Roadmap

::: incremental
1.  **The fixed effects idea** -- controlling for the unit
2.  **Visual walkthrough** -- seeing FE in action
3.  **The mechanics** -- equations, `fixest` in R, gapminder
4.  **Policy example** -- housing prices across cities
5.  **Assumptions and caveats** -- what FE can't do
6.  **Two-way fixed effects** -- and why it's tricky
:::

# Part 1: The Fixed Effects Idea

## Example: International Trade Policy {.smaller}

::: incremental
-   Imagine we want to know whether a visit from Germany's chancellor increases trade with Germany the following year

-   Global trade data is "stacked": each row is a country-year, with each country entering the data many times

-   Trade with Germany depends on culture, history, geography, resource endowments, and more

    -   This would make a very messy DAG. How can we deal with all these confounders?
:::

## The Messy DAG

```{r}
#| echo: false

library(dagitty)

g <- dagitty("
dag {
  bb=\"0,0,5,4\"

  CultureOfPopulation   [pos=\"2,0\"]
  CurrentPolitics       [pos=\"0,2\"]
  Geography            [pos=\"2,2\"]
  ChancellorVisit      [pos=\"0,3\"]
  HistoryWithGermany   [pos=\"4,2\"]
  TradeWithGermany     [pos=\"4,3\"]

  CultureOfPopulation -> ChancellorVisit
  CultureOfPopulation -> TradeWithGermany

  Geography -> ChancellorVisit
  Geography -> TradeWithGermany

  CurrentPolitics -> ChancellorVisit
  CurrentPolitics -> TradeWithGermany

  HistoryWithGermany -> TradeWithGermany
  HistoryWithGermany -> ChancellorVisit

  ChancellorVisit -> TradeWithGermany
}
")

plot(g, text = FALSE, use_labels = "name")
```

## Adding a Fixed Effect

::: incremental
-   If we add a **fixed effect for country**, we control for everything about that country that doesn't change (or changes very slowly) over time

-   Controlling for "France" implicitly controls for France's proximity to Germany, its resource endowments, its political culture, and its history with Germany

    -   All of those confounders, handled with a single control
:::

## The Clean DAG

```{r}
#| echo: false

library(dagitty)

g <- dagitty('
dag {
  CurrentPolitics      [pos="0,1"]
  ChancellorVisit      [pos="0,3"]
  TradeWithGermany     [pos="4,3"]
  Country              [pos="4,1"]

  CurrentPolitics -> ChancellorVisit
  CurrentPolitics -> TradeWithGermany

  Country         -> ChancellorVisit
  Country         -> TradeWithGermany

  ChancellorVisit -> TradeWithGermany
}
')
plot(g, text = FALSE, use_labels = "name")
```

## What Do Fixed Effects Actually Do?

::: incremental
-   Fixed effects **control for the unit** — country, individual, state, school, etc.

-   As with any control, this means we hold the unit constant and look only at variation **within** the unit

-   **Key point:** We throw away all "between-unit" variation. We only use changes *within* each unit over time.
:::

. . .

::: {.callout-note}
## In English
Fixed effects let you ask: "For the **same country**, in years when the chancellor visited, did trade go up compared to years when they didn't visit?" You're never comparing France to Finland — only France to itself.
:::

# Part 2: Visual Walkthrough

## Pooled Data {.smaller}

```{r}
#| echo: false

library(tidyverse)

df <- tibble(
  Individual = rep(c("Esnaina", "Mariam"), each = 5),
  Exercise   = c(1, 2, 3, 4, 4,   4, 5, 5, 7, 8),
  Colds      = c(4, 3, 5, 3, 2,   7, 8, 7, 5, 6)
)

ggplot(df, aes(x = Exercise, y = Colds, color = Individual, shape = Individual)) +
  geom_point(size = 3) +
  labs(
    title = "Pooled Data",
    x = "Exercise Hours Per Week",
    y = "Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

At first glance, the relationship looks positive.

## Pooled Regression

```{r}
#| echo: false

ggplot(df, aes(x = Exercise, y = Colds, color = Individual, shape = Individual)) +
  geom_point(size = 3) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Pooled Regression",
    x = "Exercise Hours Per Week",
    y = "Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

The pooled regression says more exercise → more colds. That can't be right!

## Adding Individual Means

```{r}
#| echo: false

df_means <- df %>%
  group_by(Individual) %>%
  summarize(
    mean_exercise = mean(Exercise),
    mean_colds    = mean(Colds)
  )

ggplot(df, aes(x = Exercise, y = Colds, color = Individual, shape = Individual)) +
  geom_point(size = 3) +
  geom_vline(
    data = df_means,
    aes(xintercept = mean_exercise, color = Individual),
    linetype = "dashed"
  ) +
  geom_hline(
    data = df_means,
    aes(yintercept = mean_colds, color = Individual),
    linetype = "dashed"
  ) +
  labs(
    title = "Raw Data with Individual Means",
    x     = "Exercise Hours Per Week",
    y     = "Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

Think of each person's mean as their personal origin (0,0).

## Just the Between Variation

```{r}
#| echo: false

ggplot(df_means, aes(x = mean_exercise, y = mean_colds, color = Individual, shape = Individual)) +
  geom_point(size = 4) +
  labs(
    title = "Just Between Variation",
    x = "Mean Exercise Hours Per Week",
    y = "Mean Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

These are the person-specific factors we're going to control away. Mariam exercises more *and* gets more colds — but that's a between-person difference, not a causal effect.

## Within-Person Variation

```{r}
#| echo: false

df_centered <- df %>%
  group_by(Individual) %>%
  mutate(
    Exercise_centered = Exercise - mean(Exercise),
    Colds_centered    = Colds    - mean(Colds)
  )

ggplot(df_centered, aes(x = Exercise_centered, y = Colds_centered)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(size = 3) +
  facet_wrap(~ Individual) +
  labs(
    title = "Within Variation for Each Individual",
    x = "Exercise (Deviation from Individual Mean)",
    y = "Colds (Deviation from Individual Mean)"
  ) +
  theme_minimal(base_size = 14)
```

Now we're asking: "In years when *this person* exercised more than usual, did they get fewer colds than usual?"

## Combining and Regressing

```{r}
#| echo: false

ggplot(df_centered, aes(x = Exercise_centered, y = Colds_centered)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(aes(color = Individual), size = 3) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Centered Data (Both Individuals), Single Regression Line",
    x = "Exercise (Deviation from Mean)",
    y = "Colds (Deviation from Mean)"
  ) +
  theme_minimal(base_size = 14)
```

Now the relationship is **negative** — in years when a person exercises more than their own average, they get fewer colds. The pooled regression had it backwards!

## The Intuition

::: {.callout-important}
## The tradeoff
By using fixed effects, we lose the ability to use between-group variation. If you are interested in between-group comparisons, you cannot use FE.
:::

::: incremental
-   We removed all between-person variation by subtracting each person's mean

-   By controlling for an individual, we also control for all **time-invariant** features of that person

-   We still need to control for things that **vary within the individual over time**
:::

# Part 3: The FE Equation and fixest

## The Fixed Effects Equation

$$Y_{it} = \beta_{i} + \beta_{1}X_{it} + \epsilon_{it}$$

::: {.callout-note}
## In English
$Y_{it}$ is the outcome for unit $i$ at time $t$. The key difference from standard OLS: we now have **unit-specific intercepts** $\beta_i$ — each country (or person, school, etc.) gets its own baseline level. We only estimate the effect of $X$ from variation *within* each unit.
:::

## Varying Intercepts = Dummy Variables

::: {.callout-tip}
It may not be immediately obvious, but letting the intercept vary by individual is exactly the same as adding a dummy (binary) variable for each individual!
:::

. . .

A control for "India" tells us the change we'd expect in Y given that we're in India rather than the reference category. Fixed effects just add a dummy for *every* unit.

. . .

Equivalently, we can "de-mean" the data:

$$Y_{it} - \bar{Y}_{i} = \beta_{1}(X_{it} - \bar{X}_{i}) + \epsilon_{it}$$

This is exactly what our visual walkthrough did — subtract each person's mean.

## Gapminder: Pooled Regression {.smaller}

```{r}
#| echo: true

library(tidyverse)
library(gapminder)
library(ggrepel)

df_gap <- gapminder %>%
  filter(country %in% c("Brazil", "India"))

ggplot(df_gap, aes(x = gdpPercap, y = lifeExp, color = country)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = year), size = 3, min.segment.length = 0) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE,
              linetype = "dashed", color = "black") +
  scale_x_log10(labels = scales::dollar_format(prefix = "$")) +
  labs(
    title = "Life Expectancy vs. GDP per Capita (Pooled Regression)",
    subtitle = "India and Brazil, 1952–2007 (Gapminder)",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)",
    color = "Country"
  ) +
  theme_minimal(base_size = 14)
```

## Gapminder: Country-Specific Intercepts {.smaller}

```{r}
#| echo: true

model_fe <- lm(lifeExp ~ log(gdpPercap) + country, data = df_gap)

predict_df <- expand.grid(
  country   = c("Brazil", "India"),
  gdpPercap = seq(min(df_gap$gdpPercap), max(df_gap$gdpPercap), length.out = 100)
)
predict_df$lifeExp_pred <- predict(model_fe, newdata = predict_df)

ggplot(df_gap, aes(x = gdpPercap, y = lifeExp, color = country)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = year), size = 3, min.segment.length = 0) +
  geom_line(data = predict_df,
            aes(x = gdpPercap, y = lifeExp_pred, color = country),
            linetype = "dashed", size = 1) +
  scale_x_log10(labels = scales::dollar_format(prefix = "$")) +
  labs(
    title = "Life Expectancy vs. GDP per Capita (Fixed Effects)",
    subtitle = "Same slope, different intercepts: Brazil has higher life expectancy at any GDP level",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)",
    color = "Country"
  ) +
  theme_minimal(base_size = 14)
```

## The fixest Package {.smaller}

```{r}
#| echo: true

library(fixest)

# Notation: y ~ x | fixed_effect(s)
m_fe <- feols(lifeExp ~ log(gdpPercap) | country, data = gapminder)
summary(m_fe)
```

. . .

::: {.callout-note}
## How to Interpret This
"When a country's GDP per capita roughly doubles compared to its own average, we expect life expectancy to be about **9.8 years higher than it typically is for that country**." Everything is relative to the unit's own average. That is the fixed effect at work.
:::

# Part 4: Policy Example -- Housing Prices

## The Question

::: incremental
-   Does **building more houses** lead to increases or decreases in home prices?

-   We'd want to control for crime, school quality, amenities...

-   But housing prices and new construction are both driven by unobservable, city-specific factors: desirability, zoning culture, land availability

-   We can control away those city-specific factors with fixed effects (as long as they don't vary over time)
:::

## Simulated Data {.smaller}

We simulated data for 13 US cities where popular cities both build more housing and have higher prices. The true causal effect of new construction on prices is **negative** (-$450 per new house built).

```{r}
#| echo: false

library(tidyverse)
set.seed(30317)

cities <- c("Atlanta", "Chicago", "Houston", "Boston", "New York",
            "Cleveland", "Los Angeles", "San Francisco", "Sacramento",
            "Seattle", "Philadelphia", "Washington DC", "Dallas")

n_obs_per_city <- 100

city_popularity <- runif(length(cities), min = -1, max = 1)
city_intercepts <- 850000 + 650000 * city_popularity
mean_new_houses <- 50 + 20 * city_popularity

beta_crime      <- -300
beta_edu        <-  1500
beta_new_houses <- -450

sim_data <- data.frame()
for (i in seq_along(cities)) {
  crime_vals      <- rnorm(n_obs_per_city, mean = 50, sd = 10)
  edu_vals        <- rnorm(n_obs_per_city, mean = 5, sd = 1)
  new_houses_vals <- rpois(n_obs_per_city, lambda = pmax(mean_new_houses[i], 1))

  price_vals <- city_intercepts[i] +
    beta_crime * crime_vals + beta_edu * edu_vals +
    beta_new_houses * new_houses_vals +
    rnorm(n_obs_per_city, mean = 0, sd = 30000)

  sim_data <- rbind(sim_data, data.frame(
    city = cities[i], crime = crime_vals, edu_quality = edu_vals,
    new_houses = new_houses_vals, price = price_vals
  ))
}
```

## Pooled Data: The Wrong Answer

```{r}
#| echo: false

ggplot(sim_data, aes(x = new_houses, y = price, color = city)) +
  geom_point(size = 2) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Pooled: New Houses vs. Housing Price",
       x = "New Houses Built", y = "Average Housing Price")
```

The pooled regression says more houses → higher prices. But this is confounded by city popularity!

## City-Level Averages: The Between Story

```{r}
#| echo: false

city_means <- sim_data %>%
  group_by(city) %>%
  summarise(avg_price = mean(price), avg_new_houses = mean(new_houses))

ggplot(city_means, aes(x = avg_new_houses, y = avg_price)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Between Cities: More Building = Higher Prices?",
       subtitle = "Each point = one city's average. Popular cities build more AND have higher prices.",
       x = "Avg New Houses", y = "Avg Housing Price")
```

## Within-City: The True Story

```{r}
#| echo: false

ggplot(sim_data, aes(x = new_houses, y = price)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ city, scales = "free") +
  theme_minimal() +
  labs(title = "Within Each City: More Supply → Lower Prices",
       subtitle = "Fixed effects isolate this within-city variation.",
       x = "New Houses Built", y = "Housing Price")
```

## Naive vs. Fixed Effects Model {.smaller}

```{r}
#| echo: true

library(fixest)

# Pooled (naive) model
naive_model <- feols(price ~ new_houses + crime + edu_quality, data = sim_data)

# Fixed effects model
fe_model <- feols(price ~ new_houses + crime + edu_quality | city, data = sim_data)

# Compare side by side
etable(naive_model, fe_model, headers = c("Pooled", "City FE"))
```

## Why the Naive Model Gets It Wrong

::: {.callout-important}
## The Simpson's Paradox of Policy Analysis
**Popular cities** have higher housing prices *and* more new homes. When you pool all cities together, you capture that positive *between-city* correlation. Fixed effects remove it, revealing the **negative within-city** effect: more supply → lower prices. This is exactly the kind of mistake that leads to bad policy recommendations.
:::

# Part 5: Assumptions and Caveats

## What Fixed Effects Require

::: incremental
-   Each regressor must **vary over time** for at least some units, and cannot be perfectly collinear with the fixed effects

-   Fixed effects **cannot solve reverse causality**

    -   A model of police spending will predict that more police *causes* more crime. Why? Because cities hire more police *in response to* crime.

-   Cannot handle time-varying unobserved confounders

-   Time-invariant variables must have **time-invariant effects**. For example, if you use state fixed effects but the effect of education spending on test scores has been growing over time, the fixed effect will not capture that change. See Ren (2025) in the optional readings for more on this.
:::

## A Note on Clustering Standard Errors

::: incremental
-   With panel data, observations within the same unit are not independent. Standard errors need to account for this.

-   The good news: `fixest` clusters standard errors at the fixed effect level by default, so you do not need to do anything extra.
:::

# Part 6: Two-Way Fixed Effects

## Can We Control for Multiple Dimensions? {.smaller}

::: incremental
-   Can we control for both the unit *and* the time period?

-   For example, **state fixed effects** to absorb everything about South Carolina that does not change, plus **year fixed effects** to absorb anything that affected all states in a given year, like a recession

-   This is called **two-way fixed effects** (TWFE), and it was extremely popular in applied research
:::

## The Problem with Two-Way FE

::: incremental
-   It would be very nice to adjust for both unit-specific and time-specific unobserved confounders at once

-   But Imai and Kim (2021) show that you cannot simultaneously and non-parametrically adjust for both unit-specific and time-specific unobserved confounders under this framework

-   See [Kubinec's overview of TWFE issues](https://www.robertkubinec.com/post/fixed_effects/) for a clear walkthrough of why this matters

-   We will return to these points when we cover **Diff-in-Diff**
:::

# A Brief Note on Multilevel Models

## Beyond Fixed Effects: The Shrinkage Idea {.smaller}

::: incremental
-   Fixed effects give each group its own intercept. But what if some groups are very small?

-   Imagine you are rating schools by test scores. A school with 500 students and a 95% pass rate: you trust that number. But a school with 8 students and a 100% pass rate? That is probably just luck.

-   **Multilevel models** (also called hierarchical or random effects models) pull small groups' estimates back toward the overall average. The less data a group has, the more it gets "shrunk" toward the grand mean.

-   For causal questions, fixed effects are the safer choice. But for description, prediction, and ranking, multilevel models can give more honest answers.
:::

## Where You Will See Multilevel Models in Policy {.smaller}

::: incremental
-   **School report cards:** Small schools get extreme pass rates by chance. Multilevel models give more honest rankings by shrinking those estimates toward the average.

-   **Hospital quality ratings:** CMS star ratings use hierarchical models for exactly this reason. A hospital with 10 surgeries and zero deaths is not necessarily the best hospital.

-   **Small-area poverty estimation:** The Census Bureau uses multilevel models to estimate poverty rates for small geographic areas where sample sizes are tiny.

-   You do not need to learn to run these models in this class, but knowing they exist makes you a better consumer of evidence when you encounter rankings and ratings in policy reports.
:::

# Wrapping Up

## What We Covered Today {.smaller}

::: incremental
1.  **Panel and hierarchical data** -- most policy data has this structure
2.  **Fixed effects** -- control for the unit, use only within-unit variation
3.  **The visual intuition** -- pooled regression can get the sign wrong; FE fixes it
4.  **fixest in R** -- the practical tool for FE regression
5.  **Housing prices** -- a full policy simulation showing why pooled estimates mislead
6.  **Assumptions** -- FE cannot fix reverse causality or time-varying confounders
7.  **Two-way FE** -- powerful but with important limitations
:::

## What's Next

::: incremental
-   **Difference-in-Differences** -- comparing changes over time between treated and untreated groups

-   **Regression Discontinuity** -- exploiting arbitrary cutoffs for causal inference
:::

## For Your Final Project

::: {.callout-tip}
## Think About This Now
Does your data have a panel or hierarchical structure? If you are looking at schools, hospitals, states, or any repeated observations, you almost certainly need fixed effects.
:::
