---
title: "Regression Discontinuity Designs"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

# First: One Final DiD Walkthrough

## Did Medicaid Expansion Save Lives? {.smaller}

::: incremental
-   Before we pivot to RDD, one last DiD walkthrough: this time on a live policy question
-   **The paper:** Miller, Johnson, and Wherry (2021), "Medicaid and Mortality: New Evidence from Linked Survey and Administrative Data," *Quarterly Journal of Economics*
-   **The question:** Did the ACA's Medicaid expansion reduce mortality among near Medicare-eligible adults?
-   **Why it matters:** One of the most consequential policy changes in decades + a case where credible DiD evidence has shaped the policy debate
:::

## The DiD Setup {.smaller}

::: incremental
-   The ACA let states expand Medicaid starting **January 2014**
    -   About 24 states expanded in 2014
    -   Some expanded later (Louisiana 2016, Virginia 2019)
    -   A dozen still have never expanded
-   **Treatment group:** Near Medicare-eligible adults (ages 55-64) in expansion states
-   **Control group:** Near Medicare-eligible adults in non-expansion states
-   **Before:** 2010-2013 (pre-expansion)
-   **After:** 2014 onward
-   Same 2×2 structure as the California organ donation example, just with states and years instead of hospitals and quarters
:::

## What the Data Look Like {.smaller}

```{r}
#| echo: true
library(tidyverse)
set.seed(30317)

n_exp  <- 20   # expansion states
n_ctrl <- 20   # non-expansion states
years  <- 2010:2017

# State-specific intercepts — centered within each group so the two
# groups have the same average baseline (keeps pre-trends visually parallel)
fe_exp   <- rnorm(n_exp,  0, 40); fe_exp   <- fe_exp   - mean(fe_exp)
fe_ctrl  <- rnorm(n_ctrl, 0, 40); fe_ctrl  <- fe_ctrl  - mean(fe_ctrl)
state_fe <- c(fe_exp, fe_ctrl)

df_med <- expand_grid(
  state_id = 1:(n_exp + n_ctrl),
  year     = years
) %>%
  mutate(
    expansion = state_id <= n_exp,
    post      = year >= 2014,
    trend     = 1400 - 12 * (year - 2010),     # overall downward trend
    fe        = state_fe[state_id],
    tx        = ifelse(expansion & post, -130, 0),  # ~130 per 100k reduction
    mortality = fe + trend + tx + rnorm(n(), 0, 25)
  )

# Collapse to group-year averages for plotting
df_plot <- df_med %>%
  group_by(year, expansion) %>%
  summarize(mortality = mean(mortality), .groups = "drop") %>%
  mutate(group = ifelse(expansion, "Expansion States", "Non-Expansion States"))

ggplot(df_plot, aes(x = year, y = mortality,
                    color = group, linetype = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "gray40") +
  annotate("text", x = 2014.15, y = 1420, hjust = 0, size = 3.5,
           color = "gray30", label = "ACA expansion begins") +
  scale_color_manual(values = c("Expansion States"     = "#e31a1c",
                                "Non-Expansion States" = "#1f78b4")) +
  scale_x_continuous(breaks = 2010:2017) +
  labs(title = "Mortality for Adults Ages 55-64, by Expansion Status",
       subtitle = "Restricted to Medicaid-eligible adults (income below 138% FPL)",
       x = "Year", y = "Mortality (per 100,000)",
       color = "", linetype = "") +
  theme_minimal(base_size = 14)
```

Parallel before 2014, divergence after — the visual signature of a credible DiD.

## What Makes This Credible? {.smaller}

::: incremental
-   **Parallel pre-trends:** The two groups track each other closely before 2014, necessary (but not sufficient) evidence the comparison is valid
-   **No anticipation:** The 2012 Supreme Court ruling made expansion an unexpected state-by-state choice. Hard to plan around in advance
-   **Staggered adoption:** Later expanders (Louisiana 2016, Virginia 2019) complicate TWFE. The paper uses modern estimators (Callaway-Sant'Anna) to avoid the forbidden comparisons we discussed
-   **LATE interpretation:** The estimate applies to Medicaid-eligible near-Medicare-eligible adults in expansion states: not the whole population
-   **Bottom line:** The paper estimates Medicaid expansion reduced annual mortality by \~9% in this group
:::

. . .

::: callout-tip
## The Bridge to RDD

DiD works when you can find **two groups** — one treated, one not — and argue they would have trended in parallel. But what if there's no clean comparison group? What if the only thing separating treated from untreated is a *rule*, and either side of that rule looks nearly identical? That's where **Regression Discontinuity** comes in.
:::

# Introducing RDD

## The Setup

::: incremental
-   Sometimes, the way a policy is implemented creates a **sharp cutoff** between who gets treated and who doesn't

    -   You qualify for Medicaid if your income is below 138% of the poverty line, but not at 139%

    -   You get Title I school funding if your poverty rate is above a threshold, but not one percentage point below

-   If being just above or just below the cutoff is **as good as random**, we can estimate the treatment effect by comparing the two sides
:::

## Why Should MPAs Care? {.smaller}

::: incremental
-   Government programs are **full of arbitrary cutoffs**: income thresholds, age limits, test score requirements, geographic boundaries

-   RDD lets you turn these administrative rules into **natural experiments**

-   As a policy analyst, recognizing these cutoffs, hiding in eligibility rules everywhere, unlocks powerful research designs

-   RDD is widely considered one of the **most credible** quasi-experimental designs when applicable
:::

## The Core Intuition {.smaller}

::: incremental
-   Imagine a scholarship for students scoring above 80 on an exam

-   A student who scored 81 got the scholarship. A student who scored 79 didn't.

-   Are these two students fundamentally different? Probably not - the difference is essentially **luck**

-   By comparing outcomes for students *just above* and *just below* 80, we approximate a **local randomized experiment**
:::

. . .

::: callout-note
## In English

RDD exploits the fact that people right at the cutoff are essentially identical - the only thing that differs is whether they got the treatment. That's almost as good as random assignment.
:::

## Today's Roadmap {.smaller}

::: incremental
1.  **Terminology & assumptions**: running variables, cutoffs, and what can go wrong
2.  **The graphical recipe**: seeing the discontinuity step by step
3.  **Bandwidth & estimation**: how to estimate the jump and choose a window
4.  **Policy examples**: shingles vaccines and Medicare at 65
5.  **Sharp vs. fuzzy RDD**: what if the cutoff isn't perfectly enforced? (Maimonides' Rule)
6.  **Validity checks**: manipulation tests, placebo cutoffs, donut hole tests
7.  **Practical advice** + broader lessons about quant work as an MPA
:::

# Part 1: Terminology & Assumptions

## Key Terminology {.smaller}

Let's ground the core terms in one real policy: **Pell Grants** - federal need-based aid for college students. Eligibility depends on **Expected Family Contribution (EFC)**. If your EFC is below a threshold, you qualify.

::: incremental
-   **Running Variable (aka forcing variable):** The variable that determines treatment
    -   *Pell example:* Expected Family Contribution (EFC)
    -   *In general:* test scores, income, age, birth date, distance from a border
-   **Cutoff (aka threshold):** The value where treatment "turns on"
    -   *Pell example:* the max EFC for eligibility (historically around \$5,800)
    -   *In general:* any administrative line — 138% FPL, score of 80, age 65
-   **Bandwidth:** How close to the cutoff do we look?
    -   *Pell example:* students within \$500 of the cutoff? \$1,000? \$2,000?
    -   Narrower = more credible (more similar students), less precise (fewer observations)
    -   Wider = less credible, more precise
:::

## The First Picture: Seeing the Jump {.smaller}

```{r}
#| echo: true

library(tidyverse)
set.seed(30317)

n      <- 500
cutoff <- 0.5
x      <- runif(n, 0, 1)
y      <- 2 + 5 * x + 2 * (x >= cutoff) + rnorm(n, 0, 0.3)
df     <- tibble(x, y)

# Fit a line on each side and predict the value right at the cutoff —
# the gap between the two is the treatment effect we'll draw.
left_fit  <- lm(y ~ x, data = filter(df, x <  cutoff))
right_fit <- lm(y ~ x, data = filter(df, x >= cutoff))
y_left    <- predict(left_fit,  newdata = tibble(x = cutoff))
y_right   <- predict(right_fit, newdata = tibble(x = cutoff))

ggplot(df, aes(x = x, y = y)) +
  geom_point(alpha = 0.35, size = 1.5, color = "gray40") +
  geom_smooth(data = filter(df, x <  cutoff), method = "lm", se = FALSE,
              color = "black", linewidth = 1.2) +
  geom_smooth(data = filter(df, x >= cutoff), method = "lm", se = FALSE,
              color = "black", linewidth = 1.2) +
  geom_vline(xintercept = cutoff, linetype = "dashed", linewidth = 0.8) +
  annotate("segment", x = cutoff, xend = cutoff,
           y = y_left, yend = y_right,
           color = "#6a3d9a", linewidth = 2) +
  annotate("point", x = cutoff, y = y_left,  color = "#6a3d9a", size = 3) +
  annotate("point", x = cutoff, y = y_right, color = "#6a3d9a", size = 3) +
  annotate("label", x = cutoff + 0.04, y = (y_left + y_right) / 2,
           label = "The jump\n= treatment effect",
           color = "#6a3d9a", hjust = 0, size = 4, label.size = 0) +
  labs(title = "Regression Discontinuity: Graphical Intuition",
       x = "Running Variable (X)", y = "Outcome (Y)") +
  theme_minimal(base_size = 14)
```

The **jump** at the cutoff is the treatment effect. Everything else is smooth.

## Key Assumptions {.smaller}

::: incremental
-   **No manipulation:** Subjects cannot game the running variable to choose their treatment status

    -   If students knew the scholarship cutoff was 80, could they artificially score just above it?

    -   Age-based cutoffs (like Medicare at 65) are harder to manipulate.

-   **Continuity:** Everything *other* than treatment changes smoothly through the cutoff

    -   No other policies kick in at the same threshold

    -   *Violation example:* Studying Medicaid's effect at 138% FPL gets complicated because SNAP, CHIP, and housing assistance all have similar thresholds.

-   **No anticipation:** People don't change behavior in advance of crossing the cutoff
:::

. . .

::: callout-important
## The Big Question

Ask yourself: "Is there any reason someone *just below* the cutoff would be systematically different from someone *just above* — other than the treatment?" If no, RDD is credible.
:::

## What Does RDD Estimate? {.smaller}

::: incremental
-   RDD gives us a **Local Average Treatment Effect (LATE)**

-   It tells us the effect of treatment **for people right at the cutoff**, not for everyone

    -   The effect of Medicare for 65-year-olds, not 40-year-olds

    -   The effect of the scholarship for students scoring around 80, not students scoring 95

-   This is both a **strength** (very credible for that group) and a **limitation** (may not generalize)
:::

. . .

::: callout-note
## In English

RDD gives you a very clean answer to a very specific question: "What is the effect of treatment for people right at the boundary?" Whether that answer applies more broadly is a separate question.
:::

# Part 2: The Graphical Recipe

We'll illustrate the four-step recipe with a **merit-based college scholarship**: students scoring **above 80** on a standardized test receive a scholarship; we want to know if it raises college completion rates.

## Step 1: Plot the Raw Data {.smaller}

```{r}
#| echo: true

set.seed(30317)
n      <- 500
cutoff <- 80
score  <- runif(n, 60, 100)
completion <- 0.30 + 0.005 * (score - 60) +
              0.10  * (score >= cutoff) + rnorm(n, 0, 0.04)

df <- tibble(score, completion,
             group = factor(ifelse(score < cutoff, "No Scholarship", "Scholarship"),
                            levels = c("No Scholarship", "Scholarship")))

ggplot(df, aes(x = score, y = completion, color = group)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_vline(xintercept = cutoff, linetype = "dashed",
             color = "gray40", linewidth = 1) +
  scale_color_manual(values = c("No Scholarship" = "#1f78b4",
                                "Scholarship"    = "#e31a1c")) +
  labs(title = "(a) Raw Data: Merit Scholarship at Score 80",
       x = "Test Score", y = "College Completion Rate", color = "") +
  theme_minimal(base_size = 14)
```

## Step 2: Fit a Line on Each Side {.smaller}

```{r}
#| echo: true

ggplot(df, aes(x = score, y = completion, color = group)) +
  geom_point(alpha = 0.5) +
  geom_smooth(data = filter(df, group == "No Scholarship"),
              method = "lm", se = FALSE, linewidth = 1, color = "black") +
  geom_smooth(data = filter(df, group == "Scholarship"),
              method = "lm", se = FALSE, linewidth = 1, color = "black") +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("No Scholarship" = "#1f78b4",
                                "Scholarship"    = "#e31a1c")) +
  labs(title = "(b) Fit a Line on Each Side",
       x = "Test Score", y = "College Completion Rate", color = "") +
  theme_minimal(base_size = 14)
```

We fit a simple straight line on each side. You can also use LOESS or polynomials — but linear is honest, easy to interpret, and the modern `rdrobust` default.

## Step 3: Choose a Bandwidth {.smaller}

```{r}
#| echo: true

bw  <- 4   # 4-point window around the cutoff
df2 <- df %>% mutate(in_bw = abs(score - cutoff) <= bw)

ggplot() +
  geom_point(data = filter(df2, !in_bw),
             aes(score, completion), color = "gray80") +
  geom_point(data = filter(df2, in_bw),
             aes(score, completion, color = group), size = 2) +
  geom_smooth(data = filter(df2, in_bw, score <  cutoff),
              aes(score, completion), method = "lm", se = FALSE,
              color = "black", linewidth = 1) +
  geom_smooth(data = filter(df2, in_bw, score >= cutoff),
              aes(score, completion), method = "lm", se = FALSE,
              color = "black", linewidth = 1) +
  geom_vline(xintercept = c(cutoff - bw, cutoff + bw),
             linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = cutoff, linetype = "solid", color = "gray40") +
  scale_color_manual(values = c("No Scholarship" = "#1f78b4",
                                "Scholarship"    = "#e31a1c")) +
  labs(title = "(c) Pick a Bandwidth: Look Only Near the Cutoff",
       x = "Test Score", y = "College Completion Rate", color = NULL) +
  theme_minimal(base_size = 14)
```

We restrict to students scoring within 4 points of the cutoff — these students are most comparable.

## Step 4: Estimate the Jump {.smaller}

```{r}
#| echo: true

df3 <- df %>% filter(abs(score - cutoff) <= bw)

# Fit a line on each side within the bandwidth, then predict at the cutoff
left_fit  <- lm(completion ~ score, data = filter(df3, score <  cutoff))
right_fit <- lm(completion ~ score, data = filter(df3, score >= cutoff))
y_left    <- predict(left_fit,  newdata = tibble(score = cutoff))
y_right   <- predict(right_fit, newdata = tibble(score = cutoff))
jump      <- y_right - y_left

ggplot() +
  geom_point(data = df,  aes(score, completion), color = "gray90") +
  geom_point(data = df3, aes(score, completion, color = group), size = 2) +
  geom_smooth(data = filter(df3, score <  cutoff),
              aes(score, completion), method = "lm", se = FALSE,
              color = "black", linewidth = 1) +
  geom_smooth(data = filter(df3, score >= cutoff),
              aes(score, completion), method = "lm", se = FALSE,
              color = "black", linewidth = 1) +
  geom_segment(aes(x = cutoff, xend = cutoff,
                   y = y_left, yend = y_right),
               color = "#6a3d9a", linewidth = 2) +
  annotate("point", x = cutoff, y = y_left,  color = "#6a3d9a", size = 3) +
  annotate("point", x = cutoff, y = y_right, color = "#6a3d9a", size = 3) +
  annotate("label", x = cutoff + 0.5, y = (y_left + y_right) / 2,
           label = sprintf("Estimated effect:\n+%.1f pp", jump * 100),
           color = "#6a3d9a", hjust = 0, size = 4, label.size = 0) +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("No Scholarship" = "#1f78b4",
                                "Scholarship"    = "#e31a1c")) +
  labs(title = "(d) Estimate the Jump at the Cutoff",
       x = "Test Score", y = "College Completion Rate", color = "") +
  theme_minimal(base_size = 14)
```

The purple segment is the **estimated treatment effect** — the gap between each line's predicted value *at the cutoff*.

# Part 3: Bandwidth & Estimation

## The Bandwidth Tradeoff {.smaller}

```{r}
#| echo: true

bw_values <- c(1, 2, 4, 10)    # bandwidths in test-score points

df_facet <- df %>%
  crossing(bw = bw_values) %>%
  mutate(in_bw = abs(score - cutoff) <= bw)

ggplot() +
  geom_point(data = filter(df_facet, !in_bw),
             aes(score, completion), color = "gray80") +
  geom_point(data = filter(df_facet, in_bw),
             aes(score, completion, color = group), size = 1.5) +
  geom_smooth(data = filter(df_facet, in_bw & score <  cutoff),
              aes(score, completion), method = "lm", se = FALSE,
              color = "black", linewidth = 1) +
  geom_smooth(data = filter(df_facet, in_bw & score >= cutoff),
              aes(score, completion), method = "lm", se = FALSE,
              color = "black", linewidth = 1) +
  geom_vline(data = tibble(bw = bw_values),
             aes(xintercept = cutoff - bw), linetype = "dashed", color = "gray50") +
  geom_vline(data = tibble(bw = bw_values),
             aes(xintercept = cutoff + bw), linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = cutoff, linetype = "solid", color = "gray40") +
  scale_color_manual(values = c("No Scholarship" = "#1f78b4",
                                "Scholarship"    = "#e31a1c")) +
  labs(title = "Bandwidth Choice (in Test-Score Points)",
       x = "Test Score", y = "College Completion Rate", color = NULL) +
  facet_wrap(~ bw, ncol = 2, labeller = label_both) +
  theme_minimal(base_size = 14)
```

Smaller bandwidth → more credible, less power. `rdrobust` can choose the optimal bandwidth for you.

## The Simplest RDD Regression {.smaller}

We can estimate the jump with OLS. Let's walk through each piece:

$$Y = \beta_{0} + \beta_{1}(\text{Running} - c) + \beta_{2}\text{Treated} + \beta_{3}(\text{Running} - c) \times \text{Treated} + \epsilon$$

::: incremental
-   $\text{Running} - c$ is the running variable, **centered at the cutoff**. So 0 = "right at the cutoff," negative = below, positive = above.
-   $\text{Treated}$ is a 0/1 indicator: 1 if above the cutoff, 0 if below
-   $\beta_0$ = outcome *at the cutoff* on the just-below side
-   $\beta_1$ = slope on the left side
-   $\beta_2$ = **the jump at the cutoff → this is the treatment effect**
-   $\beta_3$ × (Running − c) × Treated lets the two sides have **different slopes**. Without it, we'd force the left and right lines to tilt in exactly the same way, which is too restrictive
:::

. . .

::: callout-note
## In English

This one regression fits **two lines**, one on each side of the cutoff, in a single model. $\beta_2$ reads off the vertical gap between those lines *right at the cutoff*. That gap is the treatment effect.
:::

## Seeing the Two Lines {.smaller}

```{r}
#| echo: true

# Apply the formula above to the scholarship data: center the running
# variable at the cutoff, then fit the interaction regression.
df_scholar <- df %>%
  mutate(score_c = score - cutoff,
         side    = factor(score >= cutoff,
                          levels = c(FALSE, TRUE),
                          labels = c("Below cutoff", "At/above cutoff")),
         treated = as.integer(score >= cutoff))

rdd_model <- lm(completion ~ score_c * treated, data = df_scholar)
round(coef(rdd_model), 3)

ggplot(df_scholar, aes(x = score_c, y = completion, color = side)) +
  geom_point(alpha = 0.35, size = 1.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Below cutoff"     = "#1f78b4",
                                "At/above cutoff"  = "#e31a1c")) +
  labs(title = "Linear RDD on the Scholarship Data",
       x = "Test Score (centered at 80)",
       y = "College Completion Rate", color = "Side of Cutoff") +
  theme_minimal(base_size = 14)
```

$\beta_0$ = intercept (completion just below cutoff). $\beta_1$ = slope on the left. $\beta_2$ = **the jump** — our treatment effect. $\beta_3$ = how the slope *changes* when we cross the cutoff.

## Using `rdrobust`: The Modern Approach {.smaller}

```{r}
#| echo: true
#| eval: false

# install.packages("rdrobust")
library(rdrobust)

# rdrobust does everything for you:
# - Chooses the optimal bandwidth
# - Fits local polynomial regressions
# - Provides bias-corrected confidence intervals
rd_result <- rdrobust(y = df$completion, x = df$score, c = 80)
summary(rd_result)
```

. . .

::: callout-tip
## What `rdrobust` Handles

Instead of manually choosing bandwidths and functional forms, `rdrobust` uses data-driven methods (Calonico, Cattaneo, and Titiunik, 2014) to select the optimal bandwidth and provide robust inference. This is now the **standard tool** for applied RDD work.
:::

## How `rdrobust` Chooses the Bandwidth {.smaller}

The tool picks a bandwidth by balancing **two concerns that pull in opposite directions**:

::: incremental
-   **Too wide a window** → you include observations far from the cutoff, which may not be comparable. Your estimate gets *biased*
-   **Too narrow a window** → you don't have enough observations. Your estimate gets *noisy*
:::

. . .

The "best" bandwidth minimizes a single quantity called **mean squared error (MSE)** — which combines bias and noise into one number. Think of it as finding the right **zoom level**: close enough to the cutoff that comparisons are credible, far enough out that you have enough data to see the jump.

. . .

::: callout-tip
## You're Not Locked In

The default bandwidth is a sensible *starting point*, not a final answer. Always run the **sensitivity check** (Part 6): if the estimate is stable across a range of bandwidths, the result is robust. If it swings wildly, the data is telling you something.
:::

## `rdrobust` in Action {.smaller}

```{r}
#| echo: true

# Let's use our simulated data
library(rdrobust)

rd_fit <- rdrobust(y = df$completion, x = df$score, c = cutoff)
summary(rd_fit)
```

## Reading the `rdrobust` Output {.smaller}

Focus on these five numbers from the summary above:

::: incremental
-   **BW est. (h):** The data-driven **optimal bandwidth** (in test-score points). `rdrobust` picks it for you based on MSE
-   **Eff. Number of Obs.:** How many observations actually contributed to the estimate on each side. Narrow bandwidths shrink the effective sample
-   **Coef. (Conventional):** The **point estimate** of the jump at the cutoff: this is your treatment effect
-   **\[ 95% C.I. \] (Robust):** The bias-corrected confidence interval
-   **P\>\|z\|:** The p-value — is the jump statistically distinguishable from zero?
:::

## The `rdplot` Visualization {.smaller}

```{r}
#| echo: true

# rdrobust comes with a companion plotting function
rdplot(y = df$completion, x = df$score, c = cutoff,
       title = "RD Plot with Automatic Binning",
       x.label = "Test Score",
       y.label = "College Completion Rate")
```

`rdplot` bins the data and shows the local polynomial fit. This is basically a publication ready figure!

## Should You Add Controls? {.smaller}

::: incremental
-   The short answer: **probably not for identification, maybe for precision**

-   The logic: if the cutoff creates as-good-as-random assignment, there's no omitted variable bias to fix

-   But adding covariates that predict the outcome can **reduce noise** and tighten confidence intervals

-   **Which covariates?** Use variables measured *before* the cutoff that predict the outcome

    -   *Safe and helpful:* family income, demographics, prior GPA, baseline test scores, lagged outcomes
    -   *Dangerous:* anything measured *after* treatment. it may itself be affected by the cutoff, which biases your estimate
:::

. . .

::: callout-note
## In English

Controls in RDD aren't about fixing confounding (the cutoff handles that). They're about statistical power: predicting more of the outcome variance so the treatment effect estimate is more precise.
:::

## Kernels: Weighting by Distance {.smaller}

An alternative to sharp bandwidth cutoffs: **weight observations** by how close they are to the cutoff.

```{r}
#| echo: true

h <- 10   # bandwidth = 10 test-score points
x_kern <- seq(-h, h, length.out = 500)
w <- pmax(1 - abs(x_kern) / h, 0)

ggplot(tibble(x = x_kern, weight = w), aes(x = x, y = weight)) +
  geom_area(fill = "grey80") +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = c(-h, h), linetype = "dashed") +
  labs(title = "Triangular Kernel Weight Function",
       subtitle = "Bandwidth h = 10 test-score points around the cutoff",
       x = "Distance from Cutoff (test-score points)", y = "Weight") +
  theme_minimal(base_size = 14)
```

Observations closer to the cutoff get **more weight**. This is what `rdrobust` does under the hood.

# Part 4: Policy Examples

## Example 1: Does a Shingles Vaccine Prevent Dementia? {.smaller}

**The policy.** In September 2013, the Welsh NHS introduced a shingles vaccination program for older adults. The eligibility rule was a **sharp birth-date cutoff**: adults born on or after **September 2, 1933** were eligible; adults born before that date were deemed too old per the guidance.

**Why this matters for dementia research:**

::: incremental
-   Shingles is caused by reactivation of the varicella-zoster virus — the same virus that causes chickenpox in childhood
-   Emerging hypothesis: chronic viral infections may contribute to dementia risk, possibly by reducing inflammation of the brain.
-   Hard to test causally: people who get vaccinated differ from those who don't (wealthier, more health-conscious, better access to care)
-   The Welsh birth-date cutoff removes that confounding entirely. **You cannot manipulate your birth date.**
:::

## **The RDD setup:**

-   **Running variable:** Week of birth relative to Sept 2, 1933
-   **Cutoff:** Week 0 (Sept 2, 1933)
-   **Treatment:** Eligibility for the shingles vaccine
-   **Main outcome:** New dementia diagnosis within 7 years

*Source: Eyting et al. (2025), Nature*

## First Stage: Did the Cutoff Change Vaccination? {.smaller}

```{r}
#| echo: false

library(ggplot2)
library(dplyr)
library(patchwork)

## Simulated data inspired by Eyting et al. 2025 for pedagogical purposes
set.seed(30317)

weeks <- seq(-156, 156, by = 4)

panel_a <- tibble(
  week = weeks,
  y = ifelse(week < 0, 30 + rnorm(length(week), 0, 2),
                        77 + rnorm(length(week), 0, 2)),
  outcome = "Ever had shingles vaccine (%)"
)

plot_panel <- function(data, panel_label, coef_text) {
  data <- mutate(data,
                 side = factor(ifelse(week < 0, "Born before cutoff",
                                      "Born after cutoff"),
                               levels = c("Born before cutoff",
                                          "Born after cutoff")))
  ggplot(data, aes(x = week, y = y)) +
    geom_point(aes(color = side), alpha = 0.85, size = 1.5) +
    geom_smooth(data = filter(data, week <  0), method = "lm",
                se = FALSE, color = "black") +
    geom_smooth(data = filter(data, week >= 0), method = "lm",
                se = FALSE, color = "black") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    scale_color_manual(values = c("Born before cutoff" = "#1f78b4",
                                  "Born after cutoff"  = "#e31a1c")) +
    labs(title = paste0("Panel ", panel_label),
         x = "Week of birth (relative to Sep 2 1933)",
         y = unique(data$outcome)) +
    annotate("text", x = -120, y = max(data$y) - 2,
             label = coef_text, hjust = 0, size = 3.2) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
}

plot_panel(panel_a, "a", "Coef. 47.2\n95% CI = 46.1 to 48.7")
```

**What you're seeing:** each dot is a birth-week cohort. X-axis is week of birth relative to the Sept 2, 1933 cutoff. Y-axis is the percent who eventually received the shingles vaccine.

**The jump at week 0 is \~47 percentage points** — from about 30% to about 77%. A 4-week age difference at the cutoff produced an enormous difference in treatment. This is a very strong first stage.

## Placebo Checks: The Logic {.smaller}

**The worry:** what if people born just before vs. just after the cutoff differ in some *other* way — not just eligibility? Then we'd confound the vaccine effect with something else.

**The check:** look for jumps in *pre-treatment* outcomes — things measured BEFORE the 2013 policy started. If the groups were comparable, pre-treatment outcomes should trend smoothly through the cutoff with no discontinuity.

::: callout-note
## Two Natural Placebos for This Study

-   **Prior shingles diagnoses.** If the groups were comparable, shingles rates measured *before* the 2013 program should be similar on either side of the cutoff
-   **Prior dementia diagnoses.** Dementia tends to rise with age, but the rate measured *before* the program should vary *smoothly* across birth week — no sudden jump at Sept 2, 1933
:::

## Placebo Checks: Prior Shingles and Prior Dementia {.smaller}

```{r}
#| echo: false

set.seed(30317)

panel_b <- tibble(
  week = weeks,
  y = 10 + rnorm(length(weeks), 0, 1.5),
  outcome = "Had shingles BEFORE program start (%)"
)

panel_c <- tibble(
  week = weeks,
  y = 15 - 0.005 * weeks + rnorm(length(weeks), 0, 0.1),
  outcome = "Had dementia within 7 yrs BEFORE program (%)"
)

plot_panel(panel_b, "b", "Coef. 0.1\n95% CI = -0.5 to 0.6") +
plot_panel(panel_c, "c", "Coef. -0.2\n95% CI = -0.8 to 0.4")
```

::: incremental
-   **Panel (b):** Prior shingles diagnoses are flat across the cutoff (coef ≈ 0.1). Groups had similar shingles histories
-   **Panel (c):** Prior dementia shows a smooth downward trend (older people → more dementia), but **no jump** at the cutoff. Groups were comparable
-   **The takeaway:** the cutoff changed vaccination rates (panel a) and *nothing else* (panels b, c). Identification comes from the policy, not from underlying differences
:::

## The Punchline: Vaccines Reduce Dementia {.smaller}

```{r}
#| echo: false

set.seed(30317)
weeks <- seq(-156, 156, by = 8)
n <- length(weeks)
y <- ifelse(weeks < 0,
            18 - 0.025 * weeks + rnorm(n, 0, 0.4),
            18 - 0.025 * weeks - 3 + rnorm(n, 0, 0.4))

df_dem <- tibble(week = weeks, y = y,
                 side = factor(ifelse(weeks < 0, "Born before cutoff",
                                      "Born after cutoff"),
                               levels = c("Born before cutoff",
                                          "Born after cutoff")))

ggplot(df_dem, aes(x = week, y = y)) +
  geom_point(aes(color = side), size = 2) +
  geom_smooth(data = filter(df_dem, week <  0),
              method = "lm", se = FALSE, color = "black") +
  geom_smooth(data = filter(df_dem, week >= 0),
              method = "lm", se = FALSE, color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("Born before cutoff" = "#1f78b4",
                                "Born after cutoff"  = "#e31a1c")) +
  annotate("text", x = 100, y = 20.5, hjust = 0, size = 3.5, color = "gray30",
           label = "Effect of eligibility: -2.7 pp\n95% CI = -4 to -1.5\nP = 0.022") +
  labs(title = "Effect of Shingles Vaccine Eligibility on Dementia",
       x = "Week of birth (relative to Sep 2 1933)",
       y = "New dementia diagnosis over 7 years (%)",
       color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

When the jump is this clear, you almost don't need statistics. That's the beauty of a good RDD.

## Example 2: Medicare at Age 65 {.smaller}

Card, Dobkin, and Maestas (2008) used the 65th-birthday cutoff to study the effect of near-universal health insurance.

```{r}
#| echo: false

library(patchwork)
set.seed(30317)

age_y <- seq(-5, 5, by = 0.25)   # age in years relative to 65

# Panel A: insurance coverage rate (%)
cov <- ifelse(age_y < 0, 82 + 0.8 * age_y, 97 + 0.05 * age_y) +
       rnorm(length(age_y), 0, 1)

# Panel B: doctor visits per year
vis <- ifelse(age_y < 0, 4.0 + 0.04 * age_y, 5.2 + 0.04 * age_y) +
       rnorm(length(age_y), 0, 0.15)

df_med <- tibble(age_y, cov, vis,
                 side = factor(ifelse(age_y < 0, "Under 65", "65 and over"),
                               levels = c("Under 65", "65 and over")))

plot_med <- function(yvar, ylab, title) {
  ggplot(df_med, aes(x = age_y, y = .data[[yvar]])) +
    geom_point(aes(color = side), size = 1.8, alpha = 0.85) +
    geom_smooth(data = filter(df_med, age_y <  0),
                method = "lm", se = FALSE, color = "black") +
    geom_smooth(data = filter(df_med, age_y >= 0),
                method = "lm", se = FALSE, color = "black") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
    scale_color_manual(values = c("Under 65"    = "#1f78b4",
                                  "65 and over" = "#e31a1c")) +
    labs(title = title,
         x = "Age (years, relative to 65th birthday)",
         y = ylab) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none")
}

plot_med("cov", "Insurance coverage (%)",
         "(a) First stage: Coverage jumps at 65") +
plot_med("vis", "Doctor visits per year",
         "(b) Outcome: Healthcare use jumps")
```

## Reading the Medicare Panels {.smaller}

::: incremental
-   **Panel (a) — First stage:** Coverage jumps from \~82% to \~97% at the 65th birthday. Eligibility is sharp, and take-up is nearly universal
-   **Panel (b) — Outcome:** Doctor visits per year rise from \~4 to \~5.2 at 65 — newly-insured people use more healthcare
-   **Why this is credible:** you cannot manipulate your age, and nothing else changes abruptly at exactly 65. The cutoff is truly arbitrary
-   **LATE interpretation:** effect applies to people *right at* 65. We cannot learn the effect of Medicare on 40-year-olds from this design
:::

. . .

::: callout-tip
## Bridge Back to the Opener

We started today with **Medicaid expansion**, a DiD design exploiting *state-level* variation in policy. Now we're looking at **Medicare at 65**, an RDD design exploiting *age-level* variation in eligibility. Same substantive question ("does insurance improve health?"), two different identification strategies. Having both in your toolkit is great.
:::

## More RDD Examples in Policy {.smaller}

::: incremental
-   **Voting age and political participation:** Turning 18 creates a sharp cutoff for eligibility: used to study whether early voting experience shapes lifelong habits

-   **Drinking age and mortality:** The legal drinking age at 21 creates a sharp jump in alcohol consumption and, tragically, in traffic deaths (Carpenter & Dobkin, 2009)

-   **Title I funding:** Schools above a poverty threshold receive federal funding: used to study whether additional resources improve outcomes

-   **Close elections:** When one candidate barely wins vs. barely loses, outcomes are as-if randomly assigned: a popular design in political science (Lee, 2008)
:::

. . .

::: callout-tip
## The Pattern

Notice the pattern across these examples: an **administrative rule** creates a cutoff → people just above and below are essentially the same → comparing their outcomes reveals the causal effect. Look for these rules in your policy area!
:::

# Part 5: Sharp vs. Fuzzy RDD

## Sharp RDD: Perfect Compliance {.smaller}

```{r}
#| echo: true

library(scales)

# Local cutoff just for these illustrative diagrams
c_demo <- 0.5
x_grid <- seq(0, 1, length.out = 500)

df_sharp <- tibble(x = x_grid, p = ifelse(x_grid < c_demo, 0, 1))

ggplot(df_sharp, aes(x = x, y = p)) +
  geom_line(linewidth = 1.5, color = "black") +
  geom_vline(xintercept = c_demo, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = percent_format(accuracy = 1L),
                     breaks = seq(0, 1, by = 0.25)) +
  labs(title = "(a) Sharp Design: Treatment Probability",
       x = "Running Variable", y = "Proportion Treated") +
  theme_minimal(base_size = 14)
```

In a **sharp** design, the cutoff perfectly determines treatment: 0% below, 100% above.

## Fuzzy RDD: Imperfect Compliance {.smaller}

```{r}
#| echo: true

df_fuzzy <- tibble(
  x = x_grid,
  p = ifelse(x_grid < c_demo, 0.15 + 0.2 * x_grid, 0.55 + 0.2 * x_grid)
)

ggplot(df_fuzzy, aes(x = x, y = p)) +
  geom_line(linewidth = 1.5, color = "black") +
  geom_vline(xintercept = c_demo, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = percent_format(accuracy = 1L),
                     breaks = seq(0, 1, by = 0.25)) +
  labs(title = "(b) Fuzzy Design: Treatment Probability",
       x = "Running Variable", y = "Proportion Treated") +
  theme_minimal(base_size = 14)
```

In a **fuzzy** design, the cutoff causes a **jump** in the probability of treatment — but it's not 0-to-100%.

## Fuzzy RDD in the Wild: Maimonides' Rule {.smaller}

Angrist and Lavy (1999): Israeli schools follow a rule attributed to Maimonides — **no more than 40 students per class**. A school with 41 students *should* split into two classes of \~20.

```{r}
#| echo: false

set.seed(30317)

n_schools <- 500
enroll    <- sample(20:140, n_schools, replace = TRUE)
predicted <- enroll / ceiling(enroll / 40)

# Fuzzy compliance: 30% of schools have larger classes than the rule dictates
noncompliant <- rbinom(n_schools, 1, 0.30)
upward_dev   <- noncompliant * runif(n_schools, 4, 12)
actual       <- predicted + upward_dev + rnorm(n_schools, 0, 1.2)

df_maim   <- tibble(enroll, actual)
rule_grid <- tibble(enroll = 20:140,
                    predicted = (20:140) / ceiling((20:140) / 40))

ggplot() +
  geom_point(data = df_maim, aes(x = enroll, y = actual),
             alpha = 0.3, color = "gray30", size = 1.4) +
  geom_line(data = rule_grid, aes(x = enroll, y = predicted),
            color = "#6a3d9a", linewidth = 1.3) +
  geom_vline(xintercept = c(40.5, 80.5, 120.5),
             linetype = "dashed", color = "gray50") +
  labs(title = "Maimonides' Rule: Predicted vs. Actual Class Size",
       subtitle = "Purple line = rule-based prediction. Gray points = actual class sizes.",
       x = "School Enrollment", y = "Class Size") +
  theme_minimal(base_size = 12)
```

**Schools don't perfectly comply.** Angrist and Lavy use *only the variation in class size that the rule predicts* — setting aside school-level choices — to isolate a causal effect. Their key finding: smaller classes improve test scores, especially for disadvantaged students. We'll walk through how this actually works in a moment.

## How Fuzzy RDD Actually Works {.smaller}

When the rule *shifts* treatment probability but doesn't perfectly determine it, fuzzy RDD combines **two jumps** at the cutoff into one effect estimate.

::: incremental
1.  **Measure the jump in treatment at the cutoff (the "first stage").** For Maimonides: at enrollment = 41, actual average class size drops by about **8 students** — *not* the full \~20 the rule predicts, because some schools don't split as required

2.  **Measure the jump in the outcome at the cutoff.** Say test scores rise by **0.2 standard deviations** across the same threshold

3.  **Scale: outcome jump ÷ treatment jump.** *If* we believe the cutoff affects outcomes **only through** class size, each 1-student drop in class size explains 0.2 SD ÷ 8 students = **0.025 SD per student** of test-score improvement
:::

. . .

$$\hat{\tau}_{fuzzy} = \frac{\text{jump in outcome}}{\text{jump in treatment}}$$

**Why "treatment" and not "treatment probability"?** When treatment is **binary** (e.g., "did they enroll in the program?"), the denominator is a probability between 0 and 1 — and you'll often see the formula written that way. When treatment is **continuous** (here: class size), the denominator is the jump in the treatment variable itself, in its own units. For Maimonides: $\frac{0.2 \text{ SD}}{8 \text{ students}} = 0.025$ SD per student.

## What We Just Did: Instrumental Variables {.smaller}

This is **instrumental variables (IV)** in action — a causal-inference technique for when treatment isn't randomly assigned. The "instrument" here is the cutoff itself: it shifts treatment but, we hope, doesn't otherwise affect the outcome.

**IV in general needs three things:**

::: incremental
1.  The instrument actually **changes treatment** — we showed this with the jump in class size
2.  The instrument is **as good as random** with respect to the outcome — true at the cutoff, where a 4-week age or 1-student enrollment difference can't plausibly sort people on unobservables
3.  The instrument affects the outcome **only through treatment** — the strong assumption we unpack next
:::

. . .

**Why fuzzy RDD is a relatively well-behaved form of IV:**

::: incremental
-   The first stage is **visible in a plot** — you can literally see whether the cutoff moves treatment
-   Assignment at the cutoff is nearly random *by construction*
-   "Only through" is easier to argue about a narrow threshold than for an instrument coming from somewhere else in the world
:::

. . .

*We won't cover standalone IV in this course, it's a whole separate methodology, and more controversial than its fuzzy-RDD cousin.*

## The Catch: Why Fuzzy Estimates Deserve Scrutiny {.smaller}

Fuzzy RDD requires a strong assumption — and it can mechanically amplify what it reports.

::: incremental
-   **The "only through" part is a strong assumption.** For Maimonides, is class size really the *only* thing that changes at enrollment = 41? What about teacher assignments, classroom resources, peer composition, or parents sorting across schools? If any of these also moves the outcome, we could be over-estimating the effect.

-   **Scaling amplifies noise.** Dividing a small, imprecise outcome jump by a small, imprecise treatment jump produces a bigger estimate *and* a wider confidence interval. A modest reduced-form result can become a dramatic-sounding fuzzy estimate
:::

. . .

::: callout-important
## Bottom Line

A sharp RDD result is easier to defend than a fuzzy one. When you encounter a fuzzy RDD in a paper or report, ask: *what else might be changing at that cutoff?* and *how strong is the first stage?* Be more skeptical than you would be with a sharp design.
:::

## Fuzzy RDD in `rdrobust` {.smaller}

```{r}
#| echo: true
#| eval: false

# Fuzzy RDD: pass the actual treatment indicator to `fuzzy`.
# rdrobust handles the two-jump scaling for you — but the scaling
# is only as honest as the "only through" assumption on the prior slide.
rd_fuzzy <- rdrobust(y = outcome, x = running_var, c = cutoff,
                     fuzzy = treatment_indicator)
summary(rd_fuzzy)
```

. . .

::: callout-tip
## Sharp or Fuzzy?

If the cutoff **perfectly determines** treatment → Sharp RDD. If the cutoff **shifts the probability** of treatment → Fuzzy RDD. When in doubt, plot the treatment probability against the running variable. If it's a step function, it's sharp.
:::

# Part 6: Validity Checks

## Manipulation Testing: The McCrary Density Test {.smaller}

::: incremental
-   If people can **manipulate** the running variable, we'd see **bunching** just above (or below) the cutoff

-   The McCrary (2008) density test checks: is the distribution of the running variable smooth through the cutoff?

-   A "pile-up" on one side suggests manipulation — and invalidates the design
:::

## What Manipulation Looks Like {.smaller}

```{r}
#| echo: true

set.seed(30317)

# Build a side-by-side comparison: clean vs. manipulated scores.
n_bunch       <- 80   # ~16% of students "cheated" into the 80-82 range
bunched_score <- c(
  runif(500 - n_bunch, 60, 100),  # baseline uniform
  runif(n_bunch,        80, 82)   # concentrated spike just above cutoff
)

df_density <- bind_rows(
  tibble(score = df$score,      condition = "Clean: no manipulation"),
  tibble(score = bunched_score, condition = "Manipulated: bunching above cutoff")
)

ggplot(df_density, aes(x = score)) +
  geom_histogram(bins = 40, fill = "gray70", color = "white") +
  geom_vline(xintercept = cutoff, linetype = "dashed",
             color = "red", linewidth = 1) +
  facet_wrap(~ condition) +
  labs(title = NULL,
       x = "Test Score", y = "Count") +
  theme_minimal(base_size = 13)
```

**Left:** smooth density through the cutoff — what we want. **Right:** a clear pile-up just above 80 — a red flag that students may have manipulated their scores to qualify.

## Formal Manipulation Test {.smaller}

```{r}
#| echo: true

# The rddensity package provides a formal test
library(rddensity)

density_test <- rddensity(X = df$score, c = cutoff)
summary(density_test)
```

. . .

A non-significant p-value means we **fail to reject** the null of no manipulation — which is what we want. If significant, the RDD may be compromised.

## Placebo Tests {.smaller}

::: incremental
-   **Placebo cutoffs:** Run the RDD at *fake* cutoff values where there's no treatment. If you find "effects," something is wrong.

-   **Covariate balance:** Check whether *pre-determined characteristics* (demographics, prior outcomes) show a jump at the cutoff. They shouldn't.

-   **Donut hole test:** Remove observations *very* close to the cutoff and re-estimate. If the result changes dramatically, those boundary observations may be problematic.
:::

## Placebo Cutoffs in Action {.smaller}

```{r}
#| echo: true

# Run rdrobust at real and fake cutoffs; only the real one should show an effect
placebo_cutoffs <- c(70, 75, 80, 85, 90)   # 80 is the real cutoff

placebo_results <- map_dfr(placebo_cutoffs, function(c_val) {
  rd <- rdrobust(y = df$completion, x = df$score, c = c_val)
  tibble(cutoff   = c_val,
         is_real  = c_val == 80,
         estimate = rd$coef[1],
         ci_lower = rd$ci[1, 1],
         ci_upper = rd$ci[1, 2])
})

ggplot(placebo_results, aes(x = cutoff, y = estimate, color = is_real)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c(`TRUE` = "#e31a1c", `FALSE` = "gray40"),
                     labels = c(`TRUE` = "Real cutoff", `FALSE` = "Placebo")) +
  labs(title = "RDD Estimates at Real vs. Placebo Cutoffs",
       x = "Cutoff Value (Test Score)",
       y = "RDD Estimate (completion rate)", color = NULL) +
  theme_minimal(base_size = 13)
```

Only the cutoff at 80 should show a real effect. Placebo cutoffs should be indistinguishable from zero — that's what we see.

## The Donut Hole Test {.smaller}

```{r}
#| echo: true

donut_values <- c(0, 0.5, 1, 2)   # donut radius in test-score points

donut_results <- map_dfr(donut_values, function(d) {
  keep <- abs(df$score - cutoff) >= d
  rd <- rdrobust(y = df$completion[keep], x = df$score[keep], c = cutoff)
  tibble(donut    = d,
         estimate = rd$coef[1],
         ci_lower = rd$ci[1, 1],
         ci_upper = rd$ci[1, 2])
})

ggplot(donut_results, aes(x = donut, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.08) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Donut Hole Test: Stability When Near-Cutoff Points Are Excluded",
       x = "Donut Radius (test-score points excluded on each side)",
       y = "RDD Estimate (completion rate)") +
  theme_minimal(base_size = 13)
```

If the estimate is **stable** as we exclude observations closer to the cutoff, that's reassuring. Wild swings would suggest those boundary observations are driving — or contaminating — the result.

## Covariate Balance Check {.smaller}

```{r}
#| echo: true

# Simulate a pre-determined covariate (family income, in $000s)
# — should NOT jump at the cutoff if the RDD is credible
set.seed(30317)
df$family_income <- 30 + 0.4 * df$score + rnorm(nrow(df), 0, 5)

ggplot(df, aes(x = score, y = family_income, color = group)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  labs(title = "Covariate Balance: Family Income by Test Score",
       subtitle = "No jump at the cutoff = good news for the RDD",
       x = "Test Score", y = "Family Income ($000s)") +
  scale_color_manual(values = c("No Scholarship" = "#1f78b4",
                                "Scholarship"    = "#e31a1c")) +
  theme_minimal(base_size = 14)
```

Covariates should pass smoothly through the cutoff. A jump here would be a red flag.

## Sensitivity to Bandwidth {.smaller}

```{r}
#| echo: true

# Check if the estimate is stable across bandwidth choices
library(rdrobust)

bandwidths <- c(2, 3, 5, 7, 10, 15)   # in test-score points

sensitivity <- map_dfr(bandwidths, function(h) {
  rd <- rdrobust(y = df$completion, x = df$score, c = cutoff, h = h)
  tibble(bandwidth = h,
         estimate  = rd$coef[1],
         ci_lower  = rd$ci[1, 1],
         ci_upper  = rd$ci[1, 2])
})

ggplot(sensitivity, aes(x = bandwidth, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Sensitivity to Bandwidth Choice",
       x = "Bandwidth (test-score points)",
       y = "RD Estimate (completion rate)") +
  theme_minimal(base_size = 14)
```

If the estimate is **stable** across bandwidths, that's reassuring. Wild swings suggest fragility.

# Part 7: Practical Advice

## Checklist + Common Pitfalls {.smaller}

**Before you trust an RDD — yours or someone else's:**

::: incremental
1.  Running variable is **precisely measured** and hard to manipulate
2.  **Density** is smooth through the cutoff (no bunching)
3.  **Covariates balance** at the cutoff
4.  Estimate is **robust** to different bandwidths
5.  **Placebo cutoffs** show null effects
6.  If **fuzzy**, first stage is strong
:::

. . .

**Pitfalls to avoid:**

::: incremental
-   Too-wide bandwidth — people far from the cutoff aren't comparable
-   High-order polynomials — can manufacture fake jumps (Gelman & Imbens 2019)
-   Forgetting **LATE** — your estimate applies at the cutoff, not everywhere
-   Not showing the plot — RDD is visual; the picture *is* the evidence
-   Assuming sharp when it's fuzzy — biases the estimate toward zero
:::

## Readings {.smaller}

-   **Huntington-Klein, *The Effect*, Ch. 20** — intuitive treatment with great examples (free online)
-   **Cunningham, *Causal Inference: The Mixtape*, Ch. 6** — detailed chapter with R code (free online)
-   **Cattaneo, Idrobo, and Titiunik (2020)** — *A Practical Introduction to Regression Discontinuity Designs*. The definitive applied guide
-   **Lee and Lemieux (2010)** — "Regression Discontinuity Designs in Economics." The classic review

# Wrapping Up

## What We Covered Today {.smaller}

::: incremental
1.  **One last DiD walkthrough** — Medicaid expansion and mortality
2.  **The RDD idea** — cutoffs as local randomization
3.  **The graphical recipe** — plot, fit, bandwidth, estimate the jump
4.  **Estimation** — from OLS to `rdrobust` with optimal bandwidth selection
5.  **Policy examples** — shingles/dementia, Medicare at 65
6.  **Sharp vs. fuzzy** — with Maimonides' Rule as the fuzzy example
7.  **Validity checks** — manipulation, placebo cutoffs, donut holes, bandwidth sensitivity
:::

# Doing Quant Work as an MPA

## Four Habits That Separate Good Analysis from Statistical Theater {.smaller}

::: incremental
1.  **Start with the question, not the method.** A simple t-test on the right comparison beats a fancy model on the wrong one. Design follows the question — not the other way around

2.  **You'll mostly be a consumer, not a producer.** Most of your career you'll read, commission, and push back on analyses. Learn to ask: *"What's the comparison? How do you know they're comparable? Who does this estimate apply to?"*

3.  **Show the picture.** Every method we covered has a visual signature — parallel pre-trends for DiD, a jump at the cutoff for RDD, balanced baselines for experiments. If the plot doesn't tell the story, be suspicious of the coefficient that does

4.  **Be honest about the limits.** Every estimate has assumptions. Every result applies to a specific group. Saying "we can't tell from this data" is often more useful — and more credible — than overselling a small effect
:::

## What This Course Was Really About {.smaller}

::: callout-tip
## The Real Lesson

We spent the semester on methods — OLS, fixed effects, DiD, RDD, experiments. The real lesson is **mental discipline**: being precise about *what question you're answering*, *what assumptions you're making*, and *what your result does and doesn't mean*.

You don't need to be the person who runs every analysis. You do need to be the person who knows **what to ask, what to trust, and what to reject**. That's how quantitative work actually shapes policy — and it's what makes you valuable.
:::

## For Your Capstone {.smaller}

::: callout-tip
## Think About This Now

Does your policy area have **eligibility thresholds**, **age cutoffs**, **geographic boundaries**, or **test score requirements**? If so, you may have a natural RDD waiting to be discovered. The key ingredients: a running variable you can measure, a cutoff that changes treatment status, and an argument that people just above and below are comparable.

Even if you don't run the analysis yourself, knowing how to *read* RDD (and DiD, and IV, and experiments) critically is a career-long skill.
:::