---
title: "Difference-in-Differences for Policy Evaluation"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

## Where We Left Off {.smaller}

::: incremental
-   Last class, we learned **fixed effects**: control for the unit, use only within-unit variation over time

-   We also saw that **two-way fixed effects** (unit + time) has problems when treatment timing varies across units

-   Today we learn the method that builds directly on those ideas: **Difference-in-Differences (DiD)**

-   DiD is probably the **most commonly used** causal inference tool in applied policy research
:::

# Part 1: The DiD Idea

## A Story About Cholera {.smaller}

::: incremental
-   In the 1850s, cholera was killing thousands of people in London. Nobody knew what caused it.

-   The physician John Snow had a theory: cholera spread through contaminated water, not through the air (the prevailing view at the time)

-   Two water companies served overlapping parts of London. In 1852, the **Lambeth Company** moved its water intake upstream, away from the sewage. The other, **Southwark & Vauxhall**, did not.

-   Households served by the two companies lived in the same neighborhoods. The only difference was where their water came from.
:::

## The Problem: Before-and-After Isn't Enough {.smaller}

::: incremental
-   After the Lambeth Company moved its pipe, cholera deaths in Lambeth-served households dropped from 130.1 to 84.9 per 10,000 people. That's a decline of 45.2 deaths.

-   Can we conclude that clean water saved 45 lives per 10,000? Not yet.

-   Cholera outbreaks vary from year to year for all sorts of reasons: weather, population density, sanitation practices. Maybe 1854 was just a less severe year overall.

-   **Time is a confounder.** If we only look at one group, we can't separate the effect of the treatment from things that were changing anyway.
:::

## The Solution: A Comparison Group {.smaller}

::: incremental
-   Snow had something better than a simple before-and-after comparison. He also had data on households served by **Southwark & Vauxhall**, whose water was still contaminated.

-   If 1854 was simply a better year for cholera, we'd expect deaths to drop in *both* groups.

-   In fact, deaths in the Southwark & Vauxhall areas *rose* from 134.9 to 146.6 per 10,000, an increase of 11.7.

-   So the overall environment was getting *worse* for cholera in 1854, not better. The decline in Lambeth households happened *despite* that trend.
:::

## Snow's 2×2 Table {.smaller}

| | 1849 (Before) | 1854 (After) | Change |
|---|---|---|---|
| **Lambeth** (clean water) | 130.1 deaths per 10,000 | 84.9 | -45.2 |
| **Southwark & Vauxhall** (contaminated) | 134.9 | 146.6 | +11.7 |

. . .

**Step 1:** The change in the treated group (Lambeth): 84.9 - 130.1 = **-45.2**

. . .

**Step 2:** The change in the control group (Southwark & Vauxhall): 146.6 - 134.9 = **+11.7**

. . .

**Step 3:** The DiD estimate: (-45.2) - (+11.7) = **-56.9 deaths per 10,000**

## What Did We Just Do? {.smaller}

::: incremental
-   The Lambeth deaths fell by 45.2. But we can't credit all of that to clean water, because other things were changing too.

-   The Southwark & Vauxhall deaths *rose* by 11.7 over the same period. They serve as our benchmark for what would have happened to Lambeth without the clean water.

-   So the true effect of clean water is the Lambeth change *minus* the Southwark & Vauxhall change: -45.2 - 11.7 = -56.9

-   We're not comparing the two groups at one point in time (their levels were similar but not identical). We're comparing how much each group *changed*. That's the "difference in differences."
:::

## The DiD Formula

$$\widehat{\delta}_{DiD} = (\bar{Y}_{treat}^{post} - \bar{Y}_{treat}^{pre}) - (\bar{Y}_{control}^{post} - \bar{Y}_{control}^{pre})$$

::: {.callout-note}
## In English
Take the **change** in the treated group. Subtract the **change** in the control group. What's left is the treatment effect: the part of the change that can't be explained by common time trends.
:::

## The Key Assumption {.smaller}

::: incremental
-   This only works if we believe that, **without the treatment**, Lambeth households would have experienced the same trend as Southwark & Vauxhall households

-   This is called the **parallel trends assumption**, and it is the foundation of every DiD design

-   In Snow's case, it's plausible: the households were in the same neighborhoods, exposed to the same weather and sanitation conditions. The only difference was the water source.

-   We'll spend more time on this assumption later. For now, the point is that DiD requires a credible argument that the two groups were on the same trajectory before treatment.
:::

## Why Should MPAs Care?

::: incremental
-   DiD is everywhere in policy evaluation: minimum wage studies, Medicaid expansion, education reforms, environmental regulations

-   If you read policy evaluations, you *will* encounter DiD. If you do evaluation work, you will likely use it.

-   Real applications get more complex than Snow's 2×2 table (multiple time periods, many units, staggered policy adoption), and we'll work through those complications today
:::

## Today's Roadmap

::: incremental
1.  **DiD with real data in R**: California's organ donation policy
2.  **The parallel trends assumption**: when is DiD credible?
3.  **Event studies**: visualizing effects over time
4.  **Staggered adoption**: when different units get treated at different times
5.  **Modern solutions**: new methods that fix problems with classic TWFE
6.  **Practical advice**: a checklist for credible DiD designs
:::

# Part 2: DiD With Real Data

## California's Organ Donation Policy {.smaller}

::: incremental
-   In July 2011, **California** changed how it asked people about organ donation at the DMV

-   Previously, the question was opt-in: "Would you like to be an organ donor?" (most people skipped it)

-   California switched to **active choice**: "Do you wish to register as an organ donor?" with a required yes or no answer

-   **The question:** Did this change increase organ donation registration rates?

-   We have quarterly data on 27 states from Q4 2010 through Q1 2012. Only California changed its policy.
:::

## Loading the Data {.smaller}

This example comes from Kessler and Roth (2014) and is used in Huntington-Klein's *The Effect*.

```{r}
#| echo: true

library(tidyverse)
library(fixest)
library(causaldata)

# Load the data
od <- organ_donations

# What does it look like?
head(od, 12)
```

Each row is a state-quarter. `Rate` is the organ donation registration rate. California's policy changes in Q3 2011.

## Step 1: Look at the Raw Trends {.smaller}

Before running any model, plot the data.

```{r}
#| echo: true

# Average donation rate: California vs. everyone else
od_means <- od %>%
  mutate(group = ifelse(State == "California", "California", "Other States")) %>%
  group_by(group, Quarter) %>%
  summarise(avg_rate = mean(Rate), .groups = "drop")

ggplot(od_means, aes(x = Quarter, y = avg_rate, color = group, group = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = "Q32011", linetype = "dashed", color = "gray40") +
  labs(title = "Organ Donation Rates: California vs. Other States",
       x = "Quarter", y = "Donation Rate", color = "") +
  theme_minimal(base_size = 14)
```

Look at what happens after the dashed line. Do the trends diverge? Not really. That's already a hint about what the regression will find. But let's work through the numbers to be sure.

## Step 2: Create the Treatment Variable {.smaller}

```{r}
#| echo: true

# A unit is "treated" if it is California AND in the post-treatment period
od <- od %>%
  mutate(
    California = ifelse(State == "California", 1, 0),
    Post = ifelse(Quarter %in% c("Q32011", "Q42011", "Q12012"), 1, 0),
    Treated = California * Post
  )

# Check: how many treated observations?
od %>% count(California, Post, Treated)
```

`Treated = 1` only for California in Q3 2011 and after. Everything else is a control observation.

## Step 3: The DiD by Hand {.smaller}

Just like we did with Snow's cholera data, let's compute the DiD estimate manually before running a regression.

```{r}
#| echo: true

# Average donation rates by group and period
did_table <- od %>%
  mutate(group = ifelse(California == 1, "California", "Other States"),
         period = ifelse(Post == 1, "After", "Before")) %>%
  group_by(group, period) %>%
  summarise(avg_rate = mean(Rate), .groups = "drop") %>%
  pivot_wider(names_from = period, values_from = avg_rate) %>%
  mutate(Change = After - Before)

did_table
```

. . .

```{r}
#| echo: true

# The DiD estimate: difference in the changes
did_table$Change[did_table$group == "California"] -
  did_table$Change[did_table$group == "Other States"]
```

## Step 4: The DiD Regression {.smaller}

Now let's get the same answer from a regression with proper standard errors. This is the same `feols()` syntax from last class.

```{r}
#| echo: true

# TWFE regression: state fixed effects + quarter fixed effects
did_model <- feols(Rate ~ Treated | State + Quarter, data = od)
summary(did_model)
```

. . .

::: {.callout-note}
## Interpretation
The coefficient on `Treated` is the DiD estimate, the same number we computed by hand. It's small and not statistically significant. The active-choice policy **did not meaningfully increase** organ donation rates in California relative to the other states. Not every policy works, and DiD is just as useful for showing null effects as positive ones.
:::

## The Connection to Fixed Effects {.smaller}

::: incremental
-   The DiD regression is a **two-way fixed effects** model, which you learned last class:

$$Y_{it} = \alpha_{i} + \alpha_{t} + \beta \cdot \text{Treated}_{it} + \epsilon_{it}$$

-   $\alpha_i$ is a state fixed effect (absorbs level differences between California and other states)
-   $\alpha_t$ is a quarter fixed effect (absorbs trends common to all states)
-   $\beta$ is the DiD estimate

-   DiD builds directly on fixed effects. The new ingredient is the treatment variable that "turns on" for specific units at a specific time.
:::

## A Famous Example: Card & Krueger (1994) {.smaller}

The most famous DiD study in economics uses this same logic. In April 1992, **New Jersey** raised its minimum wage. Neighboring **Pennsylvania** did not.

| | Before (Feb 1992) | After (Nov 1992) | Change |
|---|---|---|---|
| **New Jersey** (treated) | 20.44 FTE | 21.03 FTE | +0.59 |
| **Pennsylvania** (control) | 23.33 FTE | 21.17 FTE | -2.16 |

. . .

**DiD Estimate:** (+0.59) - (-2.16) = **+2.76 FTE employees**

. . .

The minimum wage increase was associated with an *increase* in employment, the opposite of the standard prediction. This is one of the most influential papers in labor economics.

# Part 3: The Parallel Trends Assumption

## The Key Assumption {.smaller}

::: incremental
-   We introduced this briefly with Snow's data. Now let's dig into it.

-   DiD requires one critical assumption: **without the treatment**, the treated group would have followed the same trend as the control group. This is called the **parallel trends assumption**.

-   In the organ donation example, it means: if California had *not* changed its DMV question, California's donation rate would have changed at the same rate as the other states.

-   This is an assumption about a **counterfactual** that we can never observe. We can check whether trends were parallel *before* the treatment, but the assumption is about what would have happened *after*.
:::

## Levels vs. Trends {.smaller}

::: {.callout-important}
## A Common Confusion

Parallel trends does **not** require that the groups are at the same level. In the organ donation plot, California's rate was at a different level than the other states. That's fine. What matters is that they were *changing at the same rate*. Similarly, in Snow's data, the Lambeth and non-Lambeth areas had slightly different death rates in 1849, but DiD still works because we're comparing changes, not levels.
:::

::: incremental
-   Two cities can have very different crime rates but still have parallel trends if crime is rising (or falling) at the same rate in both

-   If they are changing at different rates before the policy, any post-treatment difference could just be a continuation of that pre-existing divergence
:::

## Good vs. Bad Parallel Trends {.smaller}

```{r}
#| echo: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

df <- tibble(
  time = rep(1:10, times = 4),
  group = rep(c("Treatment", "Control"), each = 10, times = 2),
  panel = rep(c("Parallel Prior Trends", "Converging Prior Trends"), each = 20),
  outcome = c(
    c(2, 2, 3, 4, 5, 7, 9, 10, 9.8, 9.5),
    c(1, 1, 2, 3, 3.5, 4.5, 5.5, 6, 5.5, 6),
    c(4, 3.5, 4, 5, 6, 6.5, 7.2, 9, 10, 10.5),
    c(1, 1.2, 2.2, 3.2, 4.8, 6.5, 8.5, 9.5, 9.7, 10)
  )
)

make_plot <- function(data, panel_title) {
  ggplot(data, aes(x = time, y = outcome, color = group)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    geom_vline(xintercept = 6, linetype = "dashed") +
    annotate("rect", xmin = 6, xmax = 10.5, ymin = -Inf, ymax = Inf, alpha = 0.1) +
    annotate("text", x = 6.3, y = max(data$outcome) * 0.2, label = "Treatment\nPeriod", hjust = 0) +
    scale_color_manual(values = c("Treatment" = "black", "Control" = "gray40")) +
    labs(title = panel_title, x = "Time", y = "Outcome", color = "Group") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom")
}

p1 <- make_plot(df %>% filter(panel == "Parallel Prior Trends"), "(a) Parallel Prior Trends")
p2 <- make_plot(df %>% filter(panel == "Converging Prior Trends"), "(b) Converging Prior Trends")

p1 + p2
```

Panel (a): pre-trends are parallel, so the post-treatment gap is plausibly causal. Panel (b): the control group was already converging before treatment, so the "effect" may just be a continuation of that pre-existing trend.

## When Is Parallel Trends Plausible? {.smaller}

::: incremental
-   **More plausible:**
    -   Units in similar environments (Snow's households were on the same streets; Card & Krueger chose neighboring NJ and PA, not NJ and Alaska)
    -   Units selected for treatment based on a fixed characteristic, not a trend
    -   Short time horizons where trends are less likely to diverge

-   **Less plausible:**
    -   Comparing a booming city to a declining one
    -   Units that self-selected into treatment based on anticipated trends (e.g., states that expanded Medicaid because their health outcomes were already improving)
    -   Very long time horizons with structural changes
:::

## Testing Parallel Trends {.smaller}

::: incremental
-   We can't prove parallel trends holds, but we can look for evidence against it

-   **Placebo tests in pre-treatment periods:** Run DiD as if treatment happened earlier. If you find "effects" before the real treatment, something is wrong.

-   **Event study plots** (which we'll build later) are the standard way to do this. They show the treatment-control difference at each time period. Before treatment, those differences should hover around zero.

-   Finding null pre-treatment effects doesn't *prove* the assumption, but it's reassuring. Finding significant pre-treatment effects is a serious red flag.
:::

## Anticipation Effects {.smaller}

::: incremental
-   What if people change behavior **before** the policy takes effect?

    -   Workers might change jobs before a minimum wage increase

    -   Firms might hire ahead of a regulation deadline

-   Anticipation violates the assumption that treatment "turns on" at a specific time. It can also make pre-trends look non-parallel even if the design is otherwise sound.

-   **Solution:** Backdate the treatment period or exclude the periods just before treatment
:::


# Part 4: Event Studies

## What If the Effect Isn't Constant? {.smaller}

::: incremental
-   Our EITC simulation assumed the treatment effect was an immediate, constant +3 points. But that's often unrealistic.

-   A job training program might not show results for a year

-   A tax credit might take time as people learn about it and adjust their behavior

-   An **event study** lets us estimate the DiD effect **separately for each time period**, so we can see how the effect evolves

-   It also gives us a built-in test of parallel trends: the pre-treatment effects should be close to zero
:::

## How an Event Study Works {.smaller}

::: incremental
-   Instead of one "post" dummy, we create a dummy for each time period relative to treatment

-   The period just before treatment (event time = -1) is the **reference category**, so all effects are measured relative to that point

-   **Pre-treatment coefficients** should hover around zero if parallel trends holds. This is the placebo test we discussed earlier.

-   **Post-treatment coefficients** show how the effect unfolds: does it appear immediately? Grow over time? Fade?
:::

## Building an Event Study {.smaller}

We'll simulate a new dataset where the EITC effect **grows over time** as more people learn about and respond to the program.

```{r}
#| echo: true

set.seed(30317)

# 60 units (30 treated, 30 control), 12 years, treatment in year 7
unit_fes <- tibble(unit = 1:60, unit_fe = rnorm(60, 0, 3))

event_data <- expand_grid(unit = 1:60, year = 1:12) %>%
  left_join(unit_fes, by = "unit") %>%
  mutate(
    treated_unit = ifelse(unit <= 30, 1, 0),
    event_time = year - 7,
    # Dynamic treatment effect: grows over time after treatment
    te = case_when(
      treated_unit == 0 ~ 0,  # Control units: no effect
      event_time < 0    ~ 0,  # Before treatment: no effect
      event_time == 0   ~ 1.5,
      event_time == 1   ~ 2.5,
      event_time == 2   ~ 3.2,
      event_time == 3   ~ 3.5,
      TRUE              ~ 4.0  # Levels off
    ),
    outcome = 50 + unit_fe + 0.5 * year + te + rnorm(n(), 0, 2)
  )
```

## The Event Study Regression {.smaller}

```{r}
#| echo: true

# i() creates a dummy for each event_time period
# ref = -1 sets the period before treatment as the reference
es_model <- feols(
  outcome ~ i(event_time, treated_unit, ref = -1) | unit + year,
  data = event_data
)

# fixest has a built-in event study plot
iplot(es_model,
      xlab = "Periods Relative to Treatment",
      ylab = "Estimated Effect",
      main = "Event Study: Effect of EITC Expansion Over Time")
```

## Reading the Event Study Plot {.smaller}

::: incremental
-   **Pre-treatment coefficients** (left of 0): they hover around zero with confidence intervals crossing zero. This supports parallel trends.

-   **The jump at time 0**: the immediate treatment effect (about 1.5 points in our simulation)

-   **Post-treatment growth**: the effect increases over time as more people respond to the program, then levels off around 4 points

-   **Confidence intervals**: if they include zero after treatment, we can't rule out no effect. Here the post-treatment intervals are clearly above zero.
:::

. . .

::: {.callout-tip}
## For Your Reports
Event study plots are the **standard way** to present DiD results in policy research. They show both the treatment effect and the parallel trends evidence in a single figure. If you do a DiD analysis, include one.
:::

# Part 5: The Staggered Adoption Problem

## What If Treatment Happens at Different Times? {.smaller}

::: incremental
-   So far, we've had a clean setup: one group treated, one group not, at a single point in time

-   In practice, policies are often adopted **at different times** by different units

    -   States expanded Medicaid in 2014, 2015, 2016, or not at all

    -   Cities adopted minimum wage increases in different years

    -   Schools implemented new curricula on different schedules

-   This is called **staggered adoption**, and it creates serious problems for classic TWFE
:::

## Visualizing Staggered Treatment {.smaller}

```{r}
#| echo: false

library(tidyverse)

stagger_viz <- tibble(
  State = c("Alabama", "Colorado", "Delaware", "Florida", "Georgia"),
  treat_year = c(2015, 2016, 2018, NA, NA)
) %>%
  crossing(Year = 2012:2020) %>%
  mutate(
    status = case_when(
      is.na(treat_year) ~ "Never Treated",
      Year < treat_year ~ "Not Yet Treated",
      Year >= treat_year ~ "Treated"
    ),
    status = factor(status, levels = c("Treated", "Not Yet Treated", "Never Treated"))
  )

ggplot(stagger_viz, aes(x = Year, y = State, fill = status)) +
  geom_tile(color = "white", linewidth = 1) +
  scale_fill_manual(values = c("Treated" = "#2C7BB6", "Not Yet Treated" = "#FFFFBF",
                                "Never Treated" = "#D7191C")) +
  labs(title = "Staggered Treatment Adoption",
       fill = "Treatment Status") +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())
```

Different states get treated at different times. Who serves as the "control" for whom?

## Why Classic TWFE Breaks {.smaller}

::: incremental
-   Classic TWFE estimates a **single treatment coefficient** from a weighted average of *all possible 2×2 comparisons*

-   The problem: some of those comparisons are **"forbidden"**

    -   TWFE compares **newly-treated units** to **already-treated units**

    -   An already-treated unit is being used as a "control," but it's already affected by treatment!

-   If treatment effects **change over time** (grow, fade, or differ across groups), these forbidden comparisons contaminate the estimate

-   The weights can even be **negative**, meaning the estimate can show a negative effect when the true effect is positive for everyone
:::

## An Intuitive Example {.smaller}

::: {.callout-warning}
## The Negative Weights Problem

Imagine a job training program that helps everyone, but the effect grows over time. Early adopters have large effects by now (they've had the program for years). Late adopters have small effects (they just started).

TWFE compares late adopters (small positive effect) to early adopters (large positive effect). The "change" in the early-adopter "controls" is actually driven by their growing treatment effect, making it look like the new treatment *reduced* outcomes. The estimated effect can flip negative, even though **everyone benefited**.

This is exactly the kind of dynamic effect we saw in our event study. If we had staggered adoption in that setting, TWFE would give us the wrong answer.
:::

## When Is TWFE Fine? {.smaller}

::: incremental
-   **TWFE is fine if:**
    -   Treatment effects are constant over time and across groups
    -   All units are treated at the same time (the classic 2×2 case)
    -   You have a "never-treated" group and only compare to them

-   **TWFE is problematic if:**
    -   Treatment effects build up, fade, or vary across cohorts
    -   You have staggered adoption with no never-treated group
    -   You suspect dynamic or heterogeneous effects

-   In practice, constant effects are a strong assumption. Most policies take time to work, which means effects change over time. When in doubt, use the newer methods we'll see next.
:::

# Part 6: Modern Solutions

## The Good News {.smaller}

Econometricians have developed several methods that fix the staggered DiD problem. The two most widely used:

::: incremental
-   **Callaway and Sant'Anna (2021):** Estimate separate treatment effects for each treatment cohort at each time period, then aggregate. R package: `did`.

-   **Sun and Abraham (2021):** An interaction-weighted event-study estimator that avoids the contamination. Available in `fixest` via `sunab()`.

-   Both approaches share a key idea: **only make "clean" comparisons**. Treated units should only be compared to not-yet-treated or never-treated units, never to already-treated units.
:::

## Simulating Staggered Adoption {.smaller}

```{r}
#| echo: true

set.seed(30317)

# 40 units in 4 groups:
# 10 treated starting year 5, 10 in year 7, 10 in year 9, 10 never
stag_unit_fes <- tibble(unit = 1:40, unit_fe = rnorm(40, 0, 2))

stag_data <- expand_grid(unit = 1:40, year = 1:12) %>%
  left_join(stag_unit_fes, by = "unit") %>%
  mutate(
    cohort = case_when(
      unit <= 10 ~ 5,   # Early adopters
      unit <= 20 ~ 7,   # Middle adopters
      unit <= 30 ~ 9,   # Late adopters
      TRUE ~ Inf         # Never treated
    ),
    treated = ifelse(year >= cohort, 1, 0),
    years_treated = pmax(year - cohort, 0),
    # Dynamic effect: grows with time since treatment (1 + 0.5 per year)
    true_te = ifelse(is.finite(cohort) & treated == 1,
                     1 + 0.5 * years_treated, 0),
    outcome = 50 + unit_fe + 0.3 * year + true_te + rnorm(n(), 0, 1.5)
  )
```

The effect grows over time since treatment. This is exactly the scenario where TWFE breaks.

## Running Both Models {.smaller}

```{r}
#| echo: true

library(fixest)

# Classic TWFE: one coefficient for all treatment variation
twfe_model <- feols(outcome ~ treated | unit + year, data = stag_data)

# Sun & Abraham: only uses clean comparisons
# sunab() needs a finite number for never-treated units
stag_data_sa <- stag_data %>%
  mutate(cohort_sa = ifelse(is.infinite(cohort), 1000, cohort))

sa_model <- feols(outcome ~ sunab(cohort_sa, year) | unit + year,
                  data = stag_data_sa)
```

`sunab()` does the heavy lifting: it figures out which comparisons are clean and estimates the effect for each time period relative to treatment, just like our event study earlier.

## TWFE: A Single (Misleading) Coefficient {.smaller}

```{r}
#| echo: true

summary(twfe_model)
```

TWFE gives us one number for the treatment effect. But we built in an effect that grows over time, so a single number can't capture the full picture. Worse, the forbidden comparisons may be pulling this estimate in the wrong direction.

## Sun & Abraham: The Full Event Study {.smaller}

```{r}
#| echo: true

iplot(sa_model,
      xlab = "Periods Relative to Treatment",
      ylab = "Estimated Effect",
      main = "Sun & Abraham Event Study (Staggered DiD)")
```

The Sun & Abraham estimator correctly shows pre-treatment coefficients near zero and a growing post-treatment effect. This is what the true data-generating process looks like.

## The Key Takeaway {.smaller}

::: {.callout-important}
## Reading DiD Papers
If you encounter a DiD paper with **staggered treatment** that was published before roughly 2020, the estimates might be unreliable, especially if treatment effects are likely to grow or change over time. Papers using modern methods (Callaway & Sant'Anna, Sun & Abraham, or similar) have addressed this problem. When evaluating research, ask: "Is treatment staggered? If so, did they use a modern estimator?"
:::

# Part 7: Practical Advice

## A Checklist for Credible DiD Designs

::: incremental
1.  Is there a **clear treatment** and a **clear control group**?
2.  Is the **treatment timing plausibly exogenous**, not driven by anticipated outcomes?
3.  Do **pre-treatment trends look parallel**? (Plot them!)
4.  Are there **anticipation effects** that could blur the treatment timing?
5.  Is treatment **staggered**? If so, use modern methods.
6.  How **sensitive** are results to the choice of control group?
:::

## What DiD Estimates: The ATT {.smaller}

::: incremental
-   One important detail: DiD does not estimate the average treatment effect for everyone (the ATE). It estimates the **Average Treatment Effect on the Treated (ATT)**.

-   In Card & Krueger, the estimate tells us what the minimum wage increase did **in New Jersey**. It does not tell us what would happen if Pennsylvania also raised its wage, because PA might be different in ways that matter.

-   In a randomized experiment, treatment and control groups are balanced, so the ATE and ATT are the same. In DiD, groups are not randomly assigned, so we can only speak to the effect on the units that were actually treated.

-   This is usually fine for policy evaluation. Legislators want to know: "Did *our* policy work?" That's an ATT question.
:::

## Common Pitfalls

::: incremental
-   **Confusing levels with trends:** Groups can be at very different levels; what matters is that they *change at the same rate* pre-treatment

-   **Cherry-picking the control group** to manufacture parallel trends

-   **Not clustering standard errors** at the level of treatment assignment

-   **Using classic TWFE with staggered adoption** when effects are dynamic

-   **Ignoring anticipation effects** that shift behavior before the treatment takes effect
:::

## Readings {.smaller}

::: incremental
-   **Cunningham, *Causal Inference: The Mixtape***: Excellent chapter on DiD with real-world examples and R code. Freely available online.

-   **Huntington-Klein, *The Effect***: Very intuitive treatment of DiD and event studies. Also freely available.

-   **Callaway & Sant'Anna (2021)**: The key paper on staggered DiD. Technical, but the introduction is accessible.

-   **Goodman-Bacon (2021)**: "Difference-in-Differences with Variation in Treatment Timing." The paper that showed why TWFE breaks with staggered adoption.
:::

# Wrapping Up

## What We Covered Today {.smaller}

::: incremental
1.  **The DiD idea**: Snow's cholera data showed us how to use a comparison group to remove common time trends
2.  **DiD with real data**: walked through California's organ donation policy step by step in R
3.  **The parallel trends assumption**: the foundation of DiD credibility
4.  **Event studies**: the standard way to present DiD results and test assumptions
5.  **Staggered adoption**: why classic TWFE breaks and how to spot the problem
6.  **Modern solutions**: Sun & Abraham and Callaway-Sant'Anna fix the staggered problem
7.  **Practical advice**: a checklist for credible designs
:::

## What's Next

::: incremental
-   **Regression Discontinuity Designs**: exploiting arbitrary cutoffs for causal inference

-   **Instrumental Variables**: finding exogenous sources of variation
:::

## For Your Capstone

::: {.callout-tip}
## Think About This Now
Is there a **policy change** in your area of interest with a natural control group? A state that changed a law while a neighbor didn't? A program that was rolled out in some districts before others? If so, DiD might be the right tool, and now you know the modern best practices for using it.
:::
