---
title: "From Samples to Decisions"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
---

## Where We Are

We've learned how to:

-   Compare group averages -- like DMV wait times by neighborhood or voting wait times by race (conditional expectations)
-   Understand that samples vary (sampling distributions) and that bigger samples = smaller standard errors

. . .

**Today's big question:** How do we use sample data to make *decisions*?

-   Did that job training program *actually* work, or did we just get lucky with our sample?
-   Is support for Medicaid expansion above 50%?
-   Do body cameras actually reduce police use of force?

## The Law of Large Numbers

One of the most useful results in statistics:

. . .

::: callout-tip
## The Law of Large Numbers (Plain English)

As you collect more data, your sample average gets closer and closer to the true population average.

**Translation:** Bigger samples = more reliable estimates.
:::

. . .

This is why a well-designed survey of 1,500 people can predict election outcomes for 150 million voters + why your agency doesn't need to survey every single citizen.

## Seeing It In Action

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
library(tidyverse)

set.seed(42)

# Simulate cumulative means of DMV wait times as sample size grows
true_mean <- 50  # True average DMV wait time (minutes)

# Generate one sequence of observations
observations <- rnorm(500, mean = true_mean, sd = 15)

# Calculate cumulative mean
cumulative_data <- tibble(
  n = 1:500,
  cumulative_mean = cumsum(observations) / (1:500)
)

ggplot(cumulative_data, aes(x = n, y = cumulative_mean)) +
  geom_line(color = "#522D80", linewidth = 1) +
  geom_hline(yintercept = true_mean, linetype = "dashed", color = "#F66733", linewidth = 1.2) +
  annotate("text", x = 450, y = true_mean + 2, label = "True average wait = 50 min",
           color = "#F66733", fontface = "bold", size = 5) +
  labs(
    title = "As We Survey More Residents, Our Estimate Converges",
    x = "Number of Residents Surveyed",
    y = "Running Average Wait Time (minutes)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

With a small sample, our estimate of average DMV wait time bounces around. As we survey more residents, it settles down near the truth.

## Why the Normal Distribution Keeps Showing Up {.smaller}

Imagine your city's 311 system. Individual response times are *skewed* -- most calls get resolved quickly, but some take forever.

. . .

Now imagine running this experiment:

::: incremental
1.  **Draw a sample** of, say, 50 response times from the full database
2.  **Calculate the average** of those 50 response times
3.  **Write it down**, then throw the sample away
4.  **Repeat** this thousands of times, each time with a fresh random sample of 50
:::

## The Central Limit Theorem

::: callout-tip
## The Central Limit Theorem (Plain English)

When you repeat that experiment many times -- drawing a sample and computing its average -- the *distribution of those averages* follows a bell curve, **even if the raw data is skewed**.

The bigger each sample, the more bell-shaped the averages become.
:::

. . .

This is what makes hypothesis testing work. Every t-test, every confidence interval we'll compute relies on the sampling distribution being approximately normal -- and the CLT is why that's a safe assumption.

## The CLT In Action

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4
library(tidyverse)
library(patchwork)

set.seed(123)

nsims <- 5000

# The raw data: skewed 311 response times
raw_data <- rexp(10000, rate = 0.5)

# Draw 5,000 samples of size 5, compute each sample's average
means_n5 <- replicate(nsims, mean(rexp(5, rate = 0.5)))

# Draw 5,000 samples of size 50, compute each sample's average
means_n50 <- replicate(nsims, mean(rexp(50, rate = 0.5)))

p0 <- ggplot(tibble(x = raw_data), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "gray50", alpha = 0.7) +
  labs(title = "Raw 311 Response Times",
       subtitle = "Very skewed — not normal at all",
       x = "Response Time (days)", y = "Density") +
  theme_minimal(base_size = 12) +
  xlim(0, 12)

p1 <- ggplot(tibble(x = means_n5), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#522D80", alpha = 0.7) +
  labs(title = "5,000 Samples (n = 5)",
       subtitle = "Each bar = the average of 5 response times",
       x = "Sample Average", y = "Density") +
  theme_minimal(base_size = 12) +
  xlim(0, 8)

p2 <- ggplot(tibble(x = means_n50), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#F66733", alpha = 0.7) +
  stat_function(fun = dnorm, args = list(mean = mean(means_n50), sd = sd(means_n50)),
                linetype = "dashed", linewidth = 1) +
  labs(title = "5,000 Samples (n = 50)",
       subtitle = "Each bar = the average of 50 response times",
       x = "Sample Average", y = "Density") +
  theme_minimal(base_size = 12) +
  xlim(0, 8)

p0 + p1 + p2
```

Left: the raw data is skewed. Middle & right: we drew 5,000 samples and computed each sample's average. **Bigger samples → the averages look more and more like a bell curve.**

## The Practical Takeaway

::: callout-important
## What This Means For You

1.  **Bigger samples** give you more reliable estimates (LLN)
2.  **Sample averages** follow a predictable pattern—the normal distribution (CLT)
3.  Because of #2, we can use **statistical tests** to make decisions about policy effectiveness
:::

. . .

**Next up:** How do we turn this into a formal test?

# Hypothesis Testing

## A Classic Example: The Lady Tasting Tea

::: incremental
-   In the 1920s, a biologist named Muriel Bristol claimed she could tell whether tea or milk was added first to a cup

-   Statistician R.A. Fisher was skeptical, so he designed a test:

    -   Make 8 cups of tea, 4 each way
    -   Present them in random order
    -   Ask Bristol to identify which 4 had milk added first

-   **She got all 8 correct!**

-   What's the probability of getting all 8 right by pure luck?
:::

## Could She Just Be Lucky?

If she was just guessing randomly:

-   There's only **1 way** to pick all 4 correct cups
-   There are **70 total ways** to pick any 4 cups from 8

. . .

$$P(\text{all correct by luck}) = \frac{1}{70} \approx 0.014 = 1.4\%$$

. . .

**Conclusion:** A 1.4% chance is pretty small. It's hard to chalk that up to luck -- she probably really could tell the difference.

. . .

This is the core logic of every hypothesis test: **If nothing were going on, how often would we see results like this?**

## What Is a Hypothesis?

A **hypothesis** is just a statement about a population that we want to test.

. . .

**Policy examples:**

-   Does a job training program increase employment rates?
-   Is public support for a carbon tax above 50%?
-   Do body cameras reduce police use of force?

. . .

Important: we can't *prove* anything with data. We **can** ask how surprising our results would be if there were no real effect.

## The Null and Alternative Hypotheses

::: callout-note
## The Null Hypothesis (H₀)

The "boring" hypothesis—usually "no effect" or "no difference"

**Example:** The job training program has no effect on employment
:::

. . .

::: callout-note
## The Alternative Hypothesis (H₁)

What we're actually interested in—usually "there IS an effect"

**Example:** The job training program does affect employment
:::

. . .

We test the null hypothesis. If the data would be very surprising under H₀, we **reject** it.

## The Logic of Hypothesis Testing

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

library(tidyverse)

# Create a normal distribution showing rejection regions
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)

df <- tibble(x = x, y = y)

ggplot(df, aes(x = x, y = y)) +
  geom_line(linewidth = 1.2) +
  geom_area(data = filter(df, x < -1.96), fill = "#F66733", alpha = 0.5) +
  geom_area(data = filter(df, x > 1.96), fill = "#F66733", alpha = 0.5) +
  geom_vline(xintercept = c(-1.96, 1.96), linetype = "dashed", color = "#522D80") +
  annotate("text", x = 0, y = 0.2, label = "If Null is true,\nmost results land here",
           size = 5, fontface = "bold") +
  annotate("text", x = -2.8, y = 0.1, label = "Reject Null", color = "#F66733",
           size = 5, fontface = "bold") +
  annotate("text", x = 2.8, y = 0.1, label = "Reject Null", color = "#F66733",
           size = 5, fontface = "bold") +
  labs(
    title = "The Logic: If Null Were True, Would Our Result Be Surprising?",
    x = "Test Statistic",
    y = "Probability"
  ) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

If our result falls in the orange "tails," it would be very unlikely under H₀—so we reject it.

## Two Types of Mistakes

When we make a decision, we can be wrong in two ways:

|                     | H₀ Actually True | H₀ Actually False    |
|---------------------|------------------|----------------------|
| **Don't Reject H₀** | Correct!         | Missed a real effect |
| **Reject H₀**       | False alarm!     | Correct!             |

. . .

::: callout-warning
## Key Terms

-   **Type I Error (False Positive):** We said there's an effect, but there isn't
-   **Type II Error (False Negative):** We missed a real effect
:::

## Statistical Power: Can We Detect What Matters?

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

library(tidyverse)

x <- seq(-4, 5, length.out = 1000)
null_y <- dnorm(x, mean = 0, sd = 1)
alt_y <- dnorm(x, mean = 2, sd = 1)

df <- tibble(x = x, null = null_y, alt = alt_y)

ggplot(df, aes(x = x)) +
  geom_line(aes(y = null), linewidth = 1.2, color = "gray40") +
  geom_line(aes(y = alt), linewidth = 1.2, color = "#522D80") +
  geom_area(data = filter(df, x > 1.96), aes(y = alt), fill = "#522D80", alpha = 0.4) +
  geom_area(data = filter(df, x > 1.96), aes(y = null), fill = "#F66733", alpha = 0.3) +
  geom_vline(xintercept = 1.96, linetype = "dashed", color = "gray40") +
  annotate("text", x = -2, y = 0.25, label = "Null:\nNo effect",
           size = 4, color = "gray40", fontface = "bold") +
  annotate("text", x = 4, y = 0.25, label = "Alternative:\nReal effect",
           size = 4, color = "#522D80", fontface = "bold") +
  annotate("text", x = 3.2, y = 0.08, label = "Power",
           size = 5, color = "#522D80", fontface = "bold") +
  labs(title = "Power = Probability of Detecting a Real Effect",
       x = "Test Statistic", y = "Density") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

## Power {.smaller}

-   **Power** is the probability of *correctly* detecting a real effect (avoiding a Type II error)
-   Low power = you might miss a program that actually works
-   Power depends on: **(1)** sample size, **(2)** effect size, **(3)** significance level
-   Rule of thumb: **80% power** is the minimum standard. Many policy studies are underpowered!

. . .

**Example:** Your city pilots body cameras with 20 officers. Even if cameras truly reduce complaints by 15%, a sample this small has only \~25% power to detect it — meaning you'd likely conclude "no effect" and kill a program that actually works. With 200 officers, power jumps above 90%.

## What Determines Power? {.smaller}

Power depends on three things, and they trade off against each other:

. . .

-   **Sample size (n):** More data → narrower sampling distribution → easier to distinguish a real effect from noise. This is the lever you have the most control over.
-   **Effect size:** Bigger effects are easier to detect. A program that boosts turnout by 8 points is much easier to pick up than one that boosts it by 0.5 points.
-   **Significance level (α):** If you set a stricter threshold (say α = 0.01 instead of 0.05), you need stronger evidence to reject H₀ — which means you'll miss more real effects. Lower α = fewer false positives, but also lower power.

. . .

That last point is worth sitting with: **there's a direct tradeoff between false positives and false negatives.** Making it harder to cry wolf (Type I) means you'll also miss more real effects (Type II). You can't minimize both at the same time: the only free lunch is collecting more data.

## A Formula for Power {.smaller}

You can see all three ingredients in one formula. For comparing two group means (like treatment vs. control):

$$\text{Power} = \Phi\left(\frac{|\mu_1 - \mu_0|}{\sigma / \sqrt{n}} - z_{1-\alpha/2}\right)$$

. . .

Where:

::: incremental
-   $|\mu_1 - \mu_0|$ is the **effect size** — how different the two groups really are
-   $\sigma / \sqrt{n}$ is the **standard error** — gets smaller as n grows
-   $z_{1-\alpha/2}$ is the **critical value** set by your significance level (1.96 for α = 0.05)
-   $\Phi$ is the cumulative distribution function of the standard normal
    -   It just converts everything to a probability
:::

. . .

Don't worry about memorizing this. The point is to see the mechanics: bigger effect in the numerator, more data shrinking the denominator, and a stricter α pushing the critical value up.

## The Power–Significance Tradeoff, Visually

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

library(tidyverse)
library(patchwork)

x <- seq(-4, 6, length.out = 1000)
null_y <- dnorm(x, mean = 0, sd = 1)
alt_y <- dnorm(x, mean = 2, sd = 1)

df <- tibble(x = x, null = null_y, alt = alt_y)

# Panel 1: α = 0.05 (cutoff at 1.645 for one-sided, but let's use 1.96 for two-sided)
p1 <- ggplot(df, aes(x = x)) +
  geom_line(aes(y = null), linewidth = 1, color = "gray40") +
  geom_line(aes(y = alt), linewidth = 1, color = "#522D80") +
  geom_area(data = filter(df, x > 1.96), aes(y = null), fill = "#F66733", alpha = 0.3) +
  geom_area(data = filter(df, x > 1.96), aes(y = alt), fill = "#522D80", alpha = 0.3) +
  geom_vline(xintercept = 1.96, linetype = "dashed", color = "gray40") +
  annotate("text", x = 2.8, y = 0.22, label = "Power ≈ 50%",
           color = "#522D80", fontface = "bold", size = 4) +
  annotate("text", x = 2.8, y = 0.04, label = "α = 0.05",
           color = "#F66733", fontface = "bold", size = 4) +
  labs(title = "Standard threshold (α = 0.05)",
       x = "Test Statistic", y = "Density") +
  theme_minimal(base_size = 13)

# Panel 2: α = 0.01 (cutoff at 2.576)
p2 <- ggplot(df, aes(x = x)) +
  geom_line(aes(y = null), linewidth = 1, color = "gray40") +
  geom_line(aes(y = alt), linewidth = 1, color = "#522D80") +
  geom_area(data = filter(df, x > 2.576), aes(y = null), fill = "#F66733", alpha = 0.3) +
  geom_area(data = filter(df, x > 2.576), aes(y = alt), fill = "#522D80", alpha = 0.3) +
  geom_vline(xintercept = 2.576, linetype = "dashed", color = "gray40") +
  annotate("text", x = 3.4, y = 0.22, label = "Power ≈ 28%",
           color = "#522D80", fontface = "bold", size = 4) +
  annotate("text", x = 3.4, y = 0.04, label = "α = 0.01",
           color = "#F66733", fontface = "bold", size = 4) +
  labs(title = "Stricter threshold (α = 0.01)",
       x = "Test Statistic", y = "Density") +
  theme_minimal(base_size = 13)

p1 + p2
```

Moving the cutoff to the right (stricter α) shrinks the orange area (fewer false positives) but also shrinks the purple area (less power). Same data, same effect.

## OK, So How Do We Actually Run a Test?

We know what hypotheses are, what errors we can make, and what power means. Now we need to nail down the mechanics:

. . .

1.  **What kind of test?** One-sided or two-sided?
2.  **What's the test statistic?** (the p-value)
3.  **What's the decision rule?** (the significance level α)

. . .

Let's take these one at a time.

## One-Sided vs Two-Sided Tests {.smaller}

::: incremental
-   **Two-sided test:** Does this program have *any* effect? (H₁: τ ≠ 0)
    -   This is what we almost always use in policy research
-   **One-sided test:** Does this program *increase* turnout? (H₁: τ \> 0)
    -   Only use when you have strong prior reason to test in one direction
    -   More powerful (easier to reject) but ignores evidence in the other direction
:::

. . .

**In practice:** Always default to two-sided tests unless you can clearly justify otherwise. Most published policy research uses two-sided tests.

## The p-value: What Does It Actually Mean?

::: callout-tip
## Definition

The **p-value** is the probability of getting a result *at least as extreme* as what we observed, *if the null hypothesis were true*.
:::

. . .

**In plain English:** "If nothing special was happening, how often would I see data this surprising?"

. . .

-   **Small p-value** (like 0.01) → "This would be rare if H₀ were true" → Evidence against H₀
-   **Large p-value** (like 0.45) → "This could easily happen if H₀ were true" → Not much evidence against H₀

## What p-values Are NOT

::: callout-warning
## Common Misconceptions

**A p-value is NOT:**

-   The probability that H₀ is true
-   The probability that H₁ is true
-   A measure of how big or important the effect is
-   A guarantee that your result is meaningful
:::

. . .

**A p-value of 0.03 does NOT mean "there's a 3% chance this is a fluke."**

It means: "If there were no real effect, we'd see data this extreme about 3% of the time."

## The 0.05 Threshold {.smaller}

By convention, researchers often use **α = 0.05** as a cutoff:

-   If p \< 0.05 → "Statistically significant" → Reject H₀
-   If p ≥ 0.05 → "Not statistically significant" → Don't reject H₀

. . .

::: callout-warning
## Important Caveats

-   The 0.05 threshold is **arbitrary**—there's nothing magical about it
-   A p-value of 0.049 isn't fundamentally different from 0.051
-   "Not significant" ≠ "No effect"—you might just need more data
-   Always consider the **practical significance**, not just statistical significance
:::

. . .

In policy settings, practical significance often matters more. A program that raises test scores by 0.01 points might be "statistically significant" with a large enough sample -- but is it worth funding?

## Testing the Voting Experiment {.smaller}

Remember Gerber, Green & Larimer (2008)? We saw the Neighbors mailer group had higher turnout — but is that a real effect, or did we just get a weird sample?

. . .

**Our hypotheses:**

-   **H₀:** The Neighbors mailer has no effect on turnout (turnout is the same in both groups)
-   **H₁:** The Neighbors mailer affects turnout (turnout differs between groups)

. . .

Let's work through this in R.

## Step 1: Load the Data {.smaller}

The actual study had about 99,000 households in the control group and 38,000 in the Neighbors treatment. We have a dataset (`gerber_example.csv`) with one row per household: which group they were in, and whether they voted.

```{r}
#| echo: true

# Load the data
voting_data <- read_csv("gerber_example.csv")

# What does it look like?
voting_data
```

. . .

Each row is a household. `voted` = 1 means they turned out; 0 means they didn't.

## Step 2: Look at the Group Means {.smaller}

Before running any test, let's just look at the data:

```{r}
#| echo: true

# Compute turnout rate by group
voting_data |>
  group_by(group) |>
  summarize(
    n = n(),
    turnout_rate = mean(voted),
    se = sd(voted) / sqrt(n())
  )
```

. . .

OK, so the turnout rates look different. But there's always some random variation in data — maybe we just got unlucky? That's what the t-test will tell us.

## Step 3: Run the Hypothesis Test {.smaller}

```{r}
#| echo: true

# Two-sample t-test: does turnout differ by group?
result <- t.test(voted ~ group, data = voting_data)
result
```

. . .

R gives us a wall of numbers. What do we actually care about?

## Step 4: Reading the Output {.smaller}

**Key numbers to focus on:**

-   **t = `r round(result$statistic, 2)`**: The test statistic — how many standard errors the difference is from zero
-   **p-value = `r format.pval(result$p.value, digits = 3)`**: The probability of seeing a difference this large (or larger) if the mailer had **no effect**
-   **95% CI: \[`r round(result$conf.int[1], 3)`, `r round(result$conf.int[2], 3)`\]**: The range of plausible values for the true difference in turnout
-   **Group means: `r round(result$estimate[2], 3)` (Neighbors) vs `r round(result$estimate[1], 3)` (Control)**: The actual turnout rates in each group

. . .

**Bottom line:** With a sample this big and a gap this large (\~8 percentage points), there's essentially no way this is just sampling noise. We reject H₀.

## We Know It's Significant — But How Big Is It?

OK, so the mailer increased turnout. Great. But if you're a campaign manager or an election administrator, "it worked" isn't enough — you want to know *how much* it worked, and how confident you should be in that number.

. . .

This is what **confidence intervals** are for.

# Confidence Intervals

## Beyond Yes/No: Confidence Intervals

Hypothesis tests give us a binary answer: reject or don't reject.

. . .

But often we want to know **how big** an effect is, not just whether it exists.

. . .

::: callout-tip
## Confidence Interval

A **confidence interval** gives us a range of plausible values for the true population parameter.

A 95% CI means: If we repeated this study many times, about 95% of the intervals we'd calculate would contain the true value.
:::

## Visualizing a Confidence Interval

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

library(tidyverse)

# Example: Survey on policy support
sample_mean <- 0.57
se <- 0.03
ci_lower <- sample_mean - 1.96 * se
ci_upper <- sample_mean + 1.96 * se

tibble(
  estimate = sample_mean,
  lower = ci_lower,
  upper = ci_upper,
  label = "Support for Medicaid Expansion"
) |>
  ggplot(aes(x = label, y = estimate)) +
  geom_point(size = 4, color = "#522D80") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1,
                color = "#522D80", linewidth = 1.2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "#F66733", linewidth = 1) +
  annotate("text", x = 1.3, y = 0.5, label = "50% (majority threshold)", color = "#F66733", size = 5) +
  labs(
    title = "Survey Results: Support for Medicaid Expansion",
    subtitle = "57% support (95% CI: 51% to 63%)",
    x = "",
    y = "Proportion Supporting"
  ) +
  ylim(0.4, 0.7) +
  coord_flip() +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold"))
```

The dot is our estimate; the bar shows the range of values for the true level of support that are consistent with our data.

## Calculating Confidence Intervals

The formula is simple:

$$\text{CI} = \text{Estimate} \pm (\text{Critical Value}) \times (\text{Standard Error})$$

. . .

**For a 95% CI:** Critical value ≈ 1.96 (or about 2)

. . .

**Example:** 57% support Medicaid expansion (sample mean = 0.57), SE = 0.03

-   Lower bound: $0.57 - 1.96 \times 0.03 = 0.51$
-   Upper bound: $0.57 + 1.96 \times 0.03 = 0.63$

**95% CI: \[0.51, 0.63\]**

## Interpreting Confidence Intervals

::: callout-important
## What a 95% CI Means

**Correct:** "If we repeated this study many times, 95% of the resulting intervals would contain the true population value."

**Incorrect:** "There's a 95% probability the true value is in this interval." (The true value is fixed—it either is or isn't in the interval!)
:::

. . .

**Think of it this way:** The interval tells you what effect sizes you can't rule out given your data.

## CIs and Hypothesis Testing Are Connected {.smaller}

There's a useful relationship:

-   If a **95% CI does NOT include zero**, the result is statistically significant at α = 0.05
-   If a **95% CI DOES include zero**, the result is NOT statistically significant

. . .

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 3.5

tibble(
  study = c("Body Cameras → Complaints", "Bias Training → Stops"),
  estimate = c(2.5, 1.2),
  lower = c(0.8, -0.5),
  upper = c(4.2, 2.9)
) |>
  ggplot(aes(x = study, y = estimate)) +
  geom_point(size = 4, color = "#522D80") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1,
                color = "#522D80", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#F66733", linewidth = 1) +
  annotate("text", x = 0.6, y = 0, label = "No effect", color = "#F66733", size = 4) +
  labs(title = "Significant vs. Not Significant", x = "", y = "Effect Size") +
  coord_flip() +
  theme_minimal(base_size = 16)
```

## CIs in R

The `t.test()` function gives you confidence intervals automatically:

```{r}
#| echo: true

# From our voting experiment
result <- t.test(voted ~ group, data = voting_data)

# Extract just the confidence interval
result$conf.int
```

. . .

The true difference in turnout between the Neighbors and Control groups is plausibly somewhere between `r round(result$conf.int[1], 3)` and `r round(result$conf.int[2], 3)`.

## Your Turn: What Would You Tell the City Council?

A city ran a **body camera pilot** with 150 officers. Results:

-   Officers with cameras received **23% fewer complaints** than those without
-   t-test: p = 0.03
-   95% CI for the reduction: \[3%, 43%\]

. . .

**The council asks:** "Should we expand the program citywide?"

. . .

Think about: What does the p-value tell you? What does the CI add? What *doesn't* this analysis tell you?

## Summary: From Samples to Decisions

::: incremental
1.  **Samples work** because of LLN and CLT -- this is why polling and program evaluation are possible

2.  **Hypothesis testing** asks: "Could this result be random noise?" (like the Gerber et al. voting experiment)

3.  **p-values** tell you how rare your result would be under the null -- NOT the probability the null is true

4.  **Confidence intervals** give a range of plausible effect sizes -- much more useful than just "significant" or "not"
:::

## Looking Ahead

So far we can compare two groups (t-tests) and quantify uncertainty (CIs). (More than two groups is an easy extension)

. . .

But what if we want to:

-   Ask whether higher per-pupil **spending** predicts better **graduation rates**?
-   **Control for** the fact that wealthier districts spend more AND have other advantages?
-   Include **multiple predictors** at once?

. . .

**Next: Introduction to Regression**: The workhorse tool for quantitative analysis.