---
title: "Difference-in-Differences for Policy Evaluation"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

## Where We Left Off {.smaller}

::: incremental
-   Last class, we learned **fixed effects**: control for the unit, use only within-unit variation over time

-   We also saw that **two-way fixed effects** (unit + time) has problems when treatment timing varies across units

-   Today we learn the method that builds directly on those ideas: **Difference-in-Differences (DiD)**

-   DiD is probably the **most commonly used** causal inference tool in applied policy research: minimum wage studies, Medicaid expansion, education reforms, environmental regulations
:::

## A Story About Cholera {.smaller}

::: incremental
-   In the 1850s, cholera was killing thousands of people in London. Nobody knew what caused it.

-   The physician John Snow had a theory: cholera spread through contaminated water, not through the air (the prevailing view at the time)

-   Two water companies served overlapping parts of London. In 1852, one of them, the **Lambeth Company**, moved its water intake pipe upstream, away from the sewage. The other, **Southwark & Vauxhall**, kept its pipe in the contaminated water.

-   Households served by the two companies lived in the same neighborhoods, on the same streets. The only difference was where their water came from.
:::

## Snow's Natural Experiment {.smaller}

::: incremental
-   Snow compared cholera death rates in households served by each company, before and after the Lambeth Company moved its pipe

-   Before the move, both groups had similarly high cholera rates

-   After the move, deaths plummeted in Lambeth households but stayed high in Southwark & Vauxhall households

-   The **difference in the changes** between the two groups was the evidence that contaminated water caused cholera
:::

. . .

::: {.callout-note}
This is Difference-in-Differences. Snow didn't compare Lambeth to Southwark & Vauxhall at one point in time (they differed in many ways). He compared how much each group *changed* after the pipe was moved. The part of the change that the control group didn't experience is plausibly caused by the treatment.
:::

## The General Logic {.smaller}

::: incremental
-   The problem with simple before-and-after comparisons is that **time is a confounder**. Things change over time regardless of any policy: the economy grows, demographics shift, diseases wax and wane.

-   If we only look at one group, we can't tell how much of the change is due to the treatment and how much would have happened anyway.

-   The solution: find a **comparison group** that was on a similar trajectory. If the treated group diverges from the control group *after* treatment, that divergence is plausibly caused by the treatment.

-   That's the core logic of DiD: compare the **change** in the treated group to the **change** in a control group.
:::

## Why Should MPAs Care?

::: incremental
-   If you read policy evaluations, you *will* encounter DiD

-   If you do evaluation work yourself, you will likely use it

-   Understanding DiD also means understanding its limits, which will make you a better consumer of policy research
:::

## Today's Roadmap

::: incremental
1.  **The classic 2×2 DiD**: Card & Krueger's minimum wage study
2.  **The parallel trends assumption**: the key to credibility
3.  **DiD in R**: a full worked example
4.  **Event studies**: visualizing effects over time
5.  **Staggered adoption**: when different units get treated at different times
6.  **Modern solutions**: new methods that fix problems with classic TWFE
7.  **Practical advice**: a checklist for credible DiD designs
:::

# Part 1: The Classic 2×2 DiD

## Card & Krueger (1994): Minimum Wage and Employment {.smaller}

::: incremental
-   Snow's cholera study is the foundational example of this kind of reasoning. The most famous modern application is Card & Krueger (1994).

-   In April 1992, **New Jersey** raised its minimum wage from \$4.25 to \$5.05

-   Neighboring **Pennsylvania** did not change its minimum wage

-   Conventional economic theory predicted that a higher minimum wage would reduce employment

-   Card & Krueger surveyed fast-food restaurants in both states, before and after the increase

-   **The question:** Did NJ's minimum wage increase reduce fast-food employment?
:::

## The 2×2 Table

| | Before (Feb 1992) | After (Nov 1992) | Change |
|---|---|---|---|
| **New Jersey** (treated) | 20.44 FTE | 21.03 FTE | +0.59 |
| **Pennsylvania** (control) | 23.33 FTE | 21.17 FTE | -2.16 |

. . .

**DiD Estimate:** (+0.59) − (−2.16) = **+2.76 FTE employees**

. . .

::: {.callout-important}
The minimum wage increase was associated with an **increase** in employment, the opposite of the standard prediction. This is one of the most influential papers in labor economics.
:::

## What Did the DiD Estimate Actually Do? {.smaller}

::: incremental
-   NJ employment went up by 0.59 FTE. But that alone doesn't tell us much, because lots of things were happening at the same time.

-   PA employment went *down* by 2.16 FTE over the same period. PA serves as our benchmark for what would have happened to NJ without the policy.

-   The difference between these two changes (0.59 minus -2.16 = 2.76) removes the common time trend and isolates the effect of the minimum wage increase.

-   This is the same logic as Snow's cholera study: we're not comparing NJ to PA in *levels* (PA had more employees). We're comparing how much each state *changed*.
:::

## The DiD Formula

$$\widehat{\delta}_{DiD} = (\bar{Y}_{treat}^{post} - \bar{Y}_{treat}^{pre}) - (\bar{Y}_{control}^{post} - \bar{Y}_{control}^{pre})$$

::: {.callout-note}
## In English
Take the **change** in the treated group. Subtract the **change** in the control group. What's left is the treatment effect: the part of the change that can't be explained by common time trends.
:::

## The DiD Logic in DAG Terms {.smaller}

```{r}
#| echo: false

library(tidyverse)
library(dagitty)
library(ggdag)

dag <- dagitty("
dag {
  Outcome      [pos=\"2,0\"]
  Time         [pos=\"1,-1\"]
  Treatment    [pos=\"0,0\"]
  Group        [pos=\"1,1\"]

  Time -> Treatment
  Time -> Outcome
  Treatment -> Outcome
  Group -> Treatment
  Group -> Outcome
}
")
ggdag(dag, text = FALSE, use_labels = "name") +
  coord_cartesian(clip = "off") +
  theme_dag() +
  theme(plot.margin = margin(t = 20, r = 10, b = 10, l = 10), legend.position = "none")
```

::: incremental
-   There are two backdoor paths: through **Time** and through **Group**
-   Differencing over time within each group closes the Time path (this is the fixed effects idea from last class)
-   Differencing *between* groups closes the Group path
-   DiD closes both at once
:::

# Part 2: The 2×2 DiD Estimator

## DiD as a Regression {.smaller}

We can express the DiD calculation as a regression. This should look familiar from OLS:

$$Y_{it} = \beta_{0} + \beta_{1}\text{Group}_{i} + \beta_{2}\text{Post}_{t} + \beta_{3}(\text{Group}_{i} \times \text{Post}_{t}) + \epsilon_{it}$$

::: incremental
-   $\beta_1$ captures the baseline difference between groups (NJ vs PA before the policy)
-   $\beta_2$ captures the common time trend (what changed for everyone)
-   $\beta_3$ is the **DiD estimate**: the extra change in the treated group beyond what the control group experienced
:::

## The Connection to Fixed Effects {.smaller}

::: incremental
-   That interaction regression is equivalent to a **two-way fixed effects** model:

$$Y_{it} = \alpha_{g} + \alpha_{t} + \beta \cdot \text{Treated}_{it} + \epsilon_{it}$$

-   $\alpha_g$ is a group fixed effect (absorbs level differences between NJ and PA)
-   $\alpha_t$ is a time fixed effect (absorbs common trends)
-   $\beta$ is the same DiD estimate as $\beta_3$ above

-   In `fixest`, this is the same syntax you learned last class, with treatment as the key variable:
:::

```r
feols(fte ~ treated:post | state + period, data = ck_data)
```

## Simulating Card & Krueger in R {.smaller}

```{r}
#| echo: true

library(tidyverse)
library(fixest)
set.seed(30317)

n_restaurants <- 400  # 200 per state

ck_data <- tibble(
  restaurant = 1:n_restaurants,
  state = rep(c("NJ", "PA"), each = n_restaurants / 2),
  # PA starts with slightly more employees on average
  base_fte = ifelse(state == "PA",
                    rnorm(n_restaurants / 2, 23.3, 4),
                    rnorm(n_restaurants / 2, 20.4, 4))
) %>%
  # Each restaurant appears twice: before and after
  crossing(period = c("before", "after")) %>%
  mutate(
    post    = ifelse(period == "after", 1, 0),
    treated = ifelse(state == "NJ", 1, 0),
    # Build the outcome from components:
    fte = base_fte +
      ifelse(post == 1, -2.2, 0) +                # Common time trend (decline)
      ifelse(post == 1 & treated == 1, 2.76, 0) +  # True treatment effect
      rnorm(n(), 0, 3)                              # Random noise
  )
```

We built in a true treatment effect of **+2.76 FTE**. Can our regression recover it?

## Running the DiD Regression {.smaller}

```{r}
#| echo: true

# TWFE with fixest (same syntax as last class)
did_twfe <- feols(fte ~ treated:post | state + period, data = ck_data)
summary(did_twfe)
```

. . .

The `treated:post` coefficient is close to 2.76, the value we built into the simulation. The state and period fixed effects are absorbed behind the scenes, just like the country fixed effects in the gapminder example.

## The Interaction Form Gives the Same Answer {.smaller}

```{r}
#| echo: true

# Interaction form with lm() -- equivalent, but shows all coefficients
did_interaction <- lm(fte ~ treated * post, data = ck_data)
summary(did_interaction)$coefficients
```

::: {.callout-note}
The `treated:post` coefficient is the DiD estimate in both models. The interaction form also shows you $\beta_1$ (group difference) and $\beta_2$ (time trend), which `feols()` absorbs into the fixed effects. Both approaches give the same answer.
:::

# Part 3: The Parallel Trends Assumption

## The Key Assumption {.smaller}

::: incremental
-   DiD requires one critical assumption: **without the treatment**, the treated group would have followed the same trend as the control group

-   This is called the **parallel trends assumption**

-   In Card & Krueger, it means: if NJ had *not* raised its minimum wage, NJ's fast-food employment would have changed at the same rate as PA's

-   Notice this is an assumption about a **counterfactual** that we can never observe. We can check whether trends were parallel *before* the treatment, but the assumption is about what would have happened *after*.
:::

## Levels vs. Trends {.smaller}

::: {.callout-important}
## A Common Confusion

Parallel trends does **not** require that the groups are at the same level. PA had more FTE employees than NJ before the policy. That's fine. What matters is that PA and NJ were *changing at the same rate*. DiD compares changes, not levels. This is the same within-unit logic from our fixed effects lecture.
:::

::: incremental
-   Two cities can have very different crime rates but still have parallel trends if crime is rising (or falling) at the same rate in both

-   If they are changing at different rates before the policy, any post-treatment difference could just be a continuation of that pre-existing divergence
:::

## Good vs. Bad Parallel Trends {.smaller}

```{r}
#| echo: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

df <- tibble(
  time = rep(1:10, times = 4),
  group = rep(c("Treatment", "Control"), each = 10, times = 2),
  panel = rep(c("Parallel Prior Trends", "Converging Prior Trends"), each = 20),
  outcome = c(
    c(2, 2, 3, 4, 5, 7, 9, 10, 9.8, 9.5),
    c(1, 1, 2, 3, 3.5, 4.5, 5.5, 6, 5.5, 6),
    c(4, 3.5, 4, 5, 6, 6.5, 7.2, 9, 10, 10.5),
    c(1, 1.2, 2.2, 3.2, 4.8, 6.5, 8.5, 9.5, 9.7, 10)
  )
)

make_plot <- function(data, panel_title) {
  ggplot(data, aes(x = time, y = outcome, color = group)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    geom_vline(xintercept = 6, linetype = "dashed") +
    annotate("rect", xmin = 6, xmax = 10.5, ymin = -Inf, ymax = Inf, alpha = 0.1) +
    annotate("text", x = 6.3, y = max(data$outcome) * 0.2, label = "Treatment\nPeriod", hjust = 0) +
    scale_color_manual(values = c("Treatment" = "black", "Control" = "gray40")) +
    labs(title = panel_title, x = "Time", y = "Outcome", color = "Group") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom")
}

p1 <- make_plot(df %>% filter(panel == "Parallel Prior Trends"), "(a) Parallel Prior Trends")
p2 <- make_plot(df %>% filter(panel == "Converging Prior Trends"), "(b) Converging Prior Trends")

p1 + p2
```

Panel (a): pre-trends are parallel, so the post-treatment gap is plausibly causal. Panel (b): the control group was already converging before treatment, so the "effect" may just be a continuation of that pre-existing trend.

## When Is Parallel Trends Plausible? {.smaller}

::: incremental
-   **More plausible:**
    -   Two neighboring states with similar economic structures (this is why Card & Krueger chose NJ and PA, not NJ and Alaska)
    -   Units selected for treatment based on a fixed characteristic, not a trend
    -   Short time horizons where trends are less likely to diverge

-   **Less plausible:**
    -   Comparing a booming city to a declining one
    -   Units that self-selected into treatment based on anticipated trends (e.g., states that expanded Medicaid because their health outcomes were already improving)
    -   Very long time horizons with structural changes
:::

## Testing Parallel Trends {.smaller}

::: incremental
-   We can't prove parallel trends holds, but we can look for evidence against it

-   **Placebo tests in pre-treatment periods:** Run DiD as if treatment happened earlier. If you find "effects" before the real treatment, something is wrong.

-   **Event study plots** (which we'll build later) are the standard way to do this. They show the treatment-control difference at each time period. Before treatment, those differences should hover around zero.

-   Finding null pre-treatment effects doesn't *prove* the assumption, but it's reassuring. Finding significant pre-treatment effects is a serious red flag.
:::

## Anticipation Effects {.smaller}

::: incremental
-   What if people change behavior **before** the policy takes effect?

    -   Workers might change jobs before a minimum wage increase

    -   Firms might hire ahead of a regulation deadline

-   Anticipation violates the assumption that treatment "turns on" at a specific time. It can also make pre-trends look non-parallel even if the design is otherwise sound.

-   **Solution:** Backdate the treatment period or exclude the periods just before treatment
:::

# Part 4: DiD in R - A Worked Example

## A New Policy Question {.smaller}

::: incremental
-   Suppose a state expands its **Earned Income Tax Credit (EITC)**, which gives tax refunds to low-income working families

-   The goal is to encourage labor force participation. Did it work?

-   We have data on 30 counties in the treatment state and 30 counties in a neighboring control state, observed over 12 years. The expansion happens in year 7.

-   This is a larger dataset than Card & Krueger, which lets us see the pre-trends more clearly and will set us up for event studies in the next section.
:::

## Simulating the EITC Data {.smaller}

```{r}
#| echo: true

library(tidyverse)
library(fixest)
set.seed(30317)

treatment_year <- 7

# 60 counties: 30 in treatment state, 30 in control state
county_info <- tibble(
  county = 1:60,
  treated_state = rep(c(1, 0), each = 30),
  state = ifelse(treated_state == 1, "Treatment", "Control"),
  # Each county has its own baseline LFP
  county_fe = rnorm(60, 0, 2)
)

eitc_data <- county_info %>%
  crossing(year = 1:12) %>%
  mutate(
    post = ifelse(year >= treatment_year, 1, 0),
    # Both states trend upward; treatment state starts lower
    base = ifelse(treated_state == 1, 55, 60) + county_fe + 0.8 * year,
    # True treatment effect: +3 percentage points
    lfp = base +
      ifelse(treated_state == 1 & post == 1, 3, 0) +
      rnorm(n(), 0, 1.5)
  )
```

## Visualizing the Data {.smaller}

```{r}
#| echo: true

# Average across counties for a clean plot
eitc_means <- eitc_data %>%
  group_by(state, year) %>%
  summarise(avg_lfp = mean(lfp), .groups = "drop")

ggplot(eitc_means, aes(x = year, y = avg_lfp, color = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = treatment_year - 0.5, linetype = "dashed", color = "gray40") +
  annotate("text", x = treatment_year + 0.2, y = min(eitc_means$avg_lfp) + 1,
           label = "EITC Expansion", hjust = 0, size = 4) +
  labs(title = "Average Labor Force Participation by State",
       x = "Year", y = "Labor Force Participation (%)",
       color = "State") +
  theme_minimal(base_size = 14)
```

Pre-treatment trends look parallel, which is a good sign for our DiD design. The treatment state starts at a lower level, but remember: parallel trends is about *rates of change*, not levels.

## Running the DiD {.smaller}

```{r}
#| echo: true

# TWFE regression with county and year fixed effects
eitc_did <- feols(lfp ~ treated_state:post | county + year, data = eitc_data)
summary(eitc_did)
```

. . .

::: {.callout-note}
## Interpretation
The EITC expansion is associated with approximately a 3 percentage-point increase in labor force participation. The county fixed effects absorb all the stable differences between counties; the year fixed effects absorb the common upward trend. What's left is the treatment effect.
:::

# Part 5: Event Studies

## What If the Effect Isn't Constant? {.smaller}

::: incremental
-   Our EITC simulation assumed the treatment effect was an immediate, constant +3 points. But that's often unrealistic.

-   A job training program might not show results for a year

-   A tax credit might take time as people learn about it and adjust their behavior

-   An **event study** lets us estimate the DiD effect **separately for each time period**, so we can see how the effect evolves

-   It also gives us a built-in test of parallel trends: the pre-treatment effects should be close to zero
:::

## How an Event Study Works {.smaller}

::: incremental
-   Instead of one "post" dummy, we create a dummy for each time period relative to treatment

-   The period just before treatment (event time = -1) is the **reference category**, so all effects are measured relative to that point

-   **Pre-treatment coefficients** should hover around zero if parallel trends holds. This is the placebo test we discussed earlier.

-   **Post-treatment coefficients** show how the effect unfolds: does it appear immediately? Grow over time? Fade?
:::

## Building an Event Study {.smaller}

We'll simulate a new dataset where the EITC effect **grows over time** as more people learn about and respond to the program.

```{r}
#| echo: true

set.seed(30317)

# 60 units (30 treated, 30 control), 12 years, treatment in year 7
unit_fes <- tibble(unit = 1:60, unit_fe = rnorm(60, 0, 3))

event_data <- expand_grid(unit = 1:60, year = 1:12) %>%
  left_join(unit_fes, by = "unit") %>%
  mutate(
    treated_unit = ifelse(unit <= 30, 1, 0),
    event_time = year - 7,
    # Dynamic treatment effect: grows over time after treatment
    te = case_when(
      treated_unit == 0 ~ 0,  # Control units: no effect
      event_time < 0    ~ 0,  # Before treatment: no effect
      event_time == 0   ~ 1.5,
      event_time == 1   ~ 2.5,
      event_time == 2   ~ 3.2,
      event_time == 3   ~ 3.5,
      TRUE              ~ 4.0  # Levels off
    ),
    outcome = 50 + unit_fe + 0.5 * year + te + rnorm(n(), 0, 2)
  )
```

## The Event Study Regression {.smaller}

```{r}
#| echo: true

# i() creates a dummy for each event_time period
# ref = -1 sets the period before treatment as the reference
es_model <- feols(
  outcome ~ i(event_time, treated_unit, ref = -1) | unit + year,
  data = event_data
)

# fixest has a built-in event study plot
iplot(es_model,
      xlab = "Periods Relative to Treatment",
      ylab = "Estimated Effect",
      main = "Event Study: Effect of EITC Expansion Over Time")
```

## Reading the Event Study Plot {.smaller}

::: incremental
-   **Pre-treatment coefficients** (left of 0): they hover around zero with confidence intervals crossing zero. This supports parallel trends.

-   **The jump at time 0**: the immediate treatment effect (about 1.5 points in our simulation)

-   **Post-treatment growth**: the effect increases over time as more people respond to the program, then levels off around 4 points

-   **Confidence intervals**: if they include zero after treatment, we can't rule out no effect. Here the post-treatment intervals are clearly above zero.
:::

. . .

::: {.callout-tip}
## For Your Reports
Event study plots are the **standard way** to present DiD results in policy research. They show both the treatment effect and the parallel trends evidence in a single figure. If you do a DiD analysis, include one.
:::

# Part 6: The Staggered Adoption Problem

## What If Treatment Happens at Different Times? {.smaller}

::: incremental
-   So far, we've had a clean setup: one group treated, one group not, at a single point in time

-   In practice, policies are often adopted **at different times** by different units

    -   States expanded Medicaid in 2014, 2015, 2016, or not at all

    -   Cities adopted minimum wage increases in different years

    -   Schools implemented new curricula on different schedules

-   This is called **staggered adoption**, and it creates serious problems for classic TWFE
:::

## Visualizing Staggered Treatment {.smaller}

```{r}
#| echo: false

library(tidyverse)

stagger_viz <- tibble(
  State = c("Alabama", "Colorado", "Delaware", "Florida", "Georgia"),
  treat_year = c(2015, 2016, 2018, NA, NA)
) %>%
  crossing(Year = 2012:2020) %>%
  mutate(
    status = case_when(
      is.na(treat_year) ~ "Never Treated",
      Year < treat_year ~ "Not Yet Treated",
      Year >= treat_year ~ "Treated"
    ),
    status = factor(status, levels = c("Treated", "Not Yet Treated", "Never Treated"))
  )

ggplot(stagger_viz, aes(x = Year, y = State, fill = status)) +
  geom_tile(color = "white", linewidth = 1) +
  scale_fill_manual(values = c("Treated" = "#2C7BB6", "Not Yet Treated" = "#FFFFBF",
                                "Never Treated" = "#D7191C")) +
  labs(title = "Staggered Treatment Adoption",
       fill = "Treatment Status") +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())
```

Different states get treated at different times. Who serves as the "control" for whom?

## Why Classic TWFE Breaks {.smaller}

::: incremental
-   Classic TWFE estimates a **single treatment coefficient** from a weighted average of *all possible 2×2 comparisons*

-   The problem: some of those comparisons are **"forbidden"**

    -   TWFE compares **newly-treated units** to **already-treated units**

    -   An already-treated unit is being used as a "control," but it's already affected by treatment!

-   If treatment effects **change over time** (grow, fade, or differ across groups), these forbidden comparisons contaminate the estimate

-   The weights can even be **negative**, meaning the estimate can show a negative effect when the true effect is positive for everyone
:::

## An Intuitive Example {.smaller}

::: {.callout-warning}
## The Negative Weights Problem

Imagine a job training program that helps everyone, but the effect grows over time. Early adopters have large effects by now (they've had the program for years). Late adopters have small effects (they just started).

TWFE compares late adopters (small positive effect) to early adopters (large positive effect). The "change" in the early-adopter "controls" is actually driven by their growing treatment effect, making it look like the new treatment *reduced* outcomes. The estimated effect can flip negative, even though **everyone benefited**.

This is exactly the kind of dynamic effect we saw in our event study. If we had staggered adoption in that setting, TWFE would give us the wrong answer.
:::

## When Is TWFE Fine? {.smaller}

::: incremental
-   **TWFE is fine if:**
    -   Treatment effects are constant over time and across groups
    -   All units are treated at the same time (the classic 2×2 case)
    -   You have a "never-treated" group and only compare to them

-   **TWFE is problematic if:**
    -   Treatment effects build up, fade, or vary across cohorts
    -   You have staggered adoption with no never-treated group
    -   You suspect dynamic or heterogeneous effects

-   In practice, constant effects are a strong assumption. Most policies take time to work, which means effects change over time. When in doubt, use the newer methods we'll see next.
:::

# Part 7: Modern Solutions

## The Good News {.smaller}

Econometricians have developed several methods that fix the staggered DiD problem. The two most widely used:

::: incremental
-   **Callaway and Sant'Anna (2021):** Estimate separate treatment effects for each treatment cohort at each time period, then aggregate. R package: `did`.

-   **Sun and Abraham (2021):** An interaction-weighted event-study estimator that avoids the contamination. Available in `fixest` via `sunab()`.

-   Both approaches share a key idea: **only make "clean" comparisons**. Treated units should only be compared to not-yet-treated or never-treated units, never to already-treated units.
:::

## Simulating Staggered Adoption {.smaller}

```{r}
#| echo: true

set.seed(30317)

# 40 units in 4 groups:
# 10 treated starting year 5, 10 in year 7, 10 in year 9, 10 never
stag_unit_fes <- tibble(unit = 1:40, unit_fe = rnorm(40, 0, 2))

stag_data <- expand_grid(unit = 1:40, year = 1:12) %>%
  left_join(stag_unit_fes, by = "unit") %>%
  mutate(
    cohort = case_when(
      unit <= 10 ~ 5,   # Early adopters
      unit <= 20 ~ 7,   # Middle adopters
      unit <= 30 ~ 9,   # Late adopters
      TRUE ~ Inf         # Never treated
    ),
    treated = ifelse(year >= cohort, 1, 0),
    years_treated = pmax(year - cohort, 0),
    # Dynamic effect: grows with time since treatment (1 + 0.5 per year)
    true_te = ifelse(is.finite(cohort) & treated == 1,
                     1 + 0.5 * years_treated, 0),
    outcome = 50 + unit_fe + 0.3 * year + true_te + rnorm(n(), 0, 1.5)
  )
```

The effect grows over time since treatment. This is exactly the scenario where TWFE breaks.

## Running Both Models {.smaller}

```{r}
#| echo: true

library(fixest)

# Classic TWFE: one coefficient for all treatment variation
twfe_model <- feols(outcome ~ treated | unit + year, data = stag_data)

# Sun & Abraham: only uses clean comparisons
# sunab() needs a finite number for never-treated units
stag_data_sa <- stag_data %>%
  mutate(cohort_sa = ifelse(is.infinite(cohort), 1000, cohort))

sa_model <- feols(outcome ~ sunab(cohort_sa, year) | unit + year,
                  data = stag_data_sa)
```

`sunab()` does the heavy lifting: it figures out which comparisons are clean and estimates the effect for each time period relative to treatment, just like our event study earlier.

## TWFE: A Single (Misleading) Coefficient {.smaller}

```{r}
#| echo: true

summary(twfe_model)
```

TWFE gives us one number for the treatment effect. But we built in an effect that grows over time, so a single number can't capture the full picture. Worse, the forbidden comparisons may be pulling this estimate in the wrong direction.

## Sun & Abraham: The Full Event Study {.smaller}

```{r}
#| echo: true

iplot(sa_model,
      xlab = "Periods Relative to Treatment",
      ylab = "Estimated Effect",
      main = "Sun & Abraham Event Study (Staggered DiD)")
```

The Sun & Abraham estimator correctly shows pre-treatment coefficients near zero and a growing post-treatment effect. This is what the true data-generating process looks like.

## The Key Takeaway {.smaller}

::: {.callout-important}
## Reading DiD Papers
If you encounter a DiD paper with **staggered treatment** that was published before roughly 2020, the estimates might be unreliable, especially if treatment effects are likely to grow or change over time. Papers using modern methods (Callaway & Sant'Anna, Sun & Abraham, or similar) have addressed this problem. When evaluating research, ask: "Is treatment staggered? If so, did they use a modern estimator?"
:::

# Part 8: Practical Advice

## A Checklist for Credible DiD Designs

::: incremental
1.  Is there a **clear treatment** and a **clear control group**?
2.  Is the **treatment timing plausibly exogenous**, not driven by anticipated outcomes?
3.  Do **pre-treatment trends look parallel**? (Plot them!)
4.  Are there **anticipation effects** that could blur the treatment timing?
5.  Is treatment **staggered**? If so, use modern methods.
6.  How **sensitive** are results to the choice of control group?
:::

## What DiD Estimates: The ATT {.smaller}

::: incremental
-   One important detail: DiD does not estimate the average treatment effect for everyone (the ATE). It estimates the **Average Treatment Effect on the Treated (ATT)**.

-   In Card & Krueger, the estimate tells us what the minimum wage increase did **in New Jersey**. It does not tell us what would happen if Pennsylvania also raised its wage, because PA might be different in ways that matter.

-   In a randomized experiment, treatment and control groups are balanced, so the ATE and ATT are the same. In DiD, groups are not randomly assigned, so we can only speak to the effect on the units that were actually treated.

-   This is usually fine for policy evaluation. Legislators want to know: "Did *our* policy work?" That's an ATT question.
:::

## Common Pitfalls

::: incremental
-   **Confusing levels with trends:** Groups can be at very different levels; what matters is that they *change at the same rate* pre-treatment

-   **Cherry-picking the control group** to manufacture parallel trends

-   **Not clustering standard errors** at the level of treatment assignment

-   **Using classic TWFE with staggered adoption** when effects are dynamic

-   **Ignoring anticipation effects** that shift behavior before the treatment takes effect
:::

## Readings {.smaller}

::: incremental
-   **Cunningham, *Causal Inference: The Mixtape***: Excellent chapter on DiD with real-world examples and R code. Freely available online.

-   **Huntington-Klein, *The Effect***: Very intuitive treatment of DiD and event studies. Also freely available.

-   **Callaway & Sant'Anna (2021)**: The key paper on staggered DiD. Technical, but the introduction is accessible.

-   **Goodman-Bacon (2021)**: "Difference-in-Differences with Variation in Treatment Timing." The paper that showed why TWFE breaks with staggered adoption.
:::

# Wrapping Up

## What We Covered Today {.smaller}

::: incremental
1.  **The DiD idea**: use a control group to remove common time trends
2.  **Card & Krueger**: the classic 2×2 minimum wage study
3.  **The parallel trends assumption**: the foundation of DiD credibility
4.  **Event studies**: the standard way to present DiD results and test assumptions
5.  **Staggered adoption**: why classic TWFE breaks and how to spot the problem
6.  **Modern solutions**: Sun & Abraham and Callaway-Sant'Anna fix the staggered problem
7.  **Practical advice**: a checklist for credible designs
:::

## What's Next

::: incremental
-   **Regression Discontinuity Designs**: exploiting arbitrary cutoffs for causal inference

-   **Instrumental Variables**: finding exogenous sources of variation
:::

## For Your Capstone

::: {.callout-tip}
## Think About This Now
Is there a **policy change** in your area of interest with a natural control group? A state that changed a law while a neighbor didn't? A program that was rolled out in some districts before others? If so, DiD might be the right tool, and now you know the modern best practices for using it.
:::
