---
title: "Regression Discontinuity Designs"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

# F**uzzy RDD is a well-behaved form of IV:**

::: incremental
-   The first stage is **visible in a plot** — you can literally see whether the cutoff moves treatment
-   Assignment at the cutoff is nearly random *by construction*
-   "Only through" is easier to argue about a narrow threshold than for an instrument coming from somewhere else in the world
:::

. . .

*We won't cover standalone IV in this course, it's a whole separate methodology, and more controversial than its fuzzy-RDD cousin.*

## The Catch: Why Fuzzy Estimates Deserve Scrutiny {.smaller}

Fuzzy RDD requires a strong assumption — and it can mechanically amplifywhat it reports.

::: incremental
-   **The "only through" part is a strong assumption.** For Maimonides, is class size really the *only* thing that changes at enrollment = 41? What about teacher assignments, classroom resources, peer composition, or parents sorting across schools? If any of these also moves the outcome, we could be over-estimating the effect.

-   **Scaling amplifies noise.** Dividing a small, imprecise outcome jump by a small, imprecise treatment jump produces a bigger estimate *and* a wider confidence interval. A modest reduced-form result can become a dramatic-sounding fuzzy estimate
:::

. . .

::: callout-important
## Bottom Line

A sharp RDD result is easier to defend than a fuzzy one. When you encounter a fuzzy RDD in a paper or report, ask: *what else might be changing at that cutoff?* and *how strong is the first stage?* Be more skeptical than you would be with a sharp design.
:::

## Fuzzy RDD in `rdrobust` {.smaller}

```{r}
#| echo: true
#| eval: false

# Fuzzy RDD: pass the actual treatment indicator to `fuzzy`.
# rdrobust handles the two-jump scaling for you — but the scaling
# is only as honest as the "only through" assumption on the prior slide.
rd_fuzzy <- rdrobust(y = outcome, x = running_var, c = cutoff,
                     fuzzy = treatment_indicator)
summary(rd_fuzzy)
```

. . .

::: callout-tip
## Sharp or Fuzzy?

If the cutoff **perfectly determines** treatment → Sharp RDD. If the cutoff **shifts the probability** of treatment → Fuzzy RDD. When in doubt, plot the treatment probability against the running variable. If it's a step function, it's sharp.
:::

# Part 6: Validity Checks

## Manipulation Testing: The McCrary Density Test {.smaller}

::: incremental
-   If people can **manipulate** the running variable, we'd see **bunching** just above (or below) the cutoff

-   The McCrary (2008) density test checks: is the distribution of the running variable smooth through the cutoff?

-   A "pile-up" on one side suggests manipulation — and invalidates the design
:::

## What Manipulation Looks Like {.smaller}

```{r}
#| echo: true

set.seed(30317)

# Build a side-by-side comparison: clean vs. manipulated scores.
n_bunch       <- 80   # ~16% of students "cheated" into the 80-82 range
bunched_score <- c(
  runif(500 - n_bunch, 60, 100),  # baseline uniform
  runif(n_bunch,        80, 82)   # concentrated spike just above cutoff
)

df_density <- bind_rows(
  tibble(score = df$score,      condition = "Clean: no manipulation"),
  tibble(score = bunched_score, condition = "Manipulated: bunching above cutoff")
)

ggplot(df_density, aes(x = score)) +
  geom_histogram(bins = 40, fill = "gray70", color = "white") +
  geom_vline(xintercept = cutoff, linetype = "dashed",
             color = "red", linewidth = 1) +
  facet_wrap(~ condition) +
  labs(title = NULL,
       x = "Test Score", y = "Count") +
  theme_minimal(base_size = 13)
```

**Left:** smooth density through the cutoff — what we want. **Right:** a clear pile-up just above 80 — a red flag that students may have manipulated their scores to qualify.

## Formal Manipulation Test {.smaller}

```{r}
#| echo: true

# The rddensity package provides a formal test
library(rddensity)

density_test <- rddensity(X = df$score, c = cutoff)
summary(density_test)
```

. . .

A non-significant p-value means we **fail to reject** the null of no manipulation — which is what we want. If significant, the RDD may be compromised.

## Placebo Tests {.smaller}

::: incremental
-   **Placebo cutoffs:** Run the RDD at *fake* cutoff values where there's no treatment. If you find "effects," something is wrong.

-   **Covariate balance:** Check whether *pre-determined characteristics* (demographics, prior outcomes) show a jump at the cutoff. They shouldn't.

-   **Donut hole test:** Remove observations *very* close to the cutoff and re-estimate. If the result changes dramatically, those boundary observations may be problematic.
:::

## Placebo Cutoffs in Action {.smaller}

```{r}
#| echo: true

# Run rdrobust at real and fake cutoffs; only the real one should show an effect
placebo_cutoffs <- c(70, 75, 80, 85, 90)   # 80 is the real cutoff

placebo_results <- map_dfr(placebo_cutoffs, function(c_val) {
  rd <- rdrobust(y = df$completion, x = df$score, c = c_val)
  tibble(cutoff   = c_val,
         is_real  = c_val == 80,
         estimate = rd$coef[1],
         ci_lower = rd$ci[1, 1],
         ci_upper = rd$ci[1, 2])
})

ggplot(placebo_results, aes(x = cutoff, y = estimate, color = is_real)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c(`TRUE` = "#e31a1c", `FALSE` = "gray40"),
                     labels = c(`TRUE` = "Real cutoff", `FALSE` = "Placebo")) +
  labs(title = "RDD Estimates at Real vs. Placebo Cutoffs",
       x = "Cutoff Value (Test Score)",
       y = "RDD Estimate (completion rate)", color = NULL) +
  theme_minimal(base_size = 13)
```

Only the cutoff at 80 should show a real effect. Placebo cutoffs should be indistinguishable from zero — that's what we see.

## The Donut Hole Test {.smaller}

```{r}
#| echo: true

donut_values <- c(0, 0.5, 1, 2)   # donut radius in test-score points

donut_results <- map_dfr(donut_values, function(d) {
  keep <- abs(df$score - cutoff) >= d
  rd <- rdrobust(y = df$completion[keep], x = df$score[keep], c = cutoff)
  tibble(donut    = d,
         estimate = rd$coef[1],
         ci_lower = rd$ci[1, 1],
         ci_upper = rd$ci[1, 2])
})

ggplot(donut_results, aes(x = donut, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.08) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Donut Hole Test: Stability When Near-Cutoff Points Are Excluded",
       x = "Donut Radius (test-score points excluded on each side)",
       y = "RDD Estimate (completion rate)") +
  theme_minimal(base_size = 13)
```

If the estimate is **stable** as we exclude observations closer to the cutoff, that's reassuring. Wild swings would suggest those boundary observations are driving — or contaminating — the result.

## Covariate Balance Check {.smaller}

```{r}
#| echo: true

# Simulate a pre-determined covariate (family income, in $000s)
# — should NOT jump at the cutoff if the RDD is credible
set.seed(30317)
df$family_income <- 30 + 0.4 * df$score + rnorm(nrow(df), 0, 5)

ggplot(df, aes(x = score, y = family_income, color = group)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_vline(xintercept = cutoff, linetype = "dashed") +
  labs(title = "Covariate Balance: Family Income by Test Score",
       subtitle = "No jump at the cutoff = good news for the RDD",
       x = "Test Score", y = "Family Income ($000s)") +
  scale_color_manual(values = c("No Scholarship" = "#1f78b4",
                                "Scholarship"    = "#e31a1c")) +
  theme_minimal(base_size = 14)
```

Covariates should pass smoothly through the cutoff. A jump here would be a red flag.

## Sensitivity to Bandwidth {.smaller}

```{r}
#| echo: true

# Check if the estimate is stable across bandwidth choices
library(rdrobust)

bandwidths <- c(2, 3, 5, 7, 10, 15)   # in test-score points

sensitivity <- map_dfr(bandwidths, function(h) {
  rd <- rdrobust(y = df$completion, x = df$score, c = cutoff, h = h)
  tibble(bandwidth = h,
         estimate  = rd$coef[1],
         ci_lower  = rd$ci[1, 1],
         ci_upper  = rd$ci[1, 2])
})

ggplot(sensitivity, aes(x = bandwidth, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Sensitivity to Bandwidth Choice",
       x = "Bandwidth (test-score points)",
       y = "RD Estimate (completion rate)") +
  theme_minimal(base_size = 14)
```

If the estimate is **stable** across bandwidths, that's reassuring. Wild swings suggest fragility.

# Part 7: Practical Advice

## Checklist + Common Pitfalls {.smaller}

**Before you trust an RDD — yours or someone else's:**

::: incremental
1.  Running variable is **precisely measured** and hard to manipulate
2.  **Density** is smooth through the cutoff (no bunching)
3.  **Covariates balance** at the cutoff
4.  Estimate is **robust** to different bandwidths
5.  **Placebo cutoffs** show null effects
6.  If **fuzzy**, first stage is strong
:::

. . .

**Pitfalls to avoid:**

::: incremental
-   Too-wide bandwidth — people far from the cutoff aren't comparable
-   High-order polynomials — can manufacture fake jumps (Gelman & Imbens 2019)
-   Forgetting **LATE** — your estimate applies at the cutoff, not everywhere
-   Not showing the plot — RDD is visual; the picture *is* the evidence
-   Assuming sharp when it's fuzzy — biases the estimate toward zero
:::

## Readings {.smaller}

-   **Huntington-Klein, *The Effect*, Ch. 20** — intuitive treatment with great examples (free online)
-   **Cunningham, *Causal Inference: The Mixtape*, Ch. 6** — detailed chapter with R code (free online)
-   **Cattaneo, Idrobo, and Titiunik (2020)** — *A Practical Introduction to Regression Discontinuity Designs*. The definitive applied guide
-   **Lee and Lemieux (2010)** — "Regression Discontinuity Designs in Economics." The classic review

# Wrapping Up

## What We Covered Today {.smaller}

::: incremental
1.  **One last DiD walkthrough** — Medicaid expansion and mortality
2.  **The RDD idea** — cutoffs as local randomization
3.  **The graphical recipe** — plot, fit, bandwidth, estimate the jump
4.  **Estimation** — from OLS to `rdrobust` with optimal bandwidth selection
5.  **Policy examples** — shingles/dementia, Medicare at 65
6.  **Sharp vs. fuzzy** — with Maimonides' Rule as the fuzzy example
7.  **Validity checks** — manipulation, placebo cutoffs, donut holes, bandwidth sensitivity
:::

# Doing Quant Work as an MPA

## Four Habits That Separate Good Analysis from Statistical Theater {.smaller}

::: incremental
1.  **Start with the question, not the method.** A simple t-test on the right comparison beats a fancy model on the wrong one. Design follows the question — not the other way around

2.  **You'll mostly be a consumer, not a producer.** Most of your career you'll read, commission, and push back on analyses. Learn to ask: *"What's the comparison? How do you know they're comparable? Who does this estimate apply to?"*

3.  **Show the picture.** Every method we covered has a visual signature — parallel pre-trends for DiD, a jump at the cutoff for RDD, balanced baselines for experiments. If the plot doesn't tell the story, be suspicious of the coefficient that does

4.  **Be honest about the limits.** Every estimate has assumptions. Every result applies to a specific group. Saying "we can't tell from this data" is often more useful — and more credible — than overselling a small effect
:::

## What This Course Was Really About {.smaller}

::: callout-tip
## The Real Lesson

We spent the semester on methods — OLS, fixed effects, DiD, RDD, experiments. The real lesson is **mental discipline**: being precise about *what question you're answering*, *what assumptions you're making*, and *what your result does and doesn't mean*.

You don't need to be the person who runs every analysis. You do need to be the person who knows **what to ask, what to trust, and what to reject**. That's how quantitative work actually shapes policy — and it's what makes you valuable.
:::

## For Your Capstone {.smaller}

::: callout-tip
## Think About This Now

Does your policy area have **eligibility thresholds**, **age cutoffs**, **geographic boundaries**, or **test score requirements**? If so, you may have a natural RDD waiting to be discovered. The key ingredients: a running variable you can measure, a cutoff that changes treatment status, and an argument that people just above and below are comparable.

Even if you don't run the analysis yourself, knowing how to *read* RDD (and DiD, and IV, and experiments) critically is a career-long skill.
:::