---
title: "Panel Data, Fixed Effects, and Multilevel Models"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

## Panel Data

::: incremental
-   **Panel data** (or longitudinal data) tracks the same units over multiple time periods

    -   A panel survey where the same individuals are re-interviewed each year

    -   Test scores for schools measured annually

    -   State-level unemployment rates over decades

-   Because the same units appear multiple times, our standard OLS assumption of independent errors is violated

    -   Are independent errors plausible when the same school appears 10 times in your data?
:::

## Hierarchical Data

::: incremental
-   Sometimes data is **nested** in ways that create a hierarchy

    -   Students are nested in classrooms

    -   Classrooms are nested in schools

    -   Schools are nested in districts

-   If we don't account for this structure, our standard errors will be wrong and our estimates may be biased
:::

## Why Should MPAs Care?

::: incremental
-   Almost all policy data has one of these structures

    -   Program evaluations track participants over time (panel)

    -   Education data has students in schools in districts (hierarchical)

    -   Health data has patients in hospitals in regions (hierarchical)

-   If you ignore this structure, you will get **wrong answers** — sometimes dramatically wrong

-   Today we learn the workhorse tool for dealing with this: **fixed effects**. And we'll preview a useful alternative: **multilevel models**.
:::

## Today's Roadmap

::: incremental
1.  **The fixed effects idea** — controlling for the unit
2.  **Visual walkthrough** — seeing FE in action
3.  **The mechanics** — equations, `fixest` in R, gapminder
4.  **Policy example** — housing prices across cities
5.  **Assumptions and caveats** — what FE can't do
6.  **Two-way fixed effects** — and why it's tricky
7.  **Beyond FE** — random effects and multilevel models
:::

# Part 1: The Fixed Effects Idea

## Example: International Trade Policy {.smaller}

::: incremental
-   Imagine we want to know whether a visit from Germany's chancellor increases trade with Germany the following year

-   Global trade data is "stacked": each row is a country-year, with each country entering the data many times

-   Trade with Germany depends on culture, history, geography, resource endowments, and more

    -   This would make a very messy DAG. How can we deal with all these confounders?
:::

## The Messy DAG

```{r}
#| echo: false

library(dagitty)

g <- dagitty("
dag {
  bb=\"0,0,5,4\"

  CultureOfPopulation   [pos=\"2,0\"]
  CurrentPolitics       [pos=\"0,2\"]
  Geography            [pos=\"2,2\"]
  ChancellorVisit      [pos=\"0,3\"]
  HistoryWithGermany   [pos=\"4,2\"]
  TradeWithGermany     [pos=\"4,3\"]

  CultureOfPopulation -> ChancellorVisit
  CultureOfPopulation -> TradeWithGermany

  Geography -> ChancellorVisit
  Geography -> TradeWithGermany

  CurrentPolitics -> ChancellorVisit
  CurrentPolitics -> TradeWithGermany

  HistoryWithGermany -> TradeWithGermany
  HistoryWithGermany -> ChancellorVisit

  ChancellorVisit -> TradeWithGermany
}
")

plot(g, text = FALSE, use_labels = "name")
```

## Adding a Fixed Effect

::: incremental
-   If we add a **fixed effect for country**, we control for everything about that country that doesn't change (or changes very slowly) over time

-   Controlling for "France" implicitly controls for France's proximity to Germany, its resource endowments, its political culture, and its history with Germany

    -   All of those confounders — absorbed in one move!
:::

## The Clean DAG

```{r}
#| echo: false

library(dagitty)

g <- dagitty('
dag {
  CurrentPolitics      [pos="0,1"]
  ChancellorVisit      [pos="0,3"]
  TradeWithGermany     [pos="4,3"]
  Country              [pos="4,1"]

  CurrentPolitics -> ChancellorVisit
  CurrentPolitics -> TradeWithGermany

  Country         -> ChancellorVisit
  Country         -> TradeWithGermany

  ChancellorVisit -> TradeWithGermany
}
')
plot(g, text = FALSE, use_labels = "name")
```

## What Do Fixed Effects Actually Do?

::: incremental
-   Fixed effects **control for the unit** — country, individual, state, school, etc.

-   As with any control, this means we hold the unit constant and look only at variation **within** the unit

-   **Key point:** We throw away all "between-unit" variation. We only use changes *within* each unit over time.
:::

. . .

::: {.callout-note}
## In English
Fixed effects let you ask: "For the **same country**, in years when the chancellor visited, did trade go up compared to years when they didn't visit?" You're never comparing France to Finland — only France to itself.
:::

# Part 2: Visual Walkthrough

## Simple Example: Exercise and Colds {.smaller}

Imagine we tracked two people and looked at how many hours per week they exercised and how many colds they got every year:

```{r}
#| echo: true

library(tidyverse)

df <- data.frame(
  Individual = c("Esnaina", "Esnaina", "Mariam", "Mariam"),
  Year       = c(2024, 2025, 2024, 2025),
  Exercise   = c(5, 7, 4, 3)
)

df_table <- df %>%
  group_by(Individual) %>%
  mutate(
    MeanExercise   = mean(Exercise),
    WithinExercise = Exercise - MeanExercise
  )

df_table
```

The "within" column is the deviation from each person's own average. This is what fixed effects use.

## Pooled Data {.smaller}

```{r}
#| echo: false

df <- tibble(
  Individual = rep(c("Esnaina", "Mariam"), each = 5),
  Exercise   = c(1, 2, 3, 4, 4,   4, 5, 5, 7, 8),
  Colds      = c(4, 3, 5, 3, 2,   7, 8, 7, 5, 6)
)

ggplot(df, aes(x = Exercise, y = Colds, color = Individual, shape = Individual)) +
  geom_point(size = 3) +
  labs(
    title = "Pooled Data",
    x = "Exercise Hours Per Week",
    y = "Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

What is the relationship between exercise and colds? It looks... positive?

## Pooled Regression

```{r}
#| echo: false

ggplot(df, aes(x = Exercise, y = Colds, color = Individual, shape = Individual)) +
  geom_point(size = 3) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Pooled Regression",
    x = "Exercise Hours Per Week",
    y = "Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

The pooled regression says more exercise → more colds. That can't be right!

## Adding Individual Means

```{r}
#| echo: false

df_means <- df %>%
  group_by(Individual) %>%
  summarize(
    mean_exercise = mean(Exercise),
    mean_colds    = mean(Colds)
  )

ggplot(df, aes(x = Exercise, y = Colds, color = Individual, shape = Individual)) +
  geom_point(size = 3) +
  geom_vline(
    data = df_means,
    aes(xintercept = mean_exercise, color = Individual),
    linetype = "dashed"
  ) +
  geom_hline(
    data = df_means,
    aes(yintercept = mean_colds, color = Individual),
    linetype = "dashed"
  ) +
  labs(
    title = "Raw Data with Individual Means",
    x     = "Exercise Hours Per Week",
    y     = "Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

Think of each person's mean as their personal origin (0,0).

## Just the Between Variation

```{r}
#| echo: false

ggplot(df_means, aes(x = mean_exercise, y = mean_colds, color = Individual, shape = Individual)) +
  geom_point(size = 4) +
  labs(
    title = "Just Between Variation",
    x = "Mean Exercise Hours Per Week",
    y = "Mean Colds Per Year"
  ) +
  theme_minimal(base_size = 14)
```

These are the person-specific factors we're going to control away. Mariam exercises more *and* gets more colds — but that's a between-person difference, not a causal effect.

## Within-Person Variation

```{r}
#| echo: false

df_centered <- df %>%
  group_by(Individual) %>%
  mutate(
    Exercise_centered = Exercise - mean(Exercise),
    Colds_centered    = Colds    - mean(Colds)
  )

ggplot(df_centered, aes(x = Exercise_centered, y = Colds_centered)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(size = 3) +
  facet_wrap(~ Individual) +
  labs(
    title = "Within Variation for Each Individual",
    x = "Exercise (Deviation from Individual Mean)",
    y = "Colds (Deviation from Individual Mean)"
  ) +
  theme_minimal(base_size = 14)
```

Now we're asking: "In years when *this person* exercised more than usual, did they get fewer colds than usual?"

## Combining and Regressing

```{r}
#| echo: false

ggplot(df_centered, aes(x = Exercise_centered, y = Colds_centered)) +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(aes(color = Individual), size = 3) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Centered Data (Both Individuals), Single Regression Line",
    x = "Exercise (Deviation from Mean)",
    y = "Colds (Deviation from Mean)"
  ) +
  theme_minimal(base_size = 14)
```

Now the relationship is **negative** — in years when a person exercises more than their own average, they get fewer colds. The pooled regression had it backwards!

## The Intuition

::: incremental
-   We removed all between-person variation by subtracting each person's mean

-   By using fixed effects, **we lose the ability to use between-group variation**

    -   If you're interested in between-group comparisons, you can't use FE!

-   By controlling for an individual, we also control for all **time-invariant** features of that person

-   We still need to control for things that **vary within the individual over time**
:::

# Part 3: The FE Equation and fixest

## The Fixed Effects Equation

$$Y_{it} = \beta_{i} + \beta_{1}X_{it} + \epsilon_{it}$$

::: {.callout-note}
## In English
$Y_{it}$ is the outcome for unit $i$ at time $t$. The key difference from standard OLS: we now have **unit-specific intercepts** $\beta_i$ — each country (or person, school, etc.) gets its own baseline level. We only estimate the effect of $X$ from variation *within* each unit.
:::

## Varying Intercepts = Dummy Variables

::: {.callout-tip}
It may not be immediately obvious, but letting the intercept vary by individual is exactly the same as adding a dummy (binary) variable for each individual!
:::

. . .

A control for "India" tells us the change we'd expect in Y given that we're in India rather than the reference category. Fixed effects just add a dummy for *every* unit.

. . .

Equivalently, we can "de-mean" the data:

$$Y_{it} - \bar{Y}_{i} = \beta_{1}(X_{it} - \bar{X}_{i}) + \epsilon_{it}$$

This is exactly what our visual walkthrough did — subtract each person's mean.

## Gapminder: Pooled Regression {.smaller}

```{r}
#| echo: true

library(tidyverse)
library(gapminder)
library(ggrepel)

df_gap <- gapminder %>%
  filter(country %in% c("Brazil", "India"))

ggplot(df_gap, aes(x = gdpPercap, y = lifeExp, color = country)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = year), size = 3, min.segment.length = 0) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE,
              linetype = "dashed", color = "black") +
  scale_x_log10(labels = scales::dollar_format(prefix = "$")) +
  labs(
    title = "Life Expectancy vs. GDP per Capita (Pooled Regression)",
    subtitle = "India and Brazil, 1952–2007 (Gapminder)",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)",
    color = "Country"
  ) +
  theme_minimal(base_size = 14)
```

## Gapminder: Country-Specific Intercepts {.smaller}

```{r}
#| echo: true

model_fe <- lm(lifeExp ~ log(gdpPercap) + country, data = df_gap)

predict_df <- expand.grid(
  country   = c("Brazil", "India"),
  gdpPercap = seq(min(df_gap$gdpPercap), max(df_gap$gdpPercap), length.out = 100)
)
predict_df$lifeExp_pred <- predict(model_fe, newdata = predict_df)

ggplot(df_gap, aes(x = gdpPercap, y = lifeExp, color = country)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = year), size = 3, min.segment.length = 0) +
  geom_line(data = predict_df,
            aes(x = gdpPercap, y = lifeExp_pred, color = country),
            linetype = "dashed", size = 1) +
  scale_x_log10(labels = scales::dollar_format(prefix = "$")) +
  labs(
    title = "Life Expectancy vs. GDP per Capita (Fixed Effects)",
    subtitle = "Same slope, different intercepts: India starts ~15 years higher at any GDP level",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)",
    color = "Country"
  ) +
  theme_minimal(base_size = 14)
```

## The fixest Package {.smaller}

```{r}
#| echo: true

library(fixest)

# Notation: y ~ x | fixed_effect(s)
m_fe <- feols(lifeExp ~ log(gdpPercap) | country, data = gapminder)
summary(m_fe)
```

. . .

::: {.callout-note}
## How to Interpret This
"In a year where GDP per capita is **one log-unit higher than it typically is for that country**, we expect life expectancy to be about **9.8 years longer than it typically is for that country**." Everything is relative to the unit's own average — that's the fixed effect at work.
:::

Note: `fixest` also works with GLMs for non-continuous outcomes.

# Part 4: Policy Example — Housing Prices

## The Question

::: incremental
-   Does **building more houses** lead to increases or decreases in home prices?

-   We'd want to control for crime, school quality, amenities...

-   But housing prices and new construction are both driven by **unobservable, city-specific factors** — desirability, zoning culture, land availability

-   We can control away those city-specific factors with fixed effects (as long as they don't vary over time)
:::

## Simulated Data {.smaller}

```{r}
#| echo: true

library(tidyverse)
set.seed(30317)

cities <- c("Atlanta", "Chicago", "Houston", "Boston", "New York",
            "Cleveland", "Los Angeles", "San Francisco", "Sacramento",
            "Seattle", "Philadelphia", "Washington DC", "Dallas")

n_obs_per_city <- 100

# City "popularity" drives both prices and construction
city_popularity <- runif(length(cities), min = -1, max = 1)
city_intercepts <- 850000 + 650000 * city_popularity
mean_new_houses <- 50 + 20 * city_popularity

# True within-city effects
beta_crime      <- -300
beta_edu        <-  1500
beta_new_houses <- -450  # More supply → lower prices WITHIN a city

sim_data <- data.frame()
for (i in seq_along(cities)) {
  crime_vals      <- rnorm(n_obs_per_city, mean = 50, sd = 10)
  edu_vals        <- rnorm(n_obs_per_city, mean = 5, sd = 1)
  new_houses_vals <- rpois(n_obs_per_city, lambda = pmax(mean_new_houses[i], 1))

  price_vals <- city_intercepts[i] +
    beta_crime * crime_vals + beta_edu * edu_vals +
    beta_new_houses * new_houses_vals +
    rnorm(n_obs_per_city, mean = 0, sd = 30000)

  sim_data <- rbind(sim_data, data.frame(
    city = cities[i], crime = crime_vals, edu_quality = edu_vals,
    new_houses = new_houses_vals, price = price_vals
  ))
}
```

## Pooled Data: The Wrong Answer

```{r}
#| echo: false

ggplot(sim_data, aes(x = new_houses, y = price, color = city)) +
  geom_point(size = 2) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Pooled: New Houses vs. Housing Price",
       x = "New Houses Built", y = "Average Housing Price")
```

The pooled regression says more houses → higher prices. But this is confounded by city popularity!

## City-Level Averages: The Between Story

```{r}
#| echo: false

city_means <- sim_data %>%
  group_by(city) %>%
  summarise(avg_price = mean(price), avg_new_houses = mean(new_houses))

ggplot(city_means, aes(x = avg_new_houses, y = avg_price)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Between Cities: More Building = Higher Prices?",
       subtitle = "Each point = one city's average. Popular cities build more AND have higher prices.",
       x = "Avg New Houses", y = "Avg Housing Price")
```

## Within-City: The True Story

```{r}
#| echo: false

ggplot(sim_data, aes(x = new_houses, y = price)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ city, scales = "free") +
  theme_minimal() +
  labs(title = "Within Each City: More Supply → Lower Prices",
       subtitle = "Fixed effects isolate this within-city variation.",
       x = "New Houses Built", y = "Housing Price")
```

## Naive vs. Fixed Effects Model {.smaller}

```{r}
#| echo: true

# Pooled (naive) model
naive_model <- lm(price ~ new_houses + crime + edu_quality, data = sim_data)
summary(naive_model)
```

## Fixed Effects Model {.smaller}

```{r}
#| echo: true

library(fixest)
fe_model <- feols(price ~ new_houses + crime + edu_quality | city, data = sim_data)
summary(fe_model)
```

## Why the Naive Model Gets It Wrong

::: {.callout-important}
## The Simpson's Paradox of Policy Analysis
**Popular cities** have higher housing prices *and* more new homes. When you pool all cities together, you capture that positive *between-city* correlation. Fixed effects remove it, revealing the **negative within-city** effect: more supply → lower prices. This is exactly the kind of mistake that leads to bad policy recommendations.
:::

# Part 5: Assumptions and Caveats

## What Fixed Effects Require

::: incremental
-   Each regressor must **vary over time** for at least some units, and cannot be perfectly collinear with the fixed effects

-   Fixed effects **cannot solve reverse causality**

    -   A model of police spending will predict that more police *causes* more crime. Why? Because cities hire more police *in response to* crime.

-   Cannot handle **time-variant** unobserved heterogeneity

-   Time-invariant variables must have **time-invariant effects** (see Ren and Allison on Canvas)
:::

## Should You Cluster Your Standard Errors?

::: incremental
-   Standard errors assume independence, but **clustered data** violates this

-   Conventional wisdom: cluster at the level of your fixed effects. `fixest` does this automatically.

-   But clustering is only necessary if:
    -   You have **treatment effect heterogeneity** across clusters
    -   Or your **treatment assignment is clustered** (e.g., a whole classroom gets treated)

-   See the Abadie et al paper on Canvas for much more detail
:::

# Part 6: Two-Way Fixed Effects

## Can We Control for Multiple Dimensions? {.smaller}

::: incremental
-   Can we get greedy and control for both individuals *and* the city they live in?

-   Yes! But now we only use variation that is within-individual *and* within-city

    -   Individuals who **never move** don't contribute to the regression — they have no variation on city, so their effect is "absorbed"
:::

## Two-Way FE With Time

::: incremental
-   It would be very nice to adjust for both unit-specific *and* time-specific unobserved confounders

-   This was extremely popular in development economics papers

-   But Imai and Kim show that: *"It is impossible to simultaneously and non-parametrically adjust for unit-specific and time-specific unobserved confounders under the two-way fixed effects framework"*

    -   Robert Kubinec has a good blog post on TWFE issues: <https://www.robertkubinec.com/post/fixed_effects/>

-   We will return to these points when we cover **Diff-in-Diff**
:::

# Part 7: Beyond Fixed Effects — Random Effects and Multilevel Models

## Three Ways to Handle Grouped Data {.smaller}

We've now seen two extremes. There's a middle ground:

::: incremental
1.  **Complete pooling** — ignore groups entirely. One regression line for everyone.
    -   Problem: ignores group-level differences → biased estimates

2.  **No pooling / Fixed Effects** — give each group its own intercept. No borrowing across groups.
    -   Problem: small groups get noisy, extreme estimates

3.  **Partial pooling / Random Effects** — group intercepts are drawn from a distribution. Small groups get "shrunk" toward the overall mean.
    -   Borrows strength from other groups to stabilize estimates
:::

## What Is Partial Pooling?

::: {.callout-note}
## The Shrinkage Idea
Imagine you're rating schools by test scores. A school with 500 students and a 95% pass rate — you trust that number. But a school with 8 students and a 100% pass rate? That's probably just luck. **Partial pooling** pulls that small school's estimate back toward the overall average. The less data a group has, the more it gets "shrunk" toward the grand mean.
:::

. . .

This is the core idea behind **random effects** and **multilevel/hierarchical models**.

## Schools in Districts: Simulated Data {.smaller}

```{r}
#| echo: true

library(tidyverse)
set.seed(30317)

# 30 schools across 5 districts, with varying sizes
n_schools <- 30
n_districts <- 5

school_data <- tibble(
  school_id   = 1:n_schools,
  district_id = sample(1:n_districts, n_schools, replace = TRUE),
  # School-level "quality" — this is what FE/RE will estimate
  school_effect = rnorm(n_schools, mean = 0, sd = 10),
  # Number of students per school (some small, some large)
  n_students = sample(c(15, 25, 50, 100, 200), n_schools, replace = TRUE)
)

# Generate student-level data
students <- school_data %>%
  uncount(n_students) %>%
  group_by(school_id) %>%
  mutate(
    student_id = row_number(),
    # Class size varies within schools
    class_size = rnorm(n(), mean = 25, sd = 5),
    # Test score depends on class size + school quality + noise
    test_score = 70 + school_effect - 0.8 * class_size + rnorm(n(), 0, 8)
  ) %>%
  ungroup()

students %>% select(school_id, district_id, student_id, class_size, test_score) %>% head(8)
```

## Complete Pooling: Ignore Schools

```{r}
#| echo: true

# One regression line — pretend schools don't exist
pooled_model <- lm(test_score ~ class_size, data = students)
summary(pooled_model)$coefficients
```

. . .

This gives us *a* number, but it ignores that students in the same school share unobserved factors.

## No Pooling: School Fixed Effects

```{r}
#| echo: true

library(fixest)

# Each school gets its own intercept
fe_model <- feols(test_score ~ class_size | school_id, data = students)
summary(fe_model)$coefficients
```

. . .

Unbiased — but small schools will have very noisy intercepts.

## Partial Pooling: Random Effects With lme4 {.smaller}

```{r}
#| echo: true

library(lme4)

# Random intercept for school: (1 | school_id)
re_model <- lmer(test_score ~ class_size + (1 | school_id), data = students)
summary(re_model)$coefficients
```

. . .

::: {.callout-note}
## In English
The `(1 | school_id)` syntax says: "Let each school have its own intercept, but assume those intercepts come from a normal distribution." Schools with lots of students keep their own estimate. Schools with few students get pulled toward the overall average.
:::

## Seeing the Shrinkage {.smaller}

```{r}
#| echo: true

# Extract school intercepts from FE and RE
fe_intercepts <- fixef(fe_model)$school_id  # Extract school-level fixed effects
re_intercepts <- ranef(re_model)$school_id$`(Intercept)`

# Get school sizes for the plot
school_sizes <- students %>%
  group_by(school_id) %>%
  summarise(n = n()) %>%
  arrange(school_id)

# Build comparison dataframe
compare_df <- tibble(
  school = 1:n_schools,
  n_students = school_sizes$n,
  FE = as.numeric(fe_intercepts[1:n_schools]),
  RE = as.numeric(re_intercepts[1:n_schools])
) %>%
  pivot_longer(cols = c(FE, RE), names_to = "Model", values_to = "Intercept")

ggplot(compare_df, aes(x = factor(school), y = Intercept, color = Model, size = n_students)) +
  geom_point(alpha = 0.7, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_size_continuous(range = c(1, 5)) +
  labs(title = "FE vs. RE School Intercepts: Shrinkage in Action",
       subtitle = "Small schools (small dots) get pulled toward zero in the RE model",
       x = "School", y = "School-Specific Intercept",
       size = "# Students") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(size = 8))
```

## The Key Tradeoff: FE vs. RE

::: incremental
-   **Fixed Effects:** Unbiased even if school quality is correlated with class size (the safe choice for **causal inference**)

-   **Random Effects:** More efficient (smaller SEs), but assumes school effects are **uncorrelated** with your predictors

    -   This is often implausible in causal settings — better schools might also have smaller classes

-   **Rule of thumb:** For causal questions, use FE. For prediction and description (especially with many small groups), RE/multilevel can be better.

-   There's a formal test (the **Hausman test**) that checks whether FE and RE give meaningfully different answers. If they do, FE is the safer choice.
:::

## When Do Multilevel Models Shine in Policy? {.smaller}

::: incremental
-   **School report cards:** Small schools get extreme pass rates by chance. Multilevel models give more honest rankings by shrinking small-school estimates toward the average.

-   **Hospital quality ratings:** CMS star ratings use hierarchical models for exactly this reason — a hospital with 10 surgeries and zero deaths isn't necessarily the best hospital.

-   **Small-area poverty estimation:** The Census Bureau uses multilevel models to estimate poverty rates for small geographic areas where sample sizes are tiny.

-   **Any time you have many groups with few observations each** — partial pooling prevents you from over-interpreting noise as signal.
:::

## Further Reading

::: incremental
-   McElreath's *Statistical Rethinking* (2020) — the best intuitive introduction to multilevel models

-   Gelman and Hill (2006) — the classic applied reference

-   For this class: fixed effects are the main tool you need. But knowing that multilevel models exist (and when they're useful) makes you a more sophisticated consumer of evidence.
:::

# Wrapping Up

## What We Covered Today

::: incremental
1.  **Panel and hierarchical data** — most policy data has this structure
2.  **Fixed effects** — control for the unit, use only within-unit variation
3.  **The visual intuition** — pooled regression can get the sign wrong; FE fixes it
4.  **fixest in R** — the practical tool for FE regression
5.  **Housing prices** — a full policy simulation showing why pooled estimates mislead
6.  **Assumptions** — FE can't fix reverse causality or time-varying confounders
7.  **Two-way FE** — powerful but with important limitations
8.  **Random effects / multilevel models** — partial pooling for when you have many small groups
:::

## What's Next

::: incremental
-   **Experiments** — the gold standard for causal inference

-   Then **design-based causal inference:**
    -   Difference-in-Differences
    -   Regression Discontinuity
    -   Instrumental Variables
:::

## For Your Capstone

::: {.callout-tip}
## Think About This Now
Does your data have a panel or hierarchical structure? If you're looking at schools, hospitals, states, or any repeated observations — you almost certainly need fixed effects. And if you're trying to rank or rate small units, multilevel models might give you more honest answers than raw averages.
:::
