---
title: "Difference-in-Differences for Policy Evaluation"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: true
execute:
  fig-align: center
---

## The Setup

::: incremental
-   Often in policy, a treatment occurs at a specific point in time

    -   We observe the world before the treatment

    -   And after the treatment

    -   We want to know how much of the difference is **caused by the treatment**

-   One obvious backdoor path: **time trends**

    -   Things change over time regardless of the policy — the economy grows, demographics shift, technology improves
:::

## The Problem: Time as a Confounder

```{r}
#| echo: false

library(tidyverse)
library(dagitty)
library(ggdag)

dag <- dagitty("
dag {
  Outcome      [pos=\"2,0\"]
  Time         [pos=\"1,-1\"]
  Treatment    [pos=\"0,0\"]

  Time -> Treatment
  Time -> Outcome
  Treatment -> Outcome
}
")
ggdag(dag, text = FALSE, use_labels = "name") +
  coord_cartesian(clip = "off") +
  theme_dag() +
  theme(plot.margin = margin(t = 20, r = 10, b = 10, l = 10), legend.position = "none")
```

If all groups are treated at the same time, we **cannot** separate the treatment effect from the time trend.

## Why We Need a Control Group

::: incremental
-   Without a control group, we could try event studies or interrupted time series

    -   But these rely on assumptions about what the trend *would have been* — often not very plausible

-   If we have a group that is **not treated**, we can use them as a benchmark

    -   The logic of a "natural experiment": comparing a treatment group to a control group over time
:::

## The DiD DAG

```{r}
#| echo: false

library(dagitty)
library(ggdag)

dag <- dagitty("
dag {
  Outcome      [pos=\"2,0\"]
  Time         [pos=\"1,-1\"]
  Treatment    [pos=\"0,0\"]
  Group        [pos=\"1,1\"]

  Time -> Treatment
  Time -> Outcome
  Treatment -> Outcome
  Group -> Treatment
  Group -> Outcome
}
")
ggdag(dag, text = FALSE, use_labels = "name") +
  coord_cartesian(clip = "off") +
  theme_dag() +
  theme(plot.margin = margin(t = 20, r = 10, b = 10, l = 10), legend.position = "none")
```

We need to close both backdoor paths — through Time *and* through Group.

## The DiD Logic

::: incremental
-   **Close the Group backdoor:** Look at *within-group* changes over time (we know how to do this — fixed effects!)

-   **Close the Time backdoor:** Compare the within-group changes between the treated and control groups

-   We're asking: **How much more did the treated group change compared to the control group?**
:::

## Why Should MPAs Care?

::: incremental
-   DiD is probably the **most commonly used** causal inference tool in applied policy research

-   Minimum wage studies, Medicaid expansion, education reforms, environmental regulations — all use DiD

-   If you read policy evaluations, you *will* encounter DiD. If you do evaluation work, you'll likely *use* it.
:::

## Today's Roadmap

::: incremental
1.  **The classic 2×2 DiD** — Card & Krueger's minimum wage study
2.  **The parallel trends assumption** — the key to credibility
3.  **DiD in R** — a full worked example
4.  **Event studies** — visualizing effects over time
5.  **Staggered adoption** — when different units get treated at different times
6.  **Modern solutions** — new methods that fix problems with classic TWFE
7.  **Practical advice** — a checklist for credible DiD designs
:::

# Part 1: The Classic 2×2 DiD

## Card & Krueger (1994): Minimum Wage and Employment {.smaller}

::: incremental
-   In April 1992, **New Jersey** raised its minimum wage from \$4.25 to \$5.05

-   Neighboring **Pennsylvania** did not change its minimum wage

-   Conventional economic theory predicted: higher minimum wage → fewer jobs

-   Card & Krueger surveyed fast-food restaurants in both states, before and after the increase

-   **The question:** Did NJ's minimum wage increase reduce fast-food employment?
:::

## The 2×2 Table

| | Before (Feb 1992) | After (Nov 1992) | Change |
|---|---|---|---|
| **New Jersey** (treated) | 20.44 FTE | 21.03 FTE | +0.59 |
| **Pennsylvania** (control) | 23.33 FTE | 21.17 FTE | -2.16 |

. . .

**DiD Estimate:** (+0.59) − (−2.16) = **+2.76 FTE employees**

. . .

::: {.callout-important}
The minimum wage increase was associated with an **increase** in employment — the opposite of the standard prediction! This paper changed the entire minimum wage debate.
:::

## The DiD Formula

$$\widehat{\delta}_{DiD} = (\bar{Y}_{treat}^{post} - \bar{Y}_{treat}^{pre}) - (\bar{Y}_{control}^{post} - \bar{Y}_{control}^{pre})$$

::: {.callout-note}
## In English
Take the **change** in the treated group. Subtract the **change** in the control group. What's left is the treatment effect — the part of the change that can't be explained by common time trends.
:::

# Part 2: The 2×2 DiD Estimator

## DiD as a Regression {.smaller}

We can recover the DiD estimate from a **two-way fixed effects** regression:

$$Y_{it} = \alpha_{g} + \alpha_{t} + \beta \cdot \text{Treated}_{it} + \epsilon_{it}$$

Or equivalently, using the interaction form:

$$Y_{it} = \beta_{0} + \beta_{1}\text{Group}_{i} + \beta_{2}\text{Post}_{t} + \beta_{3}(\text{Group}_{i} \times \text{Post}_{t}) + \epsilon_{it}$$

::: {.callout-note}
## In English
$\beta_3$ is the DiD estimate — it captures the **extra change** in the treated group beyond what the control group experienced. The group dummy absorbs level differences; the time dummy absorbs common trends.
:::

## Simulating Card & Krueger in R {.smaller}

```{r}
#| echo: true

library(tidyverse)
library(fixest)
set.seed(30317)

n_restaurants <- 400  # 200 per state

ck_data <- tibble(
  restaurant = 1:n_restaurants,
  state = rep(c("NJ", "PA"), each = n_restaurants/2),
  # PA starts with slightly more employees
  base_fte = ifelse(state == "PA", rnorm(n_restaurants/2, 23.3, 4),
                                    rnorm(n_restaurants/2, 20.4, 4))
) %>%
  # Create before and after observations
  crossing(period = c("before", "after")) %>%
  mutate(
    post = ifelse(period == "after", 1, 0),
    treated = ifelse(state == "NJ", 1, 0),
    # PA declines slightly; NJ gets a positive treatment effect
    fte = base_fte +
      ifelse(post == 1, -2.2, 0) +                   # Common time trend (decline)
      ifelse(post == 1 & treated == 1, 2.76, 0) +    # Treatment effect
      rnorm(n(), 0, 3)                                # Noise
  )
```

## Running the DiD Regression {.smaller}

```{r}
#| echo: true

# Method 1: Interaction form
did_interaction <- lm(fte ~ treated * post, data = ck_data)
summary(did_interaction)$coefficients
```

. . .

```{r}
#| echo: true

# Method 2: TWFE with fixest
did_twfe <- feols(fte ~ treated:post | state + period, data = ck_data)
summary(did_twfe)
```

. . .

Both methods recover approximately the same DiD estimate — the treatment effect of NJ's minimum wage increase.

## Why DiD Recovers the ATT, Not the ATE {.smaller}

::: incremental
-   In a randomized experiment, treatment and control groups are balanced → we get the **ATE**

-   In DiD, groups aren't randomly assigned — NJ and PA differ in many ways

-   What DiD recovers is the **Average Treatment Effect on the Treated (ATT)**

    -   "Among the units that *actually received* the treatment, what was the average effect?"

-   We can't say (without more assumptions) what would have happened if PA had also raised its minimum wage
:::

::: {.callout-note}
## In English
DiD tells you: "The minimum wage increase added about 2.76 FTE jobs per restaurant **in New Jersey**." It doesn't tell you what would happen in Pennsylvania — because PA might be different in ways that matter.
:::

# Part 3: The Parallel Trends Assumption

## The Key Assumption

::: incremental
-   DiD requires that, **without the treatment**, the treated group would have followed the same trend as the control group

-   This is called the **parallel trends assumption**

-   It does **not** require that the groups are at the same *level* — just that they would have *changed at the same rate*
:::

. . .

::: {.callout-important}
## The Unverifiable Part
We can check whether trends were parallel *before* the treatment. But the assumption is about what would have happened *after* treatment in the absence of treatment — which is a counterfactual we can never observe. Pre-trends are suggestive, not definitive.
:::

## Good vs. Bad Parallel Trends {.smaller}

```{r}
#| echo: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)

df <- tibble(
  time = rep(1:10, times = 4),
  group = rep(c("Treatment", "Control"), each = 10, times = 2),
  panel = rep(c("Parallel Prior Trends", "Converging Prior Trends"), each = 20),
  outcome = c(
    c(2, 2, 3, 4, 5, 7, 9, 10, 9.8, 9.5),
    c(1, 1, 2, 3, 3.5, 4.5, 5.5, 6, 5.5, 6),
    c(4, 3.5, 4, 5, 6, 6.5, 7.2, 9, 10, 10.5),
    c(1, 1.2, 2.2, 3.2, 4.8, 6.5, 8.5, 9.5, 9.7, 10)
  )
)

make_plot <- function(data, panel_title) {
  ggplot(data, aes(x = time, y = outcome, color = group)) +
    geom_line(size = 1) +
    geom_point(size = 2) +
    geom_vline(xintercept = 6, linetype = "dashed") +
    annotate("rect", xmin = 6, xmax = 10.5, ymin = -Inf, ymax = Inf, alpha = 0.1) +
    annotate("text", x = 6.3, y = max(data$outcome) * 0.2, label = "Treatment\nPeriod", hjust = 0) +
    scale_color_manual(values = c("Treatment" = "black", "Control" = "gray40")) +
    labs(title = panel_title, x = "Time", y = "Outcome") +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
}

p1 <- make_plot(df %>% filter(panel == "Parallel Prior Trends"), "(a) Parallel Prior Trends")
p2 <- make_plot(df %>% filter(panel == "Converging Prior Trends"), "(b) Converging Prior Trends")

p1 + p2
```

Panel (a): Pre-trends are parallel → DiD is credible. Panel (b): The control group was already converging → the "treatment effect" may just be the continuation of a pre-existing trend.

## When Is Parallel Trends Plausible?

::: incremental
-   **More plausible:**
    -   Two neighboring states with similar economic structures
    -   Units that were selected for treatment based on a fixed characteristic (not a trend)
    -   Short time horizons where trends are less likely to diverge

-   **Less plausible:**
    -   Comparing a booming city to a declining one
    -   Units that self-selected into treatment based on anticipated trends
    -   Very long time horizons with structural changes
:::

## Placebo Tests

::: incremental
-   Run DiD in **pre-treatment periods** to check for spurious effects

-   If treatment happens in April 2025 with monthly data:
    -   Run DiD comparing March to February, February to January, etc.
    -   Treat the later month as the "treatment period"

-   Finding significant effects in pre-treatment periods is strong evidence that parallel trends is violated

-   Finding null effects doesn't *prove* the assumption, but it's reassuring
:::

## Anticipation Effects

::: incremental
-   What if people change behavior **before** the policy takes effect?

    -   Workers might change jobs before a minimum wage increase

    -   Firms might hire ahead of a regulation deadline

-   Anticipation violates the assumption that treatment "turns on" at a specific time

-   **Solution:** Backdate the treatment period or exclude the periods just before treatment
:::

# Part 4: DiD in R — A Full Worked Example

## Simulating a State EITC Expansion {.smaller}

```{r}
#| echo: true

library(tidyverse)
library(fixest)
set.seed(30317)

n_states <- 2
n_years <- 10
treatment_year <- 6  # Treatment starts in year 6

eitc_data <- expand_grid(
  state = c("Treatment_State", "Control_State"),
  year = 1:n_years
) %>%
  mutate(
    treated_state = ifelse(state == "Treatment_State", 1, 0),
    post = ifelse(year >= treatment_year, 1, 0),
    # Treatment state starts lower, both trend upward
    base = ifelse(treated_state == 1, 55, 60) + 0.8 * year,
    # Treatment effect: 3 percentage points increase in LFP
    treatment_effect = ifelse(treated_state == 1 & post == 1, 3, 0),
    # Add some noise
    lfp = base + treatment_effect + rnorm(n(), 0, 0.8)
  )
```

## Visualizing the Data {.smaller}

```{r}
#| echo: true

ggplot(eitc_data, aes(x = year, y = lfp, color = state)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = treatment_year - 0.5, linetype = "dashed", color = "gray40") +
  annotate("text", x = treatment_year + 0.2, y = min(eitc_data$lfp) + 1,
           label = "EITC Expansion", hjust = 0, size = 4) +
  labs(title = "Labor Force Participation Before and After EITC Expansion",
       x = "Year", y = "Labor Force Participation (%)",
       color = "State") +
  theme_minimal(base_size = 14)
```

Pre-treatment trends look parallel — good sign for our DiD design!

## Running the DiD {.smaller}

```{r}
#| echo: true

# TWFE regression
eitc_did <- feols(lfp ~ treated_state:post | state + year, data = eitc_data)
summary(eitc_did)
```

. . .

::: {.callout-note}
## Interpretation
The EITC expansion is associated with approximately a 3 percentage-point increase in labor force participation in the treatment state, compared to what the control state experienced over the same period.
:::

# Part 5: Event Studies

## What Is an Event Study?

::: incremental
-   An event study plots the **DiD effect for each time period** relative to the treatment

-   X-axis: time periods before and after treatment (with the period just before treatment as the reference)

-   Y-axis: estimated coefficient (the difference between treated and control, relative to the reference period)

-   **Pre-treatment coefficients** should be ≈ 0 if parallel trends holds

-   **Post-treatment coefficients** show how the effect evolves over time
:::

## Why Event Studies Matter for Policy

::: incremental
-   Effects may **take time to materialize** — a job training program might not show results for a year

-   Effects may **fade over time** — an initial boost might disappear

-   Pre-treatment coefficients serve as a **visual diagnostic** for parallel trends

-   As Huntington-Klein puts it: the event study is the "visual test" of your research design
:::

## Building an Event Study With More Data {.smaller}

```{r}
#| echo: true

set.seed(30317)
n_units <- 30  # 15 treated, 15 control
n_years <- 12
treatment_year <- 7

unit_fes <- tibble(unit = 1:n_units, unit_fe = rnorm(n_units, 0, 3))

event_data <- expand_grid(
  unit = 1:n_units,
  year = 1:n_years
) %>%
  left_join(unit_fes, by = "unit") %>%
  mutate(
    treated_unit = ifelse(unit <= 15, 1, 0),
    state = ifelse(treated_unit == 1, "Treated", "Control"),
    post = ifelse(year >= treatment_year, 1, 0),
    # Event time: periods relative to treatment
    event_time = year - treatment_year,
    # Base outcome with unit FE + time trend
    base = 50 + unit_fe + 0.5 * year,
    # Dynamic treatment effect: grows over time after treatment
    te = case_when(
      treated_unit == 0 ~ 0,
      event_time < 0 ~ 0,
      event_time == 0 ~ 1.5,
      event_time == 1 ~ 2.5,
      event_time == 2 ~ 3.2,
      event_time == 3 ~ 3.5,
      event_time == 4 ~ 3.8,
      TRUE ~ 4.0
    ),
    outcome = base + te + rnorm(n(), 0, 2)
  )
```

## The Event Study Regression {.smaller}

```{r}
#| echo: true

# Event study with fixest: i() creates dummies for each event_time
# ref = -1 means the period just before treatment is the reference
es_model <- feols(outcome ~ i(event_time, treated_unit, ref = -1) | unit + year,
                  data = event_data)

# The event study plot
iplot(es_model,
      xlab = "Periods Relative to Treatment",
      ylab = "Estimated Effect",
      main = "Event Study: EITC Expansion Effect Over Time")
```

## Reading an Event Study Plot

::: incremental
-   **Pre-treatment coefficients** (left of 0): Should bounce around zero with confidence intervals crossing zero. If they don't, parallel trends is suspect.

-   **The jump at time 0**: The immediate treatment effect

-   **Post-treatment trajectory**: Does the effect grow? Stabilize? Fade?

-   **Confidence intervals**: Wider = less precise. If they include zero post-treatment, the effect isn't statistically significant.
:::

. . .

::: {.callout-tip}
## For Your Reports
Event study plots are becoming the **standard way** to present DiD results in policy research. If you ever do a DiD analysis, include one.
:::

# Part 6: The Staggered Adoption Problem

## What If Treatment Happens at Different Times?

::: incremental
-   In practice, policies are often adopted **at different times** by different units

    -   States expanded Medicaid in 2014, 2015, 2016, or not at all

    -   Cities adopted minimum wage increases in different years

    -   Schools implemented new curricula on different schedules

-   This is called **staggered adoption** — and it creates serious problems for classic TWFE
:::

## Visualizing Staggered Treatment {.smaller}

```{r}
#| echo: false

library(tidyverse)

stagger_viz <- tibble(
  State = c("Alabama", "Colorado", "Delaware", "Florida", "Georgia"),
  treat_year = c(2015, 2016, 2018, NA, NA)
) %>%
  crossing(Year = 2012:2020) %>%
  mutate(
    status = case_when(
      is.na(treat_year) ~ "Never Treated",
      Year < treat_year ~ "Not Yet Treated",
      Year >= treat_year ~ "Treated"
    ),
    status = factor(status, levels = c("Treated", "Not Yet Treated", "Never Treated"))
  )

ggplot(stagger_viz, aes(x = Year, y = State, fill = status)) +
  geom_tile(color = "white", size = 1) +
  scale_fill_manual(values = c("Treated" = "#2C7BB6", "Not Yet Treated" = "#FFFFBF",
                                "Never Treated" = "#D7191C")) +
  labs(title = "Staggered Treatment Adoption",
       fill = "Treatment Status") +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank())
```

Different states get treated at different times. Who serves as the "control" for whom?

## Why Classic TWFE Breaks {.smaller}

::: incremental
-   Classic TWFE estimates a **single treatment coefficient** from a weighted average of *all possible 2×2 comparisons*

-   The problem: some of those comparisons are **"forbidden"**

    -   TWFE compares **newly-treated units** to **already-treated units**

    -   An already-treated unit is being used as a "control" — but it's already affected by treatment!

-   If treatment effects **change over time** (grow, fade, or differ across groups), these forbidden comparisons contaminate the estimate

-   The weights can even be **negative** — meaning the estimate can show a negative effect when the true effect is positive for everyone
:::

## An Intuitive Example {.smaller}

::: {.callout-warning}
## The Negative Weights Problem (Cunningham's Framing)
Imagine a job training program that helps everyone — but the effect grows over time. Early adopters have large effects by now (they've had the program for years). Late adopters have small effects (just started).

TWFE compares late adopters (small positive effect) to early adopters (large positive effect). The "change" in the early-adopter "controls" is actually driven by their growing treatment effect — making it look like the new treatment *reduced* outcomes. The estimated effect can flip negative, even though **everyone benefited**.
:::

## When TWFE Is Fine vs. Not

::: incremental
-   **TWFE is fine if:**
    -   Treatment effects are constant over time and across groups
    -   All units are treated at the same time (the classic 2×2 case)
    -   You have a "never-treated" group and only compare to them

-   **TWFE is problematic if:**
    -   Treatment effects build up, fade, or vary across cohorts
    -   You have staggered adoption with no never-treated group
    -   You suspect dynamic or heterogeneous effects
:::

. . .

As Huntington-Klein puts it: "If you believe effects are constant, TWFE is fine. If you believe effects change over time or differ across groups — you need better tools."

# Part 7: Modern Solutions

## The Good News

Econometricians have developed several methods that fix the staggered DiD problem. The two most popular:

::: incremental
-   **Callaway and Sant'Anna (2021):** Estimate separate treatment effects for each treatment cohort at each time period, then aggregate. R package: `did`.

-   **Sun and Abraham (2021):** An interaction-weighted event-study estimator that avoids the contamination. Available in `fixest` via `sunab()`.

-   Both approaches share a key idea: **only make "clean" comparisons** — treated vs. not-yet-treated or never-treated, never treated vs. already-treated.
:::

## Simulating Staggered Adoption {.smaller}

```{r}
#| echo: true

set.seed(30317)

# 40 units: 10 treated in year 5, 10 in year 7, 10 in year 9, 10 never
stag_unit_fes <- tibble(unit = 1:40, unit_fe = rnorm(40, 0, 2))

stag_data <- expand_grid(
  unit = 1:40,
  year = 1:12
) %>%
  left_join(stag_unit_fes, by = "unit") %>%
  mutate(
    cohort = case_when(
      unit <= 10 ~ 5,
      unit <= 20 ~ 7,
      unit <= 30 ~ 9,
      TRUE ~ Inf  # Never treated
    ),
    treated = ifelse(year >= cohort, 1, 0),
    # Dynamic treatment effect: grows with time since treatment
    years_treated = pmax(year - cohort, 0),
    true_te = ifelse(is.finite(cohort) & treated == 1, 1 + 0.5 * years_treated, 0),
    # Outcome
    outcome = 50 + unit_fe + 0.3 * year + true_te + rnorm(n(), 0, 1.5)
  )
```

## TWFE vs. Sun-Abraham {.smaller}

```{r}
#| echo: true

library(fixest)

# Classic TWFE — potentially misleading with staggered treatment
twfe_model <- feols(outcome ~ treated | unit + year, data = stag_data)

# Sun & Abraham via fixest — only clean comparisons
# Need to set cohort to 1000 (or large number) for never-treated
stag_data_sa <- stag_data %>%
  mutate(cohort_sa = ifelse(is.infinite(cohort), 1000, cohort))

sa_model <- feols(outcome ~ sunab(cohort_sa, year) | unit + year,
                  data = stag_data_sa)
```

## Comparing the Two Approaches {.smaller}

```{r}
#| echo: true

# TWFE: single coefficient
summary(twfe_model)
```

. . .

```{r}
#| echo: true

# Sun & Abraham: event study plot
iplot(sa_model,
      xlab = "Periods Relative to Treatment",
      ylab = "Estimated Effect",
      main = "Sun & Abraham Event Study (Staggered DiD)")
```

## The Key Takeaway for MPAs

::: {.callout-important}
## Reading DiD Papers
If you encounter a DiD paper with **staggered treatment** that was published before ~2020, the estimates might be unreliable — especially if treatment effects are likely to grow or change over time. Papers using modern methods (Callaway & Sant'Anna, Sun & Abraham, or similar) have addressed this problem. When evaluating research, ask: "Is treatment staggered? If so, did they use a modern estimator?"
:::

# Part 8: Practical Advice

## A Checklist for Credible DiD Designs

::: incremental
1.  Is there a **clear treatment** and a **clear control group**?
2.  Is the **treatment timing plausibly exogenous** — not driven by anticipated outcomes?
3.  Do **pre-treatment trends look parallel**? (Plot them!)
4.  Are there **anticipation effects** that could blur the treatment timing?
5.  Is treatment **staggered**? If so, use modern methods.
6.  How **sensitive** are results to the choice of control group?
:::

## Common Pitfalls

::: incremental
-   **Confusing levels with trends:** Groups can be at very different levels — what matters is that they *change at the same rate* pre-treatment

-   **Cherry-picking the control group** to manufacture parallel trends

-   **Not clustering standard errors** at the level of treatment assignment

-   **Using classic TWFE with staggered adoption** when effects are dynamic

-   **Ignoring anticipation effects** that shift behavior before the treatment "officially" begins
:::

## Readings

::: incremental
-   **Cunningham, *Causal Inference: The Mixtape*** — Excellent chapter on DiD with real-world examples and R code. Freely available online.

-   **Huntington-Klein, *The Effect*** — Very intuitive treatment of DiD and event studies. Also freely available.

-   **Callaway & Sant'Anna (2021)** — The key paper on staggered DiD. Technical but the introduction is accessible.

-   **Goodman-Bacon (2021)** — "Difference-in-Differences with Variation in Treatment Timing." The paper that showed *why* TWFE breaks.
:::

# Wrapping Up

## What We Covered Today

::: incremental
1.  **The DiD idea** — use a control group to remove common time trends
2.  **Card & Krueger** — the classic 2×2 example that reshaped the minimum wage debate
3.  **The parallel trends assumption** — the core of DiD credibility
4.  **Event studies** — the standard way to present DiD results and test assumptions
5.  **Staggered adoption** — why classic TWFE breaks and how to spot the problem
6.  **Modern solutions** — Sun & Abraham and Callaway-Sant'Anna fix the staggered problem
7.  **Practical advice** — a checklist for credible designs
:::

## What's Next

::: incremental
-   **Regression Discontinuity Designs** — exploiting arbitrary cutoffs for causal inference

-   **Instrumental Variables** — finding exogenous sources of variation
:::

## For Your Capstone

::: {.callout-tip}
## Think About This Now
Is there a **policy change** in your area of interest with a natural control group? A state that changed a law while a neighbor didn't? A program that was rolled out in some districts before others? If so, DiD might be the right tool — and now you know the modern best practices for using it.
:::
