---
title: "Mini Problem Set 2: Visualizing U.S. State Poverty Data"
author: "Will Horne"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
    embed-resources: true
  pdf:
    toc: false
editor: visual
---

## Overview

In this problem set, you will practice data visualization using `ggplot2` while exploring poverty and income data across U.S. states. This builds on the data wrangling skills from Problem Set 1, but focuses on creating clear, informative graphics.

**Estimated time: 1-2 hours**

## Setup Instructions

1.  Download the `state_poverty_data.csv` file from the course website
2.  Save it in the same folder as this .qmd file
3.  Complete each question by adding your code in the provided code chunks

### Load your packages and data

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)

# Load the data
poverty_data <- read_csv("state_poverty_data.csv")
```

------------------------------------------------------------------------

## Part 1: Getting to Know the Data (25 points)

### Question 1.1: Explore the Data (5 points)

Use `glimpse()` to examine the dataset structure. Then answer:

-   How many states are included?
-   What variables are available?

```{r}
# Your code here


```

**Your written answer:**

### Question 1.2: Quick Summary (5 points)

What is the mean and median poverty rate across all states? What does the difference between mean and median suggest about the distribution?

*Hint: Use `summarize()` with `mean()` and `median()`*

```{r}
# Your code here


```

**Your written answer:**

### Question 1.3: Filtering the Data (15 points)

Let's practice using `filter()` to focus on subsets of the data.

**Part A (5 points):** Create a new dataframe called `high_poverty` that contains only states where the poverty rate is above 15%. How many states meet this criterion?

*Hint: Use `filter(poverty_rate > 15)`*

```{r}
# Your code here


```

**Your written answer:** How many states have poverty rates above 15%?

**Part B (5 points):** Create a new dataframe called `southern_states` that contains only states in the South region.

*Hint: Use `filter(region == "South")` — note the double equals sign!*

```{r}
# Your code here


```

**Part C (5 points):** Now combine what you know: filter to find Southern states that ALSO have a median income above \$60,000. Save this as `wealthy_south`.

*Hint: You can combine conditions with `&` (and) or use a comma: `filter(condition1, condition2)`*

```{r}
# Your code here


```

**Your written answer:** Which Southern states have median incomes above \$60,000?

------------------------------------------------------------------------

## Part 2: Creating Basic Visualizations (40 points)

### Question 2.1: Your First Scatter Plot (12 points)

Create a scatter plot showing the relationship between `median_income` (x-axis) and `poverty_rate` (y-axis).

**Requirements:**

-   Use `geom_point()`
-   Add a descriptive title with `ggtitle()`
-   Label your axes clearly with `labs()`
-   Apply `theme_minimal()`

```{r}
# Your code here


```

**Your written answer:** Describe the relationship you see. Does higher median income tend to be associated with higher or lower poverty rates?

### Question 2.2: Adding Color (12 points)

Improve your scatter plot by coloring points by `region` (Northeast, South, Midwest, West).

*Hint: Add `color = region` inside `aes()`*

```{r}
# Your code here


```

**Your written answer:** Do you notice any regional patterns?

### Question 2.3: A Bar Chart (16 points)

Create a bar chart showing the **average poverty rate by region**.

**Steps:**

1.  First, calculate the mean poverty rate for each region using `group_by()` and `summarize()`
2.  Then create a bar chart with `geom_col()`
3.  Add appropriate labels and a title

```{r}
# Step 1: Calculate regional averages
regional_avg <- poverty_data |>
  group_by(region) |>
  summarize(avg_poverty = mean(poverty_rate))

# Step 2: Create the bar chart
# Your code here


```

**Your written answer:** Which region has the highest average poverty rate?

------------------------------------------------------------------------

## Part 3: Making Publication-Quality Graphics (35 points)

### Question 3.1: A Polished Scatter Plot (17 points)

Take your scatter plot from Question 2.2 and make it publication-ready by adding:

-   A trend line using `geom_smooth(method = "lm", se = FALSE)`
-   Formatted axis labels (e.g., "Median Household Income (\$)" instead of "median_income")
-   A centered, larger title
-   Adjust text sizes using `theme()`

*Hint: Look at the example from the slides where we formatted the penguin plot*

```{r}
# Your code here


```

### Question 3.2: Histogram of Poverty Rates (18 points)

Create a histogram showing the distribution of poverty rates across all states.

**Requirements:**

-   Use `geom_histogram()` with `binwidth = 1`
-   Choose a fill color you like (e.g., `fill = "steelblue"`)
-   Add a white border to bars with `color = "white"`
-   Add appropriate title and axis labels
-   Use `theme_minimal()`

```{r}
# Your code here


```

**Your written answer:** Describe the shape of the distribution. Is it symmetric, left-skewed, or right-skewed? What is the typical (most common) poverty rate range?

------------------------------------------------------------------------

## Bonus Challenge (10 points)

Create a plot of your choice that tells an interesting story about poverty in America. Be creative! You might consider:

-   Comparing unemployment and poverty rates
-   Looking at education levels and poverty
-   Creating a faceted plot by region

```{r}
# Your code here (optional)


```

**Your written explanation (optional):** What does your visualization reveal?

------------------------------------------------------------------------

## Submission Instructions

1.  Make sure all code chunks run without errors
2.  Render this document to HTML or PDF
3.  Submit both the .qmd file and the rendered output

**Remember:** Good visualizations should be clear, honest, and tell a story!