---
title: "In Class 1-20"
author: "Will Horne"
format: html
editor: visual
---

## Chaining Multiple Operations with the Pipe

This block demonstrates how to chain multiple tidyverse operations together using the pipe operator (`|>`). We load the `palmerpenguins` dataset and then:

1.  **filter()** - Keep only penguins with body mass greater than 1500g
2.  **group_by()** - Group the data by species and sex
3.  **summarize()** - Calculate the average bill length and count for each group
4.  **arrange()** - Sort the results by average bill length in descending order

```{r}
library(tidyverse)
library(palmerpenguins)
new_penguins <- penguins |>
  filter(body_mass_g > 1500) |>
  group_by(species, sex) |>
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE),
            count = n()) |>
  arrange(desc(avg_bill_length))

```

## Basic Group-By and Summarize

This block shows a simpler example of grouping and summarizing. We create a small data frame with names and test scores, then calculate the average score for each person. This demonstrates the core pattern: `group_by()` defines the groups, and `summarize()` collapses each group into a single summary row.

```{r}
# Example dataset
df <- data.frame(
  Name = c('Alice', 'Bob', 'Alice', 'Bob', 'Charlie', 'Charlie'),
  Score = c(90, 85, 95, 88, 80, 78))
# Grouping by Name
result <- df |>
  group_by(Name) |>
  summarize(avg = mean(Score))

result
```

## Applying to Survey Data (ANES)

This block applies the same techniques to real survey data from the American National Election Studies (ANES). We filter to include only respondents with party affiliation less than 3 (typically Democrats and Republicans), group by their vote choice, and calculate the average feeling thermometer rating toward Biden for each voting group. The `na.rm = TRUE` argument removes missing values from the mean calculation.

```{r}

anes <-read_csv("anes.csv")

output <- anes |>
    filter(party < 3) |>
    group_by(vote) |>
    summarise(avg = mean(ft_biden, na.rm = T))

output
```

## 
