---
title: "Distributions"
author: "Will Horne"
format: revealjs
editor: visual
code-fold: True
---

## The Simplest Random Variable: Yes or No?

Many things in life have just two outcomes:

::: incremental
-   Did the candidate win? **Yes** or **No**
-   Did the patient recover? **Yes** or **No**
-   Did the bill pass? **Yes** or **No**
-   Did the coin land heads? **Yes** or **No**
:::

. . .

We call these **Bernoulli random variables**—named after mathematician Jakob Bernoulli.

## The Bernoulli Distribution

A **Bernoulli** random variable X takes only two values:

-   X = 1 (success) with probability **p**
-   X = 0 (failure) with probability **1 - p**

. . .

**Examples:**

| Situation       | X = 1 means...  | p might be... |
|-----------------|-----------------|---------------|
| Coin flip       | Heads           | 0.5           |
| Job application | Got the job     | 0.05          |
| Survey response | Supports policy | 0.62          |

. . .

**Notation:** We write X \~ Bern(p) to say "X follows a Bernoulli distribution with probability p"

## From One Trial to Many: The Binomial

A Bernoulli is one yes/no trial. But what if we repeat the trial multiple times?

. . .

**Example:** You send out 10 job applications. Each has a 20% chance of success (independent of the others). How many offers will you get?

. . .

This is a **Binomial** random variable:

-   **n** = number of trials (10 applications)
-   **p** = probability of success on each trial (0.20)
-   **X** = total number of successes (job offers)

**Notation:** X \~ Bin(n, p) or X \~ Bin(10, 0.2)

## What Can Happen? The Binomial Distribution

If you send 10 applications with 20% success rate each, you *could* get 0 offers, or 1, or 2... up to 10.

But these outcomes aren't equally likely!

. . .

```{r}
#| echo: false
#| fig-height: 4
library(ggplot2)
plot_data <- data.frame(
  x = 0:10,
  probability = dbinom(0:10, size = 10, prob = 0.2)
)
ggplot(plot_data, aes(x = factor(x), y = probability)) +
  geom_col(fill = "#522D80") +
  labs(x = "Number of Job Offers", y = "Probability",
       title = "10 Applications, 20% Success Rate Each") +
  theme_minimal(base_size = 14)
```

Most likely: 1-2 offers. Getting 0 is pretty likely too. Getting 7+ is very rare.

## The Binomial Formula (Don't Panic!)

The probability of getting **exactly k** successes out of **n** trials:

$$P(X = k) = \binom{n}{k} p^{k}(1-p)^{n-k}$$

. . .

**In plain English, this is:**

$$\text{(number of ways to get k successes)} \times \text{(prob of any single sequence with k successes)}$$

. . .

**Good news:** You don't need to calculate this by hand! R does it for you with `dbinom()`.

## Plotting a Binomial Distribution in R

```{r, echo = TRUE}
library(ggplot2)

# Set up the parameters
n <- 10        # number of trials
p <- 0.5       # probability of success

# Create a data frame with all possible outcomes (0 to n)
# and their probabilities using dbinom()
plot_data <- data.frame(
  x = 0:n,
  probability = dbinom(0:n, size = n, prob = p)
)

# Plot the distribution
ggplot(plot_data, aes(x = factor(x), y = probability)) +
  geom_col(fill = "steelblue") +
  labs(x = "Number of Successes",
       y = "Probability",
       title = "Binomial Distribution (n=10, p=0.5)") +
  theme_minimal()
```

## Working with Distributions in R

R uses a simple naming pattern for all distributions:

| Prefix | What it does | Example |
|-------------------|--------------------------------|---------------------|
| `d` | P(X = x) - probability of exactly x | `dbinom(3, size=10, prob=0.5)` |
| `p` | P(X ≤ x) - probability of x or less | `pbinom(3, size=10, prob=0.5)` |
| `r` | Generate random draws | `rbinom(100, size=10, prob=0.5)` |

## Quick Examples in R

```{r, echo = TRUE, eval = TRUE}
# What is the probability of getting exactly 3 heads in 10 coin flips?
dbinom(3, size = 10, prob = 0.5)

# What is the probability of getting 3 or fewer heads?
pbinom(3, size = 10, prob = 0.5)

# Generate 5 random binomial draws (n=10, p=0.5)
rbinom(5, size = 10, prob = 0.5)
```

## Why Does the Binomial Formula Work?

Let's think through it with a simple example: 5 coin flips, what's P(exactly 2 heads)?

. . .

**Step 1:** How many ways can we arrange 2 heads in 5 flips?

HHTTT, HTHTT, HTTHT, HTTTH, THHTT, THTHT, THTTH, TTHHT, TTHTH, TTTHH

That's $\binom{5}{2} = 10$ ways

. . .

**Step 2:** What's the probability of any ONE of these sequences?

Each flip is independent, so: $(0.5)^2 \times (0.5)^3 = 0.03125$

. . .

**Step 3:** Multiply!

$P(\text{exactly 2 heads}) = 10 \times 0.03125 = 0.3125$

. . .

**This is exactly what the formula does:** $\binom{n}{k} \times p^k \times (1-p)^{n-k}$

## What IS a Distribution, Exactly?

A distribution tells us **how probability is spread across possible values**.

. . .

For any random variable, we want to answer: "What's the probability of each possible outcome?"

. . .

Or alternatively, "what is the probability that the variables takes on a value bigger (or smaller) than some value?"

## Probability Mass Functions

For **discrete** random variables, the PMF tells us the probability of each specific value.

. . .

**Example:** Number of job offers from 10 applications (p = 0.2)

```{r}
#| echo: false
#| fig-height: 3
library(ggplot2)
plot_data <- data.frame(
  k = 0:10,
  prob = dbinom(0:10, size = 10, prob = 0.2)
)
ggplot(plot_data, aes(x = factor(k), y = prob)) +
  geom_col(fill = "#522D80") +
  geom_text(aes(label = round(prob, 3)), vjust = -0.5, size = 3) +
  labs(x = "k (number of offers)", y = "P(X = k)",
       title = "PMF of Binomial(10, 0.2)") +
  theme_minimal(base_size = 14)
```

Each bar shows P(X = k)—the probability of getting exactly k offers.

## Reading a PMF

From that PMF, we can answer questions like:

. . .

-   **P(X = 2)** = probability of exactly 2 offers ≈ 0.302
-   **P(X = 0)** = probability of no offers ≈ 0.107
-   **P(X ≥ 5)** = probability of 5 or more = sum of bars from 5 to 10

. . .

**Key property:** All the probabilities must add up to 1!

$$\sum_{k=0}^{10} P(X = k) = 1$$

(The probability of *something* happening is 100%)

## The PMF Formula for the Binomial

We've already seen this—the PMF is just the formula that gives us P(X = k):

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$$

. . .

**In R:** `dbinom(k, size = n, prob = p)` calculates this for you

```{r}
#| echo: true
# P(exactly 2 offers from 10 applications, 20% success rate)
dbinom(2, size = 10, prob = 0.2)
```

The `d` in `dbinom` stands for "density"—but for discrete distributions, it's really the PMF.

## Distributions Have "Stories"

Different distributions arise from different real-world scenarios:

| Distribution   | The Story                                         |
|----------------|---------------------------------------------------|
| Bernoulli      | One yes/no trial                                  |
| Binomial       | Count successes in n independent trials           |
| Hypergeometric | Count successes when sampling without replacement |
| Geometric      | How many failures before first success?           |

. . .

**Recognizing the story helps you choose the right distribution for your data!**

## When Sampling Without Replacement Matters

**The scenario:** A company has 500 men and 500 women (1000 total). Each year, they pick one employee for a prestigious management training program.

Over 10 years, only **1 woman** has been selected.

. . .

**The question:** Is this evidence of discrimination, or could it just be bad luck?

. . .

If selections were truly random, we'd expect about 5 women out of 10 picks. Getting only 1 seems low...but *how unlikely* is it? (Assume women have equal potential)

## Binomial vs. Hypergeometric: What's the Difference?

**Why can't we just use the Binomial here?**

. . .

The Binomial assumes each trial is **independent**—like flipping a coin.

But when you're selecting people from a fixed pool **without replacement**, the trials aren't independent!

-   If you pick a woman first, there are now 499 women and 500 men left

. . .

For large populations (like 1000 employees), this difference is tiny. But it exists!

## The Hypergeometric Distribution

The **Hypergeometric** distribution handles sampling without replacement:

-   You have a population with two types
-   You draw a sample without putting people back
-   You count how many of one type you got

. . .

**When to use it:**

-   Drawing cards from a deck
-   Selecting committee members from a group
-   Quality control (testing items from a batch)
-   Any time you're sampling from a finite population without replacement

. . .

**In R:** `dhyper()`, `phyper()`, `rhyper()`

## Back to the Discrimination Question

**Recall:** 500 women and 500 men. 10 selections over 10 years. Only 1 woman selected.

. . .

**Let's set up the problem:**

-   Population: 500 women (what we're counting) + 500 men = 1000 total
-   Sample size: 10 selections
-   Observed: only 1 woman

. . .

**The question:** If selections were truly random, how likely is it we'd see 1 or fewer women?

If this probability is low, that's evidence against fair selection.

## The Hypergeometric PMF

The probability of getting **exactly k** successes when drawing from a finite population:

$$P(X = k) = \frac{\binom{m}{k} \binom{N-m}{n-k}}{\binom{N}{n}}$$

. . .

**In plain English:**

$$\frac{\text{(ways to choose k women)} \times \text{(ways to choose remaining picks from men)}}{\text{(total ways to choose n people from everyone)}}$$

## Our Example

**For our example:** What's P(exactly 1 woman)?

$$P(X = 1) = \frac{\binom{500}{1} \binom{500}{9}}{\binom{1000}{10}}$$

This is tedious to calculate by hand—let's use R!

## Calculating in R

```{r}
#| echo: true

# Hypergeometric: P(X <= 1) where X = number of women selected
# phyper gives us P(X <= x)

phyper(1,           # x = 1 or fewer women
       m = 500,     # number of women in population
       n = 500,     # number of men in population
       k = 10)      # number of selections
```

. . .

**Only about 1% chance** of getting 1 or fewer women if selections were truly random!

This is strong evidence that something other than pure chance is happening.

## Visualizing the Distribution

What does the full distribution of "number of women selected" look like if selection is random?

## Plot of Hypergeometric

```{r, eval = F, echo = T}
# Load the ggplot2 package
library(ggplot2)

# Set parameters for the hypergeometric distribution
m <- 500  # Number of successes in the population
n <- 500  # Number of failures in the population
k <- 10   # Number of draws (sample size)

# Define the range of possible successes in the sample
x <- 0:k

# Calculate the probability mass function
probabilities <- dhyper(x, m, n, k)

# Create a data frame for plotting
df <- data.frame(Successes = x, Probability = probabilities)

# Plot using ggplot2
ggplot(df, aes(x = Successes, y = Probability)) +
  geom_col(fill = "skyblue", width = 0.7) +
  geom_point(color = "darkblue", size = 2) +
  labs(title = "Hypergeometric Distribution (m=500, n=500, k=10)",
       x = "Number of Successes",
       y = "Probability") +
  theme_minimal()


```

## Plot of Hypergeometric

```{r}
# Load the ggplot2 package
library(ggplot2)

# Set parameters for the hypergeometric distribution
m <- 500  # Number of successes in the population
n <- 500  # Number of failures in the population
k <- 10   # Number of draws (sample size)

# Define the range of possible successes in the sample
x <- 0:k

# Calculate the probability mass function
probabilities <- dhyper(x, m, n, k)

# Create a data frame for plotting
df <- data.frame(Successes = x, Probability = probabilities)

# Plot using ggplot2
ggplot(df, aes(x = Successes, y = Probability)) +
  geom_col(fill = "skyblue", width = 0.7) +
  geom_point(color = "darkblue", size = 2) +
  labs(title = "Hypergeometric Distribution (m=500, n=500, k=10)",
       x = "Number of Successes",
       y = "Probability") +
  theme_minimal()


```

## Hypergeometric vs Binomial

```{r, echo = T}

# Number of trials (number of employees selected)
k <- 10

# Total population size (total employees)
N <- 1000

# Number of successes in the population (number of women)
m <- 500

# Probability of success in the binomial distribution
p <- m / N  # Since the population is 50% women

# Hypergeometric Distribution
# Calculate P(X <= 1)
P_hyper <- phyper(1, m, N - m, k)
cat("Hypergeometric P(X <= 1):", P_hyper, "\n")

# Binomial Distribution
# Calculate P(X <= 1)
P_binom <- pbinom(1, size = k, prob = p)
cat("Binomial P(X <= 1):", P_binom, "\n")

```

HGeom and Binomial won't always be similar!

## Hypergeometric vs Binomial

**Small sample (k) relative to population size (n)**

```{r}
# Load necessary libraries
library(ggplot2)
library(reshape2)

# Parameters
N <- 100      # Total population size
m <- 20       # Number of successes in the population
k <- 5        # Sample size (small relative to N)
p <- m / N    # Probability of success

# Possible number of successes
x_vals <- 0:min(k, m)

# Hypergeometric PMF
hyper_pmf <- dhyper(x_vals, m, N - m, k)

# Binomial PMF
binom_pmf <- dbinom(x_vals, size = k, prob = p)

# Create a data frame for plotting
df <- data.frame(
  x = x_vals,
  Hypergeometric = hyper_pmf,
  Binomial = binom_pmf
)

# Melt the data frame for ggplot2
df_melt <- melt(df, id.vars = 'x', variable.name = 'Distribution', value.name = 'Probability')

# Generate the plot
ggplot(df_melt, aes(x = factor(x), y = Probability, fill = Distribution)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(
    title = paste('Figure 1: N =', N, ', m =', m, ', k =', k, ', p =', round(p, 2)),
    x = 'Number of Successes',
    y = 'Probability'
  ) +
  scale_fill_manual(values = c('blue', 'red')) +
  theme_minimal()

```

## Hypergeometric vs Binomial

**Larger sample (k=50, half of population)**

```{r}
# Load necessary libraries
library(ggplot2)
library(reshape2)

# Parameters
N <- 100
m <- 20
k <- 50      # Sample size is 50% of N
p <- m / N

# Possible number of successes
x_vals <- 0:min(k, m)

# Hypergeometric PMF
hyper_pmf <- dhyper(x_vals, m, N - m, k)

# Binomial PMF
binom_pmf <- dbinom(x_vals, size = k, prob = p)

# Create a data frame for plotting
df <- data.frame(
  x = x_vals,
  Hypergeometric = hyper_pmf,
  Binomial = binom_pmf
)

# Melt the data frame for ggplot2
df_melt <- melt(df, id.vars = 'x', variable.name = 'Distribution', value.name = 'Probability')

# Generate the plot
ggplot(df_melt, aes(x = factor(x), y = Probability, fill = Distribution)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(
    title = paste('Figure 3: N =', N, ', m =', m, ', k =', k, ', p =', round(p, 2)),
    x = 'Number of Successes',
    y = 'Probability'
  ) +
  scale_fill_manual(values = c('blue', 'red')) +
  theme_minimal()

```

## Hypergeometric vs Binomial

```{r}
# Load necessary libraries
library(ggplot2)
library(reshape2)

# Parameters
N <- 100       # Total population size
m <- 20        # Number of successes in the population
k <- 80        # Sample size is 80% of N
p <- m / N     # Probability of success

# Possible number of successes from 0 to 30
x_vals <- 0:30  # Extend x_vals to cover 0 to 30 successes

# Hypergeometric PMF
# For x > min(k, m), the hypergeometric PMF is zero
hyper_pmf <- dhyper(x_vals, m, N - m, k)

# Binomial PMF
binom_pmf <- dbinom(x_vals, size = k, prob = p)

# Create a data frame for plotting
df <- data.frame(
  x = x_vals,
  Hypergeometric = hyper_pmf,
  Binomial = binom_pmf
)

# Melt the data frame for ggplot2
df_melt <- melt(df, id.vars = 'x', variable.name = 'Distribution', value.name = 'Probability')

# Generate the plot with bar plot style
ggplot(df_melt, aes(x = factor(x), y = Probability, fill = Distribution)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(
    title = paste('Figure 4: N =', N, ', m =', m, ', k =', k, ', p =', round(p, 2)),
    x = 'Number of Successes',
    y = 'Probability'
  ) +
  scale_fill_manual(values = c('blue', 'red')) +
  theme_minimal()


```

## Discrete Uniform

Very simple story: Imagine N different items. Choose one item uniformly at random, what is the probability of choosing that item?

. . .

$$
P(X = x) = \frac{1}{N}
$$

Simple - but the intuition is important and sometimes a useful *baseline*

## Cumulative Distribution Functions

So far we've asked: "What's the probability of getting *exactly* 3 heads?" → **PMF**

. . .

But often we want to ask: "What's the probability of getting *3 or fewer* heads?" → **CDF**

. . .

The **Cumulative Distribution Function (CDF)** answers this by adding up the PMF:

$$F(x) = P(X \leq x) = P(X=0) + P(X=1) + P(X=2) + ... + P(X=x)$$

## CDF Example: Job Offers Again

What's the probability of getting **2 or fewer** offers from 10 applications?

```{r}
#| echo: true

# Add them up
dbinom(0, 10, 0.2) + dbinom(1, 10, 0.2) + dbinom(2, 10, 0.2)

# OR just use the CDF directly!
pbinom(2, 10, 0.2)  # P(X <= 2)
```

The CDF saves you from adding up all those PMF values manually!

## PMF vs CDF: The Key Difference

| Function | Question it answers | R function | Example              |
|----------|---------------------|------------|----------------------|
| **PMF**  | P(X = exactly k)    | `dbinom()` | P(exactly 2 offers)  |
| **CDF**  | P(X ≤ k)            | `pbinom()` | P(2 or fewer offers) |

. . .

**Useful trick:** P(X \> k) = 1 - P(X ≤ k) = 1 - CDF(k)

```{r}
#| echo: true
# P(more than 2 offers) = 1 - P(2 or fewer)
1 - pbinom(2, 10, 0.2)
```

## Plotting PMF vs CDF

```{r}
# Load necessary libraries
library(ggplot2)

# Parameters for Binomial Distribution
n <- 20
p <- 0.5
x <- 0:n

# Binomial distribution
binom_pmf <- dbinom(x, size = n, prob = p)
binom_cdf <- pbinom(x, size = n, prob = p)

# Create data frame for plotting
binom_data <- data.frame(x = x, pmf = binom_pmf, cdf = binom_cdf)

# Binomial Distribution: PMF and CDF on the same plot
ggplot() +
  geom_segment(aes(x = binom_data$x, xend = binom_data$x, y = 0, yend = binom_data$pmf),
               color = "blue", linewidth = 1.2) +
  geom_point(aes(x = binom_data$x, y = binom_data$pmf), color = "blue", size = 3) +
  labs(title = "Binomial Distribution (n=20, p=0.5)", x = "k", y = "P(X=k) or P(X<=k)") +
  geom_step(aes(x = binom_data$x, y = binom_data$cdf), color = "darkblue", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 15, y = 0.9, label = "CDF", color = "darkblue", size = 4) +
  annotate("text", x = 5, y = 0.2, label = "PMF", color = "blue", size = 4)

```

## Plotting PMF vs CDF

```{r}
# Load necessary libraries
library(ggplot2)

# Parameters for Uniform Distribution
n <- 20
x <- 0:n

# Uniform distribution
uniform_pmf <- rep(1 / (n + 1), n + 1)
uniform_cdf <- cumsum(uniform_pmf)

# Create data frame for plotting
uniform_data <- data.frame(x = x, pmf = uniform_pmf, cdf = uniform_cdf)

# Uniform Distribution: PMF and CDF on the same plot
ggplot() +
  geom_segment(aes(x = uniform_data$x, xend = uniform_data$x, y = 0, yend = uniform_data$pmf),
               color = "red", linewidth = 1.2) +
  geom_point(aes(x = uniform_data$x, y = uniform_data$pmf), color = "red", size = 3) +
  labs(title = "Uniform Distribution (n=20)", x = "k", y = "P(X=k) or P(X<=k)") +
  geom_step(aes(x = uniform_data$x, y = uniform_data$cdf), color = "darkred", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 15, y = 0.9, label = "CDF", color = "darkred", size = 4) +
  annotate("text", x = 5, y = 0.05, label = "PMF", color = "red", size = 4)

```

## What's Next?

We've covered:

-   Probability and conditioning
-   Random variables and several discrete distributions

. . .

**But distributions have a lot of information!** Sometimes we just want a quick summary.

**Next up:** Two key numbers that summarize any distribution:

1.  **Expectation** (the "center")
2.  **Variance** (the "spread")

## Expected Value: The Long-Run Average

**Imagine this game:** Roll a die. You win \$1 if it's a 1 or 2, and lose \$1 if it's 3, 4, 5, or 6.

. . .

If you played this game 1000 times, about how much would you win or lose overall?

. . .

-   Win \$1 about 333 times (1/3 of rolls) → +\$333
-   Lose \$1 about 667 times (2/3 of rolls) → -\$667
-   **Net: about -\$334**, or about **-\$0.33 per game**

. . .

This "long-run average" is the **expected value** of the game!

## Expected Value: The Formula

The expected value is just a **weighted average**—each outcome weighted by how likely it is:

$$E[X] = \sum_{\text{outcomes}} (\text{value}) \times (\text{probability})$$

. . .

**For our dice game:**

$$E[X] = (+1) \times \frac{2}{6} + (-1) \times \frac{4}{6} = \frac{2}{6} - \frac{4}{6} = -\frac{1}{3} \approx -0.33$$

. . .

**Key insight:** You won't win exactly -\$0.33 on any single game (impossible!), but it's what you'd average over many games.

## Expected Value: Policy Example

A job training program costs \$5,000 per participant. Research shows:

-   60% of participants get a job within 6 months (benefit: \$20,000 in wages)
-   40% don't find work in that time (benefit: \$0)

. . .

**What's the expected benefit per participant?**

$$E[\text{Benefit}] = (20000 \times 0.6) + (0 \times 0.4) = \$12,000$$

. . .

**Expected net gain:** \$12,000 - \$5,000 = **\$7,000 per participant**

This is how cost-benefit analysis works!

## Expectation is the center of gravity

```{r}
# Load necessary libraries
library(ggplot2)
library(gridExtra)

# Function to create binomial distribution plot
binom_plot <- function(size, prob) {
  x <- 0:size
  y <- dbinom(x, size, prob)
  mean_value <- size * prob
  data <- data.frame(x = x, y = y)

  ggplot(data, aes(x = x, y = y)) +
    geom_bar(stat = "identity", fill = "gray", color = "black", width = 0.7) +
    theme_minimal() +
    labs(x = "x", y = "f(x)", title = paste("Bin(", size, ",", prob, ")")) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title.x = element_text(size = 12),
      axis.title.y = element_text(size = 12)
    ) +
    geom_vline(xintercept = mean_value, linetype = "dashed", color = "black") +
    geom_point(aes(x = mean_value, y = 0), shape = 24, fill = "gray", size = 5)
}

# Function to create hypergeometric distribution plot
hgeom_plot <- function(m, n, k) {
  x <- 0:k
  y <- dhyper(x, m, n, k)
  mean_value <- k * m / (m + n)
  data <- data.frame(x = x, y = y)

  ggplot(data, aes(x = x, y = y)) +
    geom_bar(stat = "identity", fill = "gray", color = "black", width = 0.7) +
    theme_minimal() +
    labs(x = "x", y = "f(x)", title = "Hypergeom") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title.x = element_text(size = 12),
      axis.title.y = element_text(size = 12)
    ) +
    geom_vline(xintercept = mean_value, linetype = "dashed", color = "black") +
    geom_point(aes(x = mean_value, y = 0), shape = 24, fill = "gray", size = 5)
}

# Function to create uniform distribution plot
uniform_plot <- function(size) {
  x <- 0:size
  y <- dunif(x, min = 0, max = size)
  mean_value <- size / 2
  data <- data.frame(x = x, y = y)

  ggplot(data, aes(x = x, y = y)) +
    geom_bar(stat = "identity", fill = "gray", color = "black", width = 0.7) +
    theme_minimal() +
    labs(x = "x", y = "f(x)", title = "Uniform") +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title.x = element_text(size = 12),
      axis.title.y = element_text(size = 12)
    ) +
    geom_vline(xintercept = mean_value, linetype = "dashed", color = "black") +
    geom_point(aes(x = mean_value, y = 0), shape = 24, fill = "gray", size = 5)
}

# Generate the plots
plot1 <- binom_plot(10, 0.5)   # Binomial with p = 0.5
plot2 <- binom_plot(10, 0.25)  # Binomial with p = 0.25
plot3 <- hgeom_plot(m = 20, n = 15, k = 10)  # Hypergeometric distribution parameters
plot4 <- uniform_plot(10)  # Uniform distribution over {0, 1, ..., 10}

# Arrange plots in a 2x2 grid
grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)



```

## Binomial Expectation

**X** \~ (n,p). How can we find E\[X\]? Plug in the PMF to the expectation formula:

. . .

$$E[X] = \sum_{k=0}^{n} k \binom{n}{k} p^{k}q^{n-k}$$

. . .

We can simplify this - covered in the book but requires theorems and combinatorics beyond the scope of the class. But, can we use a story of what the binomial is to reason out what E\[X\] must be?

. . .

If X \~ Bin(n,p), E\[X\] = np.

## Geometric Distribution

Geom(p): Independent Bernoulli Trials. How many failures before the first success?

Let **X** \~ Geom(p), q = 1-p. Then the **PMF** must be.$$ P(X = k) = pq^{k}$$

Think through why that must be the PMF

## Geometric Expectation

From the definition of expectation, where we plug the PMF in for P(X=x):

$$
E[X] = \sum_{k = 0}^{\infty} kpq^{k}
$$

We *could* use calculus to work out how this simplifies to $\frac{q}{p}$ (see Example 4.3.6 in B&H), but the key takeaway is:

$$E[X] = \frac{q}{p} = \frac{1-p}{p}$$

## Trials Until First Success?

**What about total trials until success?** That's failures + 1 (the success):

$$E[\text{Total Trials}] = \frac{q}{p} + 1 = \frac{q + p}{p} = \frac{1}{p}$$

. . .

So if p = 0.2, we expect 4 failures and 5 total trials on average.

## Plotting Geometric Distribution

```{r, echo = TRUE}
library(ggplot2)

# Parameters
p <- 0.3  # probability of success on each trial

# Create data: possible number of failures (0, 1, 2, ..., 15)
# and their probabilities using dgeom()
plot_data <- data.frame(
  failures = 0:15,
  probability = dgeom(0:15, prob = p)
)

# Plot the distribution
ggplot(plot_data, aes(x = factor(failures), y = probability)) +
  geom_col(fill = "steelblue") +
  labs(x = "Number of Failures Before First Success",
       y = "Probability",
       title = "Geometric Distribution (p = 0.3)") +
  theme_minimal()
```

## Realistic Geometric RV Example

In labor market studies, the geometric distribution can model the **number of applications an individual submits until they receive a job offer**.

During a recession, there are many more applicants than there are jobs. Imagine that an application has a probability of success .015. How many applications would we expect an applicant to send before getting a job. If an applicant can apply to 7 jobs a week, how long will it take them to find work?

. . .

We want total applications until the first success, so we use $E[\text{Total Trials}] = \frac{1}{p} = \frac{1}{0.015} = 66.67$, or about 67 applications. At 7 applications a week, an average job search would take about 10 weeks.

## Other Discrete Distributions

There are other discrete distributions you might encounter:

-   **Negative Binomial**: How many failures before n successes
-   **Poisson**: Models count data (e.g., number of events per time period)

. . .

**Key Point**: The type of data generating process tells us which distribution (and statistical model) to use.

## The Expectation Isn't the Whole Story

**Two archers both hit the bullseye "on average"...**

```{r}
#| echo: false
#| fig-height: 4
library(ggplot2)
library(patchwork)

set.seed(42)
# Consistent archer
archer1 <- data.frame(x = rnorm(50, 0, 0.5), y = rnorm(50, 0, 0.5))
# Inconsistent archer
archer2 <- data.frame(x = rnorm(50, 0, 2), y = rnorm(50, 0, 2))

p1 <- ggplot(archer1, aes(x, y)) +
  geom_point(color = "#522D80", size = 2, alpha = 0.7) +
  geom_point(aes(x = 0, y = 0), color = "#F66733", size = 4, shape = 3, stroke = 2) +
  xlim(-6, 6) + ylim(-6, 6) +
  labs(title = "Archer A: Consistent", subtitle = "Low variance") +
  theme_minimal() +
  theme(axis.title = element_blank(), axis.text = element_blank())

p2 <- ggplot(archer2, aes(x, y)) +
  geom_point(color = "#522D80", size = 2, alpha = 0.7) +
  geom_point(aes(x = 0, y = 0), color = "#F66733", size = 4, shape = 3, stroke = 2) +
  xlim(-6, 6) + ylim(-6, 6) +
  labs(title = "Archer B: All Over the Place", subtitle = "High variance") +
  theme_minimal() +
  theme(axis.title = element_blank(), axis.text = element_blank())

p1 + p2
```

Both have the same **expected value** (bullseye). But they're very different archers!

## Why Spread Matters

::: incremental
-   **Archer A** is consistent: you know roughly what to expect

-   **Archer B** is unpredictable: could hit bullseye or miss badly

-   In policy terms: Would you rather have a program that helps everyone a little, or one that dramatically helps some and does nothing for others?

    -   Not sure that there is actually an obvious answer here!

-   **Variance** tells us how spread out the outcomes are around the average
:::

## Same Expectation, Different Spread

```{r}
#| echo: false
#| fig-height: 4.5
library(ggplot2)

set.seed(123)
dist_1 <- rnorm(1000, mean = 0, sd = 1)
dist_2 <- rnorm(1000, mean = 0, sd = 2)

df <- data.frame(
  value = c(dist_1, dist_2),
  group = rep(c("Low Variance (SD=1)", "High Variance (SD=2)"), each = 1000)
)

ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 50) +
  scale_fill_manual(values = c("#522D80", "#F66733")) +
  labs(title = "Same Mean, Different Spread",
       x = "Value", y = "Count", fill = "") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

Both distributions are centered at 0, but the orange one is much more spread out.

## Variance: Measuring Spread

**Variance** measures how far values typically fall from the mean.

. . .

**Intuition:**

1.  Take each value, measure how far it is from the mean
2.  Square those distances (so negatives don't cancel positives)
3.  Average them (weighted by probability)

. . .

$$Var[X] = E[(X - E[X])^{2}]$$

"The expected squared distance from the mean"

## Standard Deviation

$$
Var[X] = E[(X -E[X])^{2}]
$$

Note that the variance is the squared distance from the mean. If we want to know, on average, how far a realization of X will be from the $E[X]$, we can calculate the standard deviation

$$
SD(X) = \sqrt{Var[X]}
$$

In practice, we usually work with variance rather than standard deviation, but standard deviation is more immediately interpretable

## Example: Program Outcomes {.smaller}

A workforce development program tracks participant outcomes on a 1–6 satisfaction scale. Based on past cohorts, the distribution looks like:

| Rating (x) | P(X = x) | Meaning           |
|------------|----------|-------------------|
| 1          | 0.1      | Very dissatisfied |
| 2          | 0.15     | Dissatisfied      |
| 3          | 0.2      | Neutral           |
| 4          | 0.25     | Satisfied         |
| 5          | 0.2      | Very satisfied    |
| 6          | 0.1      | Exceptional       |

Calculate the Expectation, Variance, and Standard Deviation. What does the variance tell you about how *consistently* participants experience the program?

## Another Variance Example

Imagine we want to know both the average (expectation) and spread (variance/standard deviation) of incomes in a community. We randomly select 10 households who have the following incomes (in thousands of US Dollars). Let **X** be the income of each respondent. We have:

$$
X = [45,50,52,47,60,55,120,28,430,73]
$$

How can we calculate the expectation, variance and standard deviation?

## Properties of Variance

::: incremental
-   Var\[X + c\] = Var\[X\] for any constant

-   If c is a constant, $Var[cX] = c^{2}Var[X]$

-   If and only if X and Y are **independent,** Var\[X + Y\] = Var\[X\] + Var\[Y\]

-   If **X** is not a constant, Var\[X\] \> 0.
:::

## Binomial Variance

For a Bernoulli random variable: $$Var[X] = p(1-p)$$

. . .

Since a Binomial is the sum of n independent Bernoulli trials:

$$Var[X] = np(1-p)$$

## Taking Stock

::: incremental
-   So far: Probability Theory and Discrete Random Variables

    -   You should feel comfortable about distributions of data that take on discrete values

    -   You should have a good idea of what the PMF and CDF of discrete variables mean

-   Now: Same idea for variables that can take on any value

    -   Many variables we care about as social scientists are (approximately) continuous

        -   Income, Time, Tax Rates, Vote Shares

        -   Sample means of variables are also continuous
:::